---
title: "Bayesian Additive Regression Trees for Supervised Learning"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(rprojroot::find_root(rprojroot::has_file(".here")), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

This vignette demonstrates how to sample variants of the BART model (@chipman2010bart), using the `bart()` function in `stochtree`. The original BART model is 

$$
\begin{aligned}
y_i \mid X_i = x_i &\sim \mathcal{N}(f(x_i), \sigma^2)\\
\sigma^2 &\sim \text{IG}\left(\frac{\nu}{2}, \frac{\nu \lambda}{2}\right)
\end{aligned}
$$

where 

$$
f(X) = \sum_{s=1}^m g_s(X)
$$

and each $g_s$ refers to a decision tree function which partitions $X$ into $k_s$ mutually exclusive regions ($\mathcal{A}_s = \mathcal{A}_{s,1} \cup \dots \cup \mathcal{A}_{s,k_s}$) and assigns a scalar parameter $\mu_{s,j}$ to each region $\mathcal{A}_{s,j}$

$$
g_s(x) = \sum_{j = 1}^{k_s} \mu_{s,j} \mathbb{I}\left(x \in \mathcal{A}_{s,j}\right).
$$

The partitions $\mathcal{A}_s$ are defined by a series of logical split rules $X_i \leq c$ where $i$ is a variable index and $c$ is a numeric cutpoint and these partitions are guided by a uniform prior on variables and cutpoints. The prior on partitions is further specified by a probability of splitting a node

$$
P(\text{split node } \eta) = \alpha (1 + \text{depth}_{\eta})^{-\beta}
$$

The prior for each leaf node parameter is 

$$
\mu_{s,j} \sim \mathcal{N}\left(0, \sigma^2_{\mu}\right)
$$

Together, we refer to this conditional mean model as 

$$
f(X) \sim \text{BART}(\alpha, \beta, m)
$$

This is the "core" of stochtree's supervised learning interface, though we support many expanded models including

* linear leaf regression (i.e. each leaf node evaluates a linear regression on basis $W$ rather than return a constant),
* additive random effects,
* forest-based heteroskedasticity,
* binary / ordinal outcome modeling using the probit and complementary log-log (cloglog) links,

and we offer the ability to sample any of the above models using the MCMC or the Grow-From-Root sampler (@he2023stochastic).

# Setup

To begin, we load the `stochtree` and other necessary packages.

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
```

## Python

```{python}
import matplotlib.pyplot as plt
import numpy as np
import pandas as pd
from sklearn.model_selection import train_test_split
from stochtree import BARTModel
```

::::

We set a seed for reproducibility

::::{.panel-tabset group="language"}

## R

```{r}
random_seed <- 1234
set.seed(random_seed)
```

## Python

```{python}
random_seed = 1234
rng = np.random.default_rng(random_seed)
```

::::

# Demo 1: Step Function

## Data Generation

We generate data from a simple step function

::::{.panel-tabset group="language"}

## R

```{r}
# Generate the data
n <- 500
p_x <- 10
snr <- 3
X <- matrix(runif(n * p_x), ncol = p_x)
f_XW <- (((0 <= X[, 1]) & (0.25 > X[, 1])) *
  (-7.5) +
  ((0.25 <= X[, 1]) & (0.5 > X[, 1])) * (-2.5) +
  ((0.5 <= X[, 1]) & (0.75 > X[, 1])) * (2.5) +
  ((0.75 <= X[, 1]) & (1 > X[, 1])) * (7.5))
noise_sd <- sd(f_XW) / snr
y <- f_XW + rnorm(n, 0, 1) * noise_sd
```

## Python

```{python}
# Generate the data
n = 500
p_x = 10
snr = 3
X = rng.uniform(0, 1, (n, p_x))
f_XW = (
    ((X[:, 0] >= 0.0) & (X[:, 0] < 0.25)) * (-7.5)
    + ((X[:, 0] >= 0.25) & (X[:, 0] < 0.5)) * (-2.5)
    + ((X[:, 0] >= 0.5) & (X[:, 0] < 0.75)) * (2.5)
    + ((X[:, 0] >= 0.75) & (X[:, 0] < 1.0)) * (7.5)
)
noise_sd = np.std(f_XW) / snr
y = f_XW + rng.normal(0, noise_sd, n)
```

::::

Split the data into train and test sets

::::{.panel-tabset group="language"}

## R

```{r}
test_set_pct <- 0.2
n_test <- round(test_set_pct * n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds, ])
X_train <- as.data.frame(X[train_inds, ])
y_test <- y[test_inds]
y_train <- y[train_inds]
```

## Python

```{python}
test_set_pct = 0.2
X_train, X_test, y_train, y_test = train_test_split(
    X, y, test_size=test_set_pct, random_state=random_seed
)
```

::::

## Sampling and Analysis

We sample from a BART model of $y \mid X$ with 10 grow-from-root GFR samples (@he2023stochastic) followed by 100 MCMC samples (this is the default in `stochtree`), run for 4 chains initialized by different GFR iterations. 

We also specify $m = 100$ trees and we let both $\sigma^2$ and $\sigma^2_{\mu}$ be updated by Gibbs samplers.

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
general_params <- list(
  sample_sigma2_global = T,
  num_threads = 1,
  num_chains = 4,
  random_seed = random_seed
)
mean_forest_params <- list(sample_sigma2_leaf = T, num_trees = 100)
bart_model <- stochtree::bart(
  X_train = X_train,
  y_train = y_train,
  X_test = X_test,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = general_params,
  mean_forest_params = mean_forest_params
)
```

## Python

```{python}
num_gfr = 10
num_burnin = 0
num_mcmc = 100
general_params = {
    "sample_sigma2_global": True,
    "num_threads": 1,
    "num_chains": 4,
    "random_seed": random_seed,
}
mean_forest_params = {"sample_sigma2_leaf": True, "num_trees": 100}
bart_model = BARTModel()
bart_model.sample(
    X_train=X_train,
    y_train=y_train,
    X_test=X_test,
    num_gfr=num_gfr,
    num_burnin=num_burnin,
    num_mcmc=num_mcmc,
    general_params=general_params,
    mean_forest_params=mean_forest_params,
)
```

::::

Plot the mean outcome predictions versus the true outcomes

::::{.panel-tabset group="language"}

## R

```{r}
y_hat_test <- predict(
  bart_model,
  X = X_test, 
  terms = "y_hat", 
  type = "mean"
)
plot(
  y_hat_test,
  y_test,
  xlab = "predicted",
  ylab = "actual",
  main = "Outcome"
)
abline(0, 1, col = "red", lty = 3, lwd = 3)
```

## Python

```{python}
y_hat_test = bart_model.predict(X=X_test, terms="y_hat", type="mean")
lo, hi = min(y_hat_test.min(), y_test.min()), max(y_hat_test.max(), y_test.max())
plt.scatter(y_hat_test, y_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted")
plt.ylabel("Actual")
plt.title("Outcome")
plt.show()
```

::::

Plot the $\sigma^2$ traceplot

::::{.panel-tabset group="language"}

## R

```{r}
sigma_observed <- var(y - f_XW)
sigma2_global_samples <- extractParameter(bart_model, "sigma2_global")
plot_bounds <- c(
  min(c(sigma2_global_samples, sigma_observed)),
  max(c(sigma2_global_samples, sigma_observed))
)
plot(
  sigma2_global_samples,
  ylim = plot_bounds,
  ylab = "sigma^2",
  xlab = "Sample",
  main = "Global variance parameter"
)
abline(h = sigma_observed, lty = 3, lwd = 3, col = "blue")
```

## Python

```{python}
sigma_observed = np.var(y - f_XW)
global_var_samples = bart_model.extract_parameter("sigma2_global")
plt.plot(global_var_samples)
plt.axhline(sigma_observed, color="blue", linestyle="dashed", linewidth=2)
plt.xlabel("Sample")
plt.ylabel(r"$\sigma^2$")
plt.title("Global variance parameter")
plt.show()
```

::::

# References
