---
title: "BCF: Bayesian Causal Forests"
---

Bayesian Causal Forests (BCF) for estimating heterogeneous treatment effects,
with separate tree ensembles for the prognostic function and treatment effect.

::: {.callout-note}
This vignette is under construction. Content will be ported from:

- **Python**: `stochtree_repo/demo/notebooks/causal_inference.ipynb`
- **R**: pkgdown article `CausalInference`
:::

## Background

BCF [@hahn2020bayesian] models the conditional average treatment effect (CATE)
by fitting two separate tree ensembles:

$$
Y_i = \mu(X_i) + \tau(X_i) Z_i + \epsilon_i, \quad \epsilon_i \sim \mathcal{N}(0, \sigma^2)
$$

where $\mu(\cdot)$ is a prognostic forest and $\tau(\cdot)$ is a treatment effect forest.
The estimated propensity score $\hat{\pi}(X_i)$ is included as a covariate in $\mu(\cdot)$
to reduce confounding bias.

## Setup

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
library(stochtree)
```

## Python

```{python}
#| eval: false
import stochtree
import numpy as np
```

::::

## Data Simulation

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Simulate observational data with heterogeneous treatment effects
```

## Python

```{python}
#| eval: false
# Simulate observational data with heterogeneous treatment effects
```

::::

## Propensity Score Estimation

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Estimate propensity scores
```

## Python

```{python}
#| eval: false
# Estimate propensity scores
```

::::

## Model Fitting

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Fit BCF model
```

## Python

```{python}
#| eval: false
# Fit BCF model
```

::::

## Posterior Summaries

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Summarize CATE posterior
```

## Python

```{python}
#| eval: false
# Summarize CATE posterior
```

::::

## References
