---
title: "Causal Inference with BCF"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(here::here(), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

This vignette demonstrates how to use the `bcf()` function for causal inference
(@hahn2020bayesian). BCF models the conditional average treatment effect (CATE)
by fitting two separate tree ensembles

$$
Y_i = \mu(X_i) + \tau(X_i) Z_i + \epsilon_i, \quad \epsilon_i \sim \mathcal{N}(0, \sigma^2)
$$

where $\mu(\cdot)$ is a prognostic forest and $\tau(\cdot)$ is a treatment effect
forest. The estimated propensity score $\hat{\pi}(X_i)$ is included as a covariate
in $\mu(\cdot)$ to reduce confounding bias.

# Setup

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
```

## Python

```{python}
import numpy as np
import pandas as pd
import matplotlib.pyplot as plt
from scipy.stats import norm
from stochtree import BCFModel
```

::::

We also define several simple functions that configure the data generating processes
used in this vignette.

::::{.panel-tabset group="language"}

## R

```{r}
g <- function(x) {
  ifelse(x[, 5] == 1, 2, ifelse(x[, 5] == 2, -1, -4))
}
mu1 <- function(x) {
  1 + g(x) + x[, 1] * x[, 3]
}
mu2 <- function(x) {
  1 + g(x) + 6 * abs(x[, 3] - 1)
}
tau1 <- function(x) {
  rep(3, nrow(x))
}
tau2 <- function(x) {
  1 + 2 * x[, 2] * x[, 4]
}
```

## Python

```{python}
# Note: Python uses 0-based column indexing (R's X[,1] = Python's X[:,0])
def g(x):
    return np.where(x[:, 4] == 1, 2, np.where(x[:, 4] == 2, -1, -4))

def mu1(x):
    return 1 + g(x) + x[:, 0] * x[:, 2]

def mu2(x):
    return 1 + g(x) + 6 * np.abs(x[:, 2] - 1)

def tau1(x):
    return np.full(x.shape[0], 3.0)

def tau2(x):
    return 1 + 2 * x[:, 1] * x[:, 3]

rng = np.random.default_rng(101)
```

::::

# Binary Treatment

## Demo 1: Nonlinear Outcome Model, Heterogeneous Treatment Effect

We consider the following data generating process from @hahn2020bayesian:

\begin{equation*}
\begin{aligned}
y &= \mu(X) + \tau(X) Z + \epsilon\\
\epsilon &\sim N\left(0,\sigma^2\right)\\
\mu(X) &= 1 + g(X) + 6 \lvert X_3 - 1 \rvert\\
\tau(X) &= 1 + 2 X_2 X_4\\
g(X) &= \mathbb{I}(X_5=1) \times 2 - \mathbb{I}(X_5=2) \times 1 - \mathbb{I}(X_5=3) \times 4\\
s_{\mu} &= \sqrt{\mathbb{V}(\mu(X))}\\
\pi(X) &= 0.8 \phi\left(\frac{3\mu(X)}{s_{\mu}}\right) - \frac{X_1}{2} + \frac{2U+1}{20}\\
X_1,X_2,X_3 &\sim N\left(0,1\right)\\
X_4 &\sim \text{Bernoulli}(1/2)\\
X_5 &\sim \text{Categorical}(1/3,1/3,1/3)\\
U &\sim \text{Uniform}\left(0,1\right)\\
Z &\sim \text{Bernoulli}\left(\pi(X)\right)
\end{aligned}
\end{equation*}

### Simulation

We draw from the DGP defined above.

::::{.panel-tabset group="language"}

## R

```{r}
n <- 500
snr <- 3
x1 <- rnorm(n)
x2 <- rnorm(n)
x3 <- rnorm(n)
x4 <- as.numeric(rbinom(n, 1, 0.5))
x5 <- as.numeric(sample(1:3, n, replace = TRUE))
X <- cbind(x1, x2, x3, x4, x5)
p <- ncol(X)
mu_x <- mu1(X)
tau_x <- tau2(X)
pi_x <- 0.8 * pnorm((3 * mu_x / sd(mu_x)) - 0.5 * X[, 1]) + 0.05 + runif(n) / 10
Z <- rbinom(n, 1, pi_x)
E_XZ <- mu_x + Z * tau_x
y <- E_XZ + rnorm(n, 0, 1) * (sd(E_XZ) / snr)
X <- as.data.frame(X)
X$x4 <- factor(X$x4, ordered = TRUE)
X$x5 <- factor(X$x5, ordered = TRUE)

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct * n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- X[test_inds, ]
X_train <- X[train_inds, ]
pi_test <- pi_x[test_inds]
pi_train <- pi_x[train_inds]
Z_test <- Z[test_inds]
Z_train <- Z[train_inds]
y_test <- y[test_inds]
y_train <- y[train_inds]
mu_test <- mu_x[test_inds]
mu_train <- mu_x[train_inds]
tau_test <- tau_x[test_inds]
tau_train <- tau_x[train_inds]
```

## Python

```{python}
n = 500
snr = 3
x1 = rng.normal(size=n)
x2 = rng.normal(size=n)
x3 = rng.normal(size=n)
x4 = rng.binomial(1, 0.5, n).astype(float)
x5 = rng.choice([1, 2, 3], size=n).astype(float)
X = np.column_stack([x1, x2, x3, x4, x5])
mu_x = mu1(X)
tau_x = tau2(X)
pi_x = (0.8 * norm.cdf((3 * mu_x / np.std(mu_x)) - 0.5 * X[:, 0])
        + 0.05 + rng.uniform(size=n) / 10)
Z = rng.binomial(1, pi_x, n).astype(float)
E_XZ = mu_x + Z * tau_x
y = E_XZ + rng.normal(size=n) * (np.std(E_XZ) / snr)

# Convert to DataFrame with ordered categoricals (matching R's factor(..., ordered=TRUE))
X_df = pd.DataFrame({"x1": x1, "x2": x2, "x3": x3, "x4": x4, "x5": x5})
X_df["x4"] = pd.Categorical(X_df["x4"].astype(int), categories=[0, 1], ordered=True)
X_df["x5"] = pd.Categorical(X_df["x5"].astype(int), categories=[1, 2, 3], ordered=True)

# Split data into test and train sets
test_set_pct = 0.2
n_test = round(test_set_pct * n)
n_train = n - n_test
test_inds = rng.choice(n, n_test, replace=False)
train_inds = np.setdiff1d(np.arange(n), test_inds)
X_test  = X_df.iloc[test_inds]
X_train = X_df.iloc[train_inds]
pi_test, pi_train   = pi_x[test_inds], pi_x[train_inds]
Z_test,  Z_train    = Z[test_inds],    Z[train_inds]
y_test,  y_train    = y[test_inds],    y[train_inds]
mu_test, mu_train   = mu_x[test_inds], mu_x[train_inds]
tau_test, tau_train = tau_x[test_inds], tau_x[train_inds]
```

::::

### Sampling and Analysis

#### Warmstart

We first simulate from an ensemble model of $y \mid X$ using "warm-start"
initialization samples (@krantsevich2023stochastic). This is the default in
`stochtree`.

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
general_params <- list(keep_every = 5)
prognostic_forest_params <- list(sample_sigma2_leaf = F)
treatment_effect_forest_params <- list(sample_sigma2_leaf = F)
bcf_model_warmstart <- bcf(
  X_train = X_train,
  Z_train = Z_train,
  y_train = y_train,
  propensity_train = pi_train,
  X_test = X_test,
  Z_test = Z_test,
  propensity_test = pi_test,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = general_params,
  prognostic_forest_params = prognostic_forest_params,
  treatment_effect_forest_params = treatment_effect_forest_params
)
```

## Python

```{python}
bcf_model_warmstart = BCFModel()
bcf_model_warmstart.sample(
    X_train=X_train, Z_train=Z_train, y_train=y_train, propensity_train=pi_train,
    X_test=X_test,   Z_test=Z_test,   propensity_test=pi_test,
    num_gfr=10, num_burnin=0, num_mcmc=100,
    general_params={"num_threads": 1, "keep_every": 5},
    prognostic_forest_params={"sample_sigma2_leaf": False},
    treatment_effect_forest_params={"sample_sigma2_leaf": False},
)
```

::::

Inspect the samples initialized with an XBART warm-start.

::::{.panel-tabset group="language"}

## R

```{r}
plot(
  rowMeans(bcf_model_warmstart$mu_hat_test),
  mu_test,
  xlab = "predicted",
  ylab = "actual",
  main = "Prognostic function"
)
abline(0, 1, col = "red", lty = 3, lwd = 3)
plot(
  rowMeans(bcf_model_warmstart$tau_hat_test),
  tau_test,
  xlab = "predicted",
  ylab = "actual",
  main = "Treatment effect"
)
abline(0, 1, col = "red", lty = 3, lwd = 3)
sigma_observed <- var(y - E_XZ)
plot_bounds <- c(
  min(c(bcf_model_warmstart$sigma2_global_samples, sigma_observed)),
  max(c(bcf_model_warmstart$sigma2_global_samples, sigma_observed))
)
plot(
  bcf_model_warmstart$sigma2_global_samples,
  ylim = plot_bounds,
  ylab = "sigma^2",
  xlab = "Sample",
  main = "Global variance parameter"
)
abline(h = sigma_observed, lty = 3, lwd = 3, col = "blue")
```

## Python

```{python}
sigma_observed = np.var(y - E_XZ)

mu_pred = bcf_model_warmstart.mu_hat_test.mean(axis=1)
lo, hi = min(mu_pred.min(), mu_test.min()), max(mu_pred.max(), mu_test.max())
plt.scatter(mu_pred, mu_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Prognostic function")
plt.show()

tau_pred = bcf_model_warmstart.tau_hat_test.mean(axis=1)
lo, hi = min(tau_pred.min(), tau_test.min()), max(tau_pred.max(), tau_test.max())
plt.scatter(tau_pred, tau_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Treatment effect")
plt.show()

plt.plot(bcf_model_warmstart.global_var_samples)
plt.axhline(sigma_observed, color="blue", linestyle="dashed", linewidth=2)
plt.xlabel("Sample"); plt.ylabel(r"$\sigma^2$"); plt.title("Global variance parameter")
plt.show()
```

::::

Examine test set interval coverage.

::::{.panel-tabset group="language"}

## R

```{r}
test_lb <- apply(bcf_model_warmstart$tau_hat_test, 1, quantile, 0.025)
test_ub <- apply(bcf_model_warmstart$tau_hat_test, 1, quantile, 0.975)
cover <- ((test_lb <= tau_x[test_inds]) &
  (test_ub >= tau_x[test_inds]))
mean(cover)
```

## Python

```{python}
test_lb = np.quantile(bcf_model_warmstart.tau_hat_test, 0.025, axis=1)
test_ub = np.quantile(bcf_model_warmstart.tau_hat_test, 0.975, axis=1)
cover = (test_lb <= tau_test) & (test_ub >= tau_test)
print(f"Coverage: {cover.mean():.3f}")
```

::::

#### BART MCMC without Warmstart

Next, we simulate from this ensemble model without any warm-start initialization.

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 0
num_burnin <- 2000
num_mcmc <- 100
general_params <- list(keep_every = 5)
prognostic_forest_params <- list(sample_sigma2_leaf = F)
treatment_effect_forest_params <- list(sample_sigma2_leaf = F)
bcf_model_root <- bcf(
  X_train = X_train,
  Z_train = Z_train,
  y_train = y_train,
  propensity_train = pi_train,
  X_test = X_test,
  Z_test = Z_test,
  propensity_test = pi_test,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = general_params,
  prognostic_forest_params = prognostic_forest_params,
  treatment_effect_forest_params = treatment_effect_forest_params
)
```

## Python

```{python}
bcf_model_root = BCFModel()
bcf_model_root.sample(
    X_train=X_train, Z_train=Z_train, y_train=y_train, propensity_train=pi_train,
    X_test=X_test,   Z_test=Z_test,   propensity_test=pi_test,
    num_gfr=0, num_burnin=2000, num_mcmc=100,
    general_params={"num_threads": 1, "keep_every": 5},
    prognostic_forest_params={"sample_sigma2_leaf": False},
    treatment_effect_forest_params={"sample_sigma2_leaf": False},
)
```

::::

Inspect the samples after burnin.

::::{.panel-tabset group="language"}

## R

```{r}
plot(
  rowMeans(bcf_model_root$mu_hat_test),
  mu_test,
  xlab = "predicted",
  ylab = "actual",
  main = "Prognostic function"
)
abline(0, 1, col = "red", lty = 3, lwd = 3)
plot(
  rowMeans(bcf_model_root$tau_hat_test),
  tau_test,
  xlab = "predicted",
  ylab = "actual",
  main = "Treatment effect"
)
abline(0, 1, col = "red", lty = 3, lwd = 3)
sigma_observed <- var(y - E_XZ)
plot_bounds <- c(
  min(c(bcf_model_root$sigma2_global_samples, sigma_observed)),
  max(c(bcf_model_root$sigma2_global_samples, sigma_observed))
)
plot(
  bcf_model_root$sigma2_global_samples,
  ylim = plot_bounds,
  ylab = "sigma^2",
  xlab = "Sample",
  main = "Global variance parameter"
)
abline(h = sigma_observed, lty = 3, lwd = 3, col = "blue")
```

## Python

```{python}
mu_pred = bcf_model_root.mu_hat_test.mean(axis=1)
lo, hi = min(mu_pred.min(), mu_test.min()), max(mu_pred.max(), mu_test.max())
plt.scatter(mu_pred, mu_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Prognostic function")
plt.show()

tau_pred = bcf_model_root.tau_hat_test.mean(axis=1)
lo, hi = min(tau_pred.min(), tau_test.min()), max(tau_pred.max(), tau_test.max())
plt.scatter(tau_pred, tau_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Treatment effect")
plt.show()

plt.plot(bcf_model_root.global_var_samples)
plt.axhline(sigma_observed, color="blue", linestyle="dashed", linewidth=2)
plt.xlabel("Sample"); plt.ylabel(r"$\sigma^2$"); plt.title("Global variance parameter")
plt.show()
```

::::

Examine test set interval coverage.

::::{.panel-tabset group="language"}

## R

```{r}
test_lb <- apply(bcf_model_root$tau_hat_test, 1, quantile, 0.025)
test_ub <- apply(bcf_model_root$tau_hat_test, 1, quantile, 0.975)
cover <- ((test_lb <= tau_x[test_inds]) &
  (test_ub >= tau_x[test_inds]))
mean(cover)
```

## Python

```{python}
test_lb = np.quantile(bcf_model_root.tau_hat_test, 0.025, axis=1)
test_ub = np.quantile(bcf_model_root.tau_hat_test, 0.975, axis=1)
cover = (test_lb <= tau_test) & (test_ub >= tau_test)
print(f"Coverage: {cover.mean():.3f}")
```

::::

## Demo 2: Linear Outcome Model, Heterogeneous Treatment Effect

We consider the following data generating process from @hahn2020bayesian:

\begin{equation*}
\begin{aligned}
y &= \mu(X) + \tau(X) Z + \epsilon\\
\epsilon &\sim N\left(0,\sigma^2\right)\\
\mu(X) &= 1 + g(X) + 6 X_1 X_3\\
\tau(X) &= 1 + 2 X_2 X_4\\
g(X) &= \mathbb{I}(X_5=1) \times 2 - \mathbb{I}(X_5=2) \times 1 - \mathbb{I}(X_5=3) \times 4\\
s_{\mu} &= \sqrt{\mathbb{V}(\mu(X))}\\
\pi(X) &= 0.8 \phi\left(\frac{3\mu(X)}{s_{\mu}}\right) - \frac{X_1}{2} + \frac{2U+1}{20}\\
X_1,X_2,X_3 &\sim N\left(0,1\right)\\
X_4 &\sim \text{Bernoulli}(1/2)\\
X_5 &\sim \text{Categorical}(1/3,1/3,1/3)\\
U &\sim \text{Uniform}\left(0,1\right)\\
Z &\sim \text{Bernoulli}\left(\pi(X)\right)
\end{aligned}
\end{equation*}

### Simulation

We draw from the DGP defined above.

::::{.panel-tabset group="language"}

## R

```{r}
n <- 500
snr <- 3
x1 <- rnorm(n)
x2 <- rnorm(n)
x3 <- rnorm(n)
x4 <- as.numeric(rbinom(n, 1, 0.5))
x5 <- as.numeric(sample(1:3, n, replace = TRUE))
X <- cbind(x1, x2, x3, x4, x5)
p <- ncol(X)
mu_x <- mu2(X)
tau_x <- tau2(X)
pi_x <- 0.8 * pnorm((3 * mu_x / sd(mu_x)) - 0.5 * X[, 1]) + 0.05 + runif(n) / 10
Z <- rbinom(n, 1, pi_x)
E_XZ <- mu_x + Z * tau_x
y <- E_XZ + rnorm(n, 0, 1) * (sd(E_XZ) / snr)
X <- as.data.frame(X)
X$x4 <- factor(X$x4, ordered = TRUE)
X$x5 <- factor(X$x5, ordered = TRUE)

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct * n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- X[test_inds, ]
X_train <- X[train_inds, ]
pi_test <- pi_x[test_inds]
pi_train <- pi_x[train_inds]
Z_test <- Z[test_inds]
Z_train <- Z[train_inds]
y_test <- y[test_inds]
y_train <- y[train_inds]
mu_test <- mu_x[test_inds]
mu_train <- mu_x[train_inds]
tau_test <- tau_x[test_inds]
tau_train <- tau_x[train_inds]
```

## Python

```{python}
n = 500
snr = 3
x1 = rng.normal(size=n)
x2 = rng.normal(size=n)
x3 = rng.normal(size=n)
x4 = rng.binomial(1, 0.5, n).astype(float)
x5 = rng.choice([1, 2, 3], size=n).astype(float)
X = np.column_stack([x1, x2, x3, x4, x5])
mu_x = mu2(X)  # mu2 for Demo 2
tau_x = tau2(X)
pi_x = (0.8 * norm.cdf((3 * mu_x / np.std(mu_x)) - 0.5 * X[:, 0])
        + 0.05 + rng.uniform(size=n) / 10)
Z = rng.binomial(1, pi_x, n).astype(float)
E_XZ = mu_x + Z * tau_x
y = E_XZ + rng.normal(size=n) * (np.std(E_XZ) / snr)

X_df = pd.DataFrame({"x1": x1, "x2": x2, "x3": x3, "x4": x4, "x5": x5})
X_df["x4"] = pd.Categorical(X_df["x4"].astype(int), categories=[0, 1], ordered=True)
X_df["x5"] = pd.Categorical(X_df["x5"].astype(int), categories=[1, 2, 3], ordered=True)

test_set_pct = 0.2
n_test = round(test_set_pct * n)
n_train = n - n_test
test_inds = rng.choice(n, n_test, replace=False)
train_inds = np.setdiff1d(np.arange(n), test_inds)
X_test  = X_df.iloc[test_inds]
X_train = X_df.iloc[train_inds]
pi_test, pi_train   = pi_x[test_inds], pi_x[train_inds]
Z_test,  Z_train    = Z[test_inds],    Z[train_inds]
y_test,  y_train    = y[test_inds],    y[train_inds]
mu_test, mu_train   = mu_x[test_inds], mu_x[train_inds]
tau_test, tau_train = tau_x[test_inds], tau_x[train_inds]
```

::::

### Sampling and Analysis

#### Warmstart

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
general_params <- list(keep_every = 5)
prognostic_forest_params <- list(sample_sigma2_leaf = F)
treatment_effect_forest_params <- list(sample_sigma2_leaf = F)
bcf_model_warmstart <- bcf(
  X_train = X_train,
  Z_train = Z_train,
  y_train = y_train,
  propensity_train = pi_train,
  X_test = X_test,
  Z_test = Z_test,
  propensity_test = pi_test,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = general_params,
  prognostic_forest_params = prognostic_forest_params,
  treatment_effect_forest_params = treatment_effect_forest_params
)
```

## Python

```{python}
bcf_model_warmstart = BCFModel()
bcf_model_warmstart.sample(
    X_train=X_train, Z_train=Z_train, y_train=y_train, propensity_train=pi_train,
    X_test=X_test,   Z_test=Z_test,   propensity_test=pi_test,
    num_gfr=10, num_burnin=0, num_mcmc=100,
    general_params={"num_threads": 1, "keep_every": 5},
    prognostic_forest_params={"sample_sigma2_leaf": False},
    treatment_effect_forest_params={"sample_sigma2_leaf": False},
)
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
plot(
  rowMeans(bcf_model_warmstart$mu_hat_test),
  mu_test,
  xlab = "predicted",
  ylab = "actual",
  main = "Prognostic function"
)
abline(0, 1, col = "red", lty = 3, lwd = 3)
plot(
  rowMeans(bcf_model_warmstart$tau_hat_test),
  tau_test,
  xlab = "predicted",
  ylab = "actual",
  main = "Treatment effect"
)
abline(0, 1, col = "red", lty = 3, lwd = 3)
sigma_observed <- var(y - E_XZ)
plot_bounds <- c(
  min(c(bcf_model_warmstart$sigma2_global_samples, sigma_observed)),
  max(c(bcf_model_warmstart$sigma2_global_samples, sigma_observed))
)
plot(
  bcf_model_warmstart$sigma2_global_samples,
  ylim = plot_bounds,
  ylab = "sigma^2",
  xlab = "Sample",
  main = "Global variance parameter"
)
abline(h = sigma_observed, lty = 3, lwd = 3, col = "blue")
```

## Python

```{python}
sigma_observed = np.var(y - E_XZ)

mu_pred = bcf_model_warmstart.mu_hat_test.mean(axis=1)
lo, hi = min(mu_pred.min(), mu_test.min()), max(mu_pred.max(), mu_test.max())
plt.scatter(mu_pred, mu_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Prognostic function")
plt.show()

tau_pred = bcf_model_warmstart.tau_hat_test.mean(axis=1)
lo, hi = min(tau_pred.min(), tau_test.min()), max(tau_pred.max(), tau_test.max())
plt.scatter(tau_pred, tau_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Treatment effect")
plt.show()

plt.plot(bcf_model_warmstart.global_var_samples)
plt.axhline(sigma_observed, color="blue", linestyle="dashed", linewidth=2)
plt.xlabel("Sample"); plt.ylabel(r"$\sigma^2$"); plt.title("Global variance parameter")
plt.show()
```

::::

Examine test set interval coverage.

::::{.panel-tabset group="language"}

## R

```{r}
test_lb <- apply(bcf_model_warmstart$tau_hat_test, 1, quantile, 0.025)
test_ub <- apply(bcf_model_warmstart$tau_hat_test, 1, quantile, 0.975)
cover <- ((test_lb <= tau_x[test_inds]) &
  (test_ub >= tau_x[test_inds]))
mean(cover)
```

## Python

```{python}
test_lb = np.quantile(bcf_model_warmstart.tau_hat_test, 0.025, axis=1)
test_ub = np.quantile(bcf_model_warmstart.tau_hat_test, 0.975, axis=1)
cover = (test_lb <= tau_test) & (test_ub >= tau_test)
print(f"Coverage: {cover.mean():.3f}")
```

::::

#### BART MCMC without Warmstart

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 0
num_burnin <- 2000
num_mcmc <- 100
general_params <- list(keep_every = 5)
prognostic_forest_params <- list(sample_sigma2_leaf = F)
treatment_effect_forest_params <- list(sample_sigma2_leaf = F)
bcf_model_root <- bcf(
  X_train = X_train,
  Z_train = Z_train,
  y_train = y_train,
  propensity_train = pi_train,
  X_test = X_test,
  Z_test = Z_test,
  propensity_test = pi_test,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = general_params,
  prognostic_forest_params = prognostic_forest_params,
  treatment_effect_forest_params = treatment_effect_forest_params
)
```

## Python

```{python}
bcf_model_root = BCFModel()
bcf_model_root.sample(
    X_train=X_train, Z_train=Z_train, y_train=y_train, propensity_train=pi_train,
    X_test=X_test,   Z_test=Z_test,   propensity_test=pi_test,
    num_gfr=0, num_burnin=2000, num_mcmc=100,
    general_params={"num_threads": 1, "keep_every": 5},
    prognostic_forest_params={"sample_sigma2_leaf": False},
    treatment_effect_forest_params={"sample_sigma2_leaf": False},
)
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
plot(
  rowMeans(bcf_model_root$mu_hat_test),
  mu_test,
  xlab = "predicted",
  ylab = "actual",
  main = "Prognostic function"
)
abline(0, 1, col = "red", lty = 3, lwd = 3)
plot(
  rowMeans(bcf_model_root$tau_hat_test),
  tau_test,
  xlab = "predicted",
  ylab = "actual",
  main = "Treatment effect"
)
abline(0, 1, col = "red", lty = 3, lwd = 3)
sigma_observed <- var(y - E_XZ)
plot_bounds <- c(
  min(c(bcf_model_root$sigma2_global_samples, sigma_observed)),
  max(c(bcf_model_root$sigma2_global_samples, sigma_observed))
)
plot(
  bcf_model_root$sigma2_global_samples,
  ylim = plot_bounds,
  ylab = "sigma^2",
  xlab = "Sample",
  main = "Global variance parameter"
)
abline(h = sigma_observed, lty = 3, lwd = 3, col = "blue")
```

## Python

```{python}
mu_pred = bcf_model_root.mu_hat_test.mean(axis=1)
lo, hi = min(mu_pred.min(), mu_test.min()), max(mu_pred.max(), mu_test.max())
plt.scatter(mu_pred, mu_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Prognostic function")
plt.show()

tau_pred = bcf_model_root.tau_hat_test.mean(axis=1)
lo, hi = min(tau_pred.min(), tau_test.min()), max(tau_pred.max(), tau_test.max())
plt.scatter(tau_pred, tau_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Treatment effect")
plt.show()

plt.plot(bcf_model_root.global_var_samples)
plt.axhline(sigma_observed, color="blue", linestyle="dashed", linewidth=2)
plt.xlabel("Sample"); plt.ylabel(r"$\sigma^2$"); plt.title("Global variance parameter")
plt.show()
```

::::

Examine test set interval coverage.

::::{.panel-tabset group="language"}

## R

```{r}
test_lb <- apply(bcf_model_root$tau_hat_test, 1, quantile, 0.025)
test_ub <- apply(bcf_model_root$tau_hat_test, 1, quantile, 0.975)
cover <- ((test_lb <= tau_x[test_inds]) &
  (test_ub >= tau_x[test_inds]))
mean(cover)
```

## Python

```{python}
test_lb = np.quantile(bcf_model_root.tau_hat_test, 0.025, axis=1)
test_ub = np.quantile(bcf_model_root.tau_hat_test, 0.975, axis=1)
cover = (test_lb <= tau_test) & (test_ub >= tau_test)
print(f"Coverage: {cover.mean():.3f}")
```

::::

# References
