---
title: "Calibrating StochTree Model Parameters"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
code-annotations: hover
---

```{r}
#| include: false
Sys.setenv(OMP_NUM_THREADS = "1")
Sys.setenv(OPENBLAS_NUM_THREADS = "1")
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(
      rprojroot::find_root(rprojroot::has_file(".here")),
      ".venv",
      "bin",
      "python"
    )
  ),
  required = TRUE
)
```

# Introduction

`stochtree` exposes many user-facing modeling parameters. Even for a simple BART model, users can specify

1. how many trees in the forest, 
2. prior split parameters $\alpha$ and $\beta$, 
3. how to initialize the leaf scale parameter and whether to sample it,
4. max depth of each tree (default: unlimited), and 
5. minimum size of a leaf node.

We have made every effort to set experience-informed defaults or calibration procedures for each of these options so that stochtree with largely work "out of the box," but it's nonetheless reasonable to wonder how good these parameters are for a given dataset. 

In this vignette, we'll walkthrough a simple (scaled-down) example of how one might go about evaluating (and updating) their BART model's hyperparameters.

# Framework

Broadly, our procedure is

<style>.algorithm p { margin: 0.1em 0; line-height: 1.7; }</style>

::: {.algorithm style="border-top: 2px solid; border-bottom: 2px solid; padding: 0.6em 1em; margin: 1.5em 0;"}

Define a data-generating process (DGP), with careful attention to: (1) sample size ($n$), (2) feature dimensionality ($p$), (3) outcome mean function (sparsity of features, nonlinearity), and (4) signal-to-noise ratio. These aspects should be chosen based on their resemblance to a dataset of interest.

Pick a number, $mc$, of replicates to draw of this DGP. Fix BART hyperparameters $\alpha$, $\beta$, $m$ trees, max depth, min samples in a leaf, etc...

Specify a range of interval probabilities (i.e. $r = \left\{0.50, 0.75, 0.90, 0.95\right\}$)

For $i$ in $\left\{1,\dots,mc\right\}$:

::::: {style="margin-left: 2em"}

Draw a dataset $y, X$ from the DGP defined above

Split the dataset into train and test sets

Sample a BART model of $y \sim \mathcal{N}(f(X),\sigma^2)$ with hyperparameters $\alpha$, $\beta$, $m$, ....

Compute test set coverage of true $f(X)$ for intervals defined by the probabilities in $r$

:::::

Compute average coverage across $mc$ replications for each probability in $r$

Plot average coverage against expected coverage $r$

:::

This plot should be roughly diagonal for a well-calibrated BART model, so we can use it to evaluate and refine hyperparameters.

# Demo

We now illustrate this procedure with a simple continuous outcome BART demo. We start by importing the necessary packages.

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
```

## Python

```{python}
from typing import Union
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from stochtree import BARTModel
```

::::

Set a seed for reproducibility

::::{.panel-tabset group="language"}

## R

```{r}
random_seed = 1234
set.seed(random_seed)
```

## Python

```{python}
random_seed = 1234
rng = np.random.default_rng(random_seed)
```

::::

## Data Generation

Let's define a sparse, nonlinear DGP with additive Gaussian errors

$$
\begin{aligned}
y &\sim \mathcal{N}\left(f(X), \sigma^2\right)\\
X_1, \dots, X_p &\sim \text{U}\left(0,1\right), \quad p > 5\\
f(X) &= 5 \sin(2 \pi X_1) + g(X_2, X_3) - 2 X_4 X_5\\
g(X_2, X_3) &= \begin{cases}
-3 & \text{if } X_2 \leq 0.5 \text{ and } X_3 \leq 0.5 \\
-1 & \text{if } X_2 \leq 0.5 \text{ and } X_3 > 0.5 \\
-1 & \text{if } X_2 > 0.5 \text{ and } X_3 \leq 0.5 \\
3 & \text{if } X_2 > 0.5 \text{ and } X_3 > 0.5 \\
\end{cases}\\
\sigma^2 &= \text{Var}(f(X)) / a^2
\end{aligned}
$$

where $a$ is a scaling factor that determines the signal-to-noise ratio.

::::{.panel-tabset group="language"}

## R

```{r}
continuous_dgp <- function(X) {
  g <- function(x) {
    ifelse(
      x[, 2] <= 0.5 & x[, 3] <= 0.5,
      -3,
      ifelse(
        x[, 2] <= 0.5 & x[, 3] > 0.5,
        -1,
        ifelse(x[, 2] > 0.5 & x[, 3] <= 0.5, -1, 3)
      )
    )
  }
  5 * sin(2 * pi * X[, 1]) + g(X) - 2 * X[, 4] * X[, 5]
}

generate_data <- function(n, p, snr, mean_fn) {
  if (p < 5) {
    stop("p must be at least 5")
  }
  if (n < 1) {
    stop("n must be at least 1")
  }
  if (snr <= 0) {
    stop("snr must be positive")
  }
  X <- matrix(runif(n * p), ncol = p)
  f_X <- mean_fn(X)
  eps <- rnorm(n, mean = 0, sd = sqrt(var(f_X) / snr))
  y <- f_X + eps
  list(X = X, f_X = f_X, y = y)
}
```

## Python

```{python}
def continuous_dgp(X):
    def g(x):
        return np.where(
            (x[:, 1] <= 0.5) & (x[:, 2] <= 0.5),
            -3,
            np.where(
                (x[:, 1] <= 0.5) & (x[:, 2] > 0.5),
                -1,
                np.where((x[:, 1] > 0.5) & (x[:, 2] <= 0.5), -1, 3),
            ),
        )

    return 5 * np.sin(2 * np.pi * X[:, 0]) + g(X) - 2 * X[:, 3] * X[:, 4]


def generate_data(n, p, snr, mean_fn):
    if p < 5:
        raise ValueError("p must be at least 5")
    if n < 1:
        raise ValueError("n must be at least 1")
    if snr <= 0:
        raise ValueError("snr must be positive")
    X = rng.uniform(size=(n, p))
    f_X = mean_fn(X)
    eps = rng.normal(loc=0, scale=np.sqrt(np.var(f_X) / snr), size=n)
    y = f_X + eps
    return X, f_X, y
```

::::

And we define helper functions to perform train/test splits

::::{.panel-tabset group="language"}

## R

```{r}
compute_test_train_indices <- function(n, test_set_pct) {
  n_test <- round(test_set_pct * n)
  n_train <- n - n_test
  test_inds <- sort(sample(1:n, n_test, replace = FALSE))
  train_inds <- (1:n)[!((1:n) %in% test_inds)]
  return(list(test_inds = test_inds, train_inds = train_inds))
}
subset_data <- function(data, subset_inds) {
  if (is.matrix(data)) {
    return(data[subset_inds, ])
  } else {
    return(data[subset_inds])
  }
}
```

## Python

```{python}
def compute_test_train_indices(n, test_set_pct):
    n_test = round(test_set_pct * n)
    test_inds = np.sort(rng.choice(n, size=n_test, replace=False))
    train_inds = np.setdiff1d(np.arange(n), test_inds)
    return test_inds, train_inds


def subset_data(
    data: Union[np.array, pd.DataFrame], subset_inds: np.array
) -> Union[np.array, pd.DataFrame]:
    if data.ndim == 2:
        if isinstance(data, np.ndarray):
            return data[subset_inds, :]
        else:
            return data.iloc[subset_inds, :]
    else:
        return data[subset_inds]
```

::::

## Simulation

Now, we run the simulation procedure described above to assess the calibration of the model.
Let's define a helper function that runs the procedure end-to-end, producing a vector of coverage results for each interval level.

::::{.panel-tabset group="language"}

## R

```{r}
run_coverage_simulation <- function(
  n,
  p,
  snr,
  num_gfr,
  num_burnin,
  num_mcmc,
  num_trees,
  max_depth,
  mc,
  interval_levels,
  mean_fn
) {
  # Simulation
  coverage_results <- matrix(
    NA_real_,
    nrow = mc,
    ncol = length(interval_levels)
  )
  for (i in 1:mc) {
    # Generate data
    data <- generate_data(n, p, snr, mean_fn = mean_fn)
    X <- data$X
    f_X <- data$f_X
    y <- data$y

    # Split into train and test sets
    indices <- compute_test_train_indices(n, test_set_pct = 0.2)
    test_inds <- indices$test_inds
    train_inds <- indices$train_inds
    X_train <- subset_data(X, train_inds)
    X_test <- subset_data(X, test_inds)
    f_X_train <- subset_data(f_X, train_inds)
    f_X_test <- subset_data(f_X, test_inds)
    y_train <- subset_data(y, train_inds)
    y_test <- subset_data(y, test_inds)

    # Fit BART
    bart_model <- bart(
      X_train = X_train,
      y_train = y_train,
      num_gfr = num_gfr,
      num_burnin = num_burnin,
      num_mcmc = num_mcmc,
      general_params = list(random_seed = random_seed + i),
      mean_forest_params = list(num_trees = num_trees, max_depth = max_depth)
    )

    # Compute intervals
    level_idx <- 1
    for (level in interval_levels) {
      # Compute intervals for each level
      yhat_test_interval <- computeBARTPosteriorInterval(
        bart_model,
        level = level,
        terms = "y_hat",
        X = X_test
      )
      yhat_test_interval_lb <- yhat_test_interval$lower
      yhat_test_interval_ub <- yhat_test_interval$upper
      # Compute average mean function coverage
      cover <- mean(
        f_X_test >= yhat_test_interval_lb & f_X_test <= yhat_test_interval_ub
      )
      coverage_results[i, level_idx] <- cover
      level_idx <- level_idx + 1
    }
  }
  return(apply(coverage_results, 2, mean))
}
```

## Python

```{python}
def run_coverage_simulation(
    n,
    p,
    snr,
    num_gfr,
    num_burnin,
    num_mcmc,
    num_trees,
    max_depth,
    mc,
    interval_levels,
    mean_fn,
):
    coverage_results = np.empty((mc, len(interval_levels)))

    # Simulation
    for i in range(mc):
        # Generate data
        X, f_X, y = generate_data(n, p, snr, mean_fn=mean_fn)

        # Split into train and test sets
        test_inds, train_inds = compute_test_train_indices(n, test_set_pct=0.2)
        X_train = subset_data(X, train_inds)
        X_test = subset_data(X, test_inds)
        f_X_train = subset_data(f_X, train_inds)
        f_X_test = subset_data(f_X, test_inds)
        y_train = subset_data(y, train_inds)
        y_test = subset_data(y, test_inds)

        # Fit BART
        bart_model = BARTModel()
        bart_model.sample(
            X_train=X_train,
            y_train=y_train,
            num_gfr=num_gfr,
            num_burnin=num_burnin,
            num_mcmc=num_mcmc,
            general_params={"random_seed": random_seed + i},
            mean_forest_params={"num_trees": num_trees, "max_depth": max_depth},
        )

        # Compute intervals
        for level_idx, level in enumerate(interval_levels):
            # Compute intervals for each level
            yhat_test_interval = bart_model.compute_posterior_interval(
                level=level, terms="y_hat", X=X_test
            )
            yhat_test_interval_lb = yhat_test_interval["lower"]
            yhat_test_interval_ub = yhat_test_interval["upper"]
            # Compute average mean function coverage
            cover = np.mean(
                (f_X_test >= yhat_test_interval_lb)
                & (f_X_test <= yhat_test_interval_ub)
            )
            coverage_results[i, level_idx] = cover

    return np.mean(coverage_results, axis=0)
```

::::

We start with a simple BART model with 10 decision trees each with a maximum depth of 3.

::::{.panel-tabset group="language"}

## R

```{r}
# Data generating process parameters
n <- 250
p <- 50
snr <- 2

# Modeling parameters
num_gfr <- 5
num_burnin <- 100
num_mcmc <- 500
num_trees <- 10
max_depth <- 3

# Simulation setup
mc <- 50
interval_levels <- c(0.5, 0.6, 0.7, 0.8, 0.9, 0.95)

# Run simulation
average_coverage_rates <- run_coverage_simulation(
  n,
  p,
  snr,
  num_gfr,
  num_burnin,
  num_mcmc,
  num_trees,
  max_depth,
  mc,
  interval_levels,
  mean_fn = continuous_dgp
)
```

## Python

```{python}
# Data generating process parameters
n = 250
p = 50
snr = 2

# Modeling parameters
num_gfr = 5
num_burnin = 100
num_mcmc = 500
num_trees = 10
max_depth = 3

# Simulation setup
mc = 50
interval_levels = [0.5, 0.6, 0.7, 0.8, 0.9, 0.95]

# Run simulation
average_coverage_rates = run_coverage_simulation(
    n,
    p,
    snr,
    num_gfr,
    num_burnin,
    num_mcmc,
    num_trees,
    max_depth,
    mc,
    interval_levels,
    mean_fn=continuous_dgp,
)
```

::::

Now, we plot the average observed coverage against the nominal coverage level

::::{.panel-tabset group="language"}

## R

```{r}
plot(
  interval_levels,
  average_coverage_rates,
  type = "b",
  xlab = "Nominal Coverage Level",
  ylab = "Average Observed Coverage",
  main = "Calibration Plot"
)
abline(0, 1, lty = 2)
```

## Python

```{python}
fig, ax = plt.subplots()
ax.plot(interval_levels, average_coverage_rates, marker='o')
ax.set_xlabel("Nominal Coverage Level")
ax.set_ylabel("Average Observed Coverage")
ax.set_title("Calibration Plot")
lims = [max(ax.get_xlim()[0], ax.get_ylim()[0]), min(ax.get_xlim()[1], ax.get_ylim()[1])]
ax.plot(lims, lims, linestyle='--', color='gray')
plt.show()
```

::::

Clearly, this is quite off. The posterior intervals are too narrow at every nominal coverage probability. Let's try increasing the number of trees to 200 and the max depth to 10.

::::{.panel-tabset group="language"}

## R

```{r}
# Data generating process parameters
n <- 250
p <- 50
snr <- 2

# Modeling parameters
num_gfr <- 5
num_burnin <- 100
num_mcmc <- 500
num_trees <- 200
max_depth <- 10

# Simulation setup
mc <- 50
interval_levels <- c(0.5, 0.6, 0.7, 0.8, 0.9, 0.95)

# Run simulation
average_coverage_rates <- run_coverage_simulation(
  n,
  p,
  snr,
  num_gfr,
  num_burnin,
  num_mcmc,
  num_trees,
  max_depth,
  mc,
  interval_levels,
  mean_fn = continuous_dgp
)
```

## Python

```{python}
# Data generating process parameters
n = 250
p = 50
snr = 2

# Modeling parameters
num_gfr = 5
num_burnin = 100
num_mcmc = 500
num_trees = 200
max_depth = 10

# Simulation setup
mc = 50
interval_levels = [0.5, 0.6, 0.7, 0.8, 0.9, 0.95]

# Run simulation
average_coverage_rates = run_coverage_simulation(
    n,
    p,
    snr,
    num_gfr,
    num_burnin,
    num_mcmc,
    num_trees,
    max_depth,
    mc,
    interval_levels,
    mean_fn=continuous_dgp,
)
```

::::

Now, we plot the average observed coverage against the nominal coverage level

::::{.panel-tabset group="language"}

## R

```{r}
plot(
  interval_levels,
  average_coverage_rates,
  type = "b",
  xlab = "Nominal Coverage Level",
  ylab = "Average Observed Coverage",
  main = "Calibration Plot"
)
abline(0, 1, lty = 2)
```

## Python

```{python}
fig, ax = plt.subplots()
ax.plot(interval_levels, average_coverage_rates, marker='o')
ax.set_xlabel("Nominal Coverage Level")
ax.set_ylabel("Average Observed Coverage")
ax.set_title("Calibration Plot")
lims = [max(ax.get_xlim()[0], ax.get_ylim()[0]), min(ax.get_xlim()[1], ax.get_ylim()[1])]
ax.plot(lims, lims, linestyle='--', color='gray')
plt.show()
```

::::

This is much better but now we have the opposite problem — intervals are slightly conservative (wider than their nominal coverage probability). Let's try reducing the number of trees to 50.

::::{.panel-tabset group="language"}

## R

```{r}
# Data generating process parameters
n <- 250
p <- 50
snr <- 2

# Modeling parameters
num_gfr <- 5
num_burnin <- 100
num_mcmc <- 500
num_trees <- 50
max_depth <- 10

# Simulation setup
mc <- 50
interval_levels <- c(0.5, 0.6, 0.7, 0.8, 0.9, 0.95)

# Run simulation
average_coverage_rates <- run_coverage_simulation(
  n,
  p,
  snr,
  num_gfr,
  num_burnin,
  num_mcmc,
  num_trees,
  max_depth,
  mc,
  interval_levels,
  mean_fn = continuous_dgp
)
```

## Python

```{python}
# Data generating process parameters
n = 250
p = 50
snr = 2

# Modeling parameters
num_gfr = 5
num_burnin = 100
num_mcmc = 500
num_trees = 50
max_depth = 10

# Simulation setup
mc = 50
interval_levels = [0.5, 0.6, 0.7, 0.8, 0.9, 0.95]

# Run simulation
average_coverage_rates = run_coverage_simulation(
    n,
    p,
    snr,
    num_gfr,
    num_burnin,
    num_mcmc,
    num_trees,
    max_depth,
    mc,
    interval_levels,
    mean_fn=continuous_dgp,
)
```

::::

Now, we plot the average observed coverage against the nominal coverage level

::::{.panel-tabset group="language"}

## R

```{r}
plot(
  interval_levels,
  average_coverage_rates,
  type = "b",
  xlab = "Nominal Coverage Level",
  ylab = "Average Observed Coverage",
  main = "Calibration Plot"
)
abline(0, 1, lty = 2)
```

## Python

```{python}
fig, ax = plt.subplots()
ax.plot(interval_levels, average_coverage_rates, marker='o')
ax.set_xlabel("Nominal Coverage Level")
ax.set_ylabel("Average Observed Coverage")
ax.set_title("Calibration Plot")
lims = [max(ax.get_xlim()[0], ax.get_ylim()[0]), min(ax.get_xlim()[1], ax.get_ylim()[1])]
ax.plot(lims, lims, linestyle='--', color='gray')
plt.show()
```

::::

With 50 trees, coverage is nearly perfect — this configuration is well-calibrated for this DGP.

This is a relatively straightforward demo of how we can calibrate priors based on a target data generating process.
