---
title: "Custom Sampling Routine"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(here::here(), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

While the functions `bart()` and `bcf()` provide simple and performant interfaces for
supervised learning / causal inference, `stochtree` also offers access to many of the
"low-level" data structures that are typically implemented in C++. This low-level
interface is not designed for performance or even simplicity — rather the intent is
to provide a "prototype" interface to the C++ code that doesn't require modifying any
C++.

# Motivation

To illustrate when such a prototype interface might be useful, consider the classic
BART algorithm:

|    **INPUT**: $y$, $X$, $\tau$, $\nu$, $\lambda$, $\alpha$, $\beta$
|    **OUTPUT**: $m$ samples of a decision forest with $k$ trees and global variance parameter $\sigma^2$
|    Initialize $\sigma^2$ via a default or a data-dependent calibration exercise
|    Initialize "forest 0" with $k$ trees with a single root node, referring to tree $j$'s prediction vector as $f_{0,j}$
|    Compute residual as $r = y - \sum_{j=1}^k f_{0,j}$
|    **FOR** $i$ **IN** $\left\{1,\dots,m\right\}$:
|       Initialize forest $i$ from forest $i-1$
|       **FOR** $j$ **IN** $\left\{1,\dots,k\right\}$:
|          Add predictions for tree $j$ to residual: $r = r + f_{i,j}$
|          Sample tree $j$ of forest $i$ from $p\left(\mathcal{T}_{i,j} \mid r, \sigma^2\right)$
|          Update residual by removing tree $j$'s predictions: $r = r - f_{i,j}$
|       Sample $\sigma^2$ from $p\left(\sigma^2 \mid r\right)$
|    **RETURN** forests $\left\{1,\dots,m\right\}$ and $\sigma^2$ samples

This algorithm is implemented in the `bart()` function, but the low-level interface
allows you to customize this loop — for example, to add random effects, modify the
variance model, or implement a novel sampling scheme.

# Setup

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
```

## Python

```{python}
import numpy as np
import matplotlib.pyplot as plt
from stochtree import (
    RNG, Dataset, Forest, ForestContainer, ForestSampler,
    GlobalVarianceModel, LeafVarianceModel, Residual,
    ForestModelConfig, GlobalModelConfig,
)
```

::::

# Supervised Learning Demo

## Simulation

Simulate a simple partitioned linear model.

::::{.panel-tabset group="language"}

## R

```{r}
# Generate the data
n <- 500
p_X <- 10
p_W <- 1
X <- matrix(runif(n*p_X), ncol = p_X)
W <- matrix(runif(n*p_W), ncol = p_W)
f_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (-3*W[,1]) +
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (-1*W[,1]) +
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (1*W[,1]) +
    ((0.75 <= X[,1]) & (1 > X[,1])) * (3*W[,1])
)
y <- f_XW + rnorm(n, 0, 1)

# Standardize outcome
y_bar <- mean(y)
y_std <- sd(y)
resid <- (y-y_bar)/y_std
```

## Python

```{python}
random_seed = 1234
rng = np.random.default_rng(random_seed)

n = 500
p_X = 10
p_W = 1
X = rng.uniform(size=(n, p_X))
W = rng.uniform(size=(n, p_W))
# R uses X[,1] (1-indexed) = Python X[:,0]
f_XW = (
    ((X[:, 0] >= 0)    & (X[:, 0] < 0.25)) * (-3 * W[:, 0]) +
    ((X[:, 0] >= 0.25) & (X[:, 0] < 0.5))  * (-1 * W[:, 0]) +
    ((X[:, 0] >= 0.5)  & (X[:, 0] < 0.75)) * ( 1 * W[:, 0]) +
    ((X[:, 0] >= 0.75) & (X[:, 0] < 1.0))  * ( 3 * W[:, 0])
)
y = f_XW + rng.standard_normal(n)

# Standardize outcome
y_bar = np.mean(y)
y_std = np.std(y)
resid = (y - y_bar) / y_std
```

::::

## Sampling

Set some parameters that inform the forest and variance parameter samplers.

::::{.panel-tabset group="language"}

## R

```{r}
alpha <- 0.9
beta <- 1.25
min_samples_leaf <- 1
max_depth <- 10
num_trees <- 100
cutpoint_grid_size <- 100
global_variance_init <- 1.
current_sigma2 <- global_variance_init
tau_init <- 1/num_trees
leaf_prior_scale <- as.matrix(ifelse(p_W >= 1, diag(tau_init, p_W), diag(tau_init, 1)))
nu <- 4
lambda <- 0.5
a_leaf <- 2.
b_leaf <- 0.5
leaf_regression <- T
feature_types <- as.integer(rep(0, p_X)) # 0 = numeric
var_weights <- rep(1/p_X, p_X)
```

## Python

```{python}
alpha = 0.9
beta = 1.25
min_samples_leaf = 1
max_depth = 10
num_trees = 100
cutpoint_grid_size = 100
global_variance_init = 1.0
tau_init = 1.0 / num_trees
leaf_prior_scale = np.array([[tau_init]], order="C")
a_global = 4.0
b_global = 2.0
a_leaf = 2.0
b_leaf = 0.5
leaf_regression = True
feature_types = np.zeros(p_X, dtype=int)  # 0 = numeric
var_weights = np.full(p_X, 1.0 / p_X)
leaf_model = 1 if leaf_regression else 0  # 1 = linear leaf (with basis)
leaf_dimension = p_W if leaf_regression else 1
```

::::

Initialize R-level access to the C++ classes needed to sample our model.

::::{.panel-tabset group="language"}

## R

```{r}
# Data
if (leaf_regression) {
    forest_dataset <- createForestDataset(X, W)
    outcome_model_type <- 1
    leaf_dimension <- p_W
} else {
    forest_dataset <- createForestDataset(X)
    outcome_model_type <- 0
    leaf_dimension <- 1
}
outcome <- createOutcome(resid)

# Random number generator (std::mt19937)
rng <- createCppRNG()

# Sampling data structures
forest_model_config <- createForestModelConfig(
    feature_types = feature_types, num_trees = num_trees, num_features = p_X,
    num_observations = n, variable_weights = var_weights, leaf_dimension = leaf_dimension,
    alpha = alpha, beta = beta, min_samples_leaf = min_samples_leaf, max_depth = max_depth,
    leaf_model_type = outcome_model_type, leaf_model_scale = leaf_prior_scale,
    cutpoint_grid_size = cutpoint_grid_size
)
global_model_config <- createGlobalModelConfig(global_error_variance = global_variance_init)
forest_model <- createForestModel(forest_dataset, forest_model_config, global_model_config)

# "Active forest" (which gets updated by the sample) and
# container of forest samples (which is written to when
# a sample is not discarded due to burn-in / thinning)
if (leaf_regression) {
    forest_samples <- createForestSamples(num_trees, 1, F)
    active_forest <- createForest(num_trees, 1, F)
} else {
    forest_samples <- createForestSamples(num_trees, 1, T)
    active_forest <- createForest(num_trees, 1, T)
}

# Initialize the leaves of each tree in the forest
active_forest$prepare_for_sampler(forest_dataset, outcome, forest_model, outcome_model_type, mean(resid))
active_forest$adjust_residual(forest_dataset, outcome, forest_model, ifelse(outcome_model_type==1, T, F), F)
```

## Python

```{python}
# Dataset: covariates + basis
dataset = Dataset()
dataset.add_covariates(X)
dataset.add_basis(W)

# Residual
residual = Residual(resid)

# Random number generator
cpp_rng = RNG(random_seed)

# Forest container (stores samples) and active forest (updated each iteration)
forest_container = ForestContainer(num_trees, leaf_dimension, False, False)
active_forest = Forest(num_trees, leaf_dimension, False, False)

# Sampler config
global_model_config = GlobalModelConfig(global_error_variance=global_variance_init)
forest_model_config = ForestModelConfig(
    num_trees=num_trees,
    num_features=p_X,
    num_observations=n,
    feature_types=feature_types,
    variable_weights=var_weights,
    leaf_dimension=leaf_dimension,
    alpha=alpha,
    beta=beta,
    min_samples_leaf=min_samples_leaf,
    max_depth=max_depth,
    leaf_model_type=leaf_model,
    leaf_model_scale=leaf_prior_scale,
    cutpoint_grid_size=cutpoint_grid_size,
)
forest_sampler = ForestSampler(dataset, global_model_config, forest_model_config)
global_var_model = GlobalVarianceModel()
leaf_var_model = LeafVarianceModel()

# Initialize leaves of each tree
forest_init_val = np.zeros(leaf_dimension)
forest_sampler.prepare_for_sampler(dataset, residual, active_forest, leaf_model, forest_init_val)
```

::::

Prepare to run the sampler.

::::{.panel-tabset group="language"}

## R

```{r}
num_warmstart <- 10
num_mcmc <- 100
num_samples <- num_warmstart + num_mcmc
global_var_samples <- c(global_variance_init, rep(0, num_samples))
leaf_scale_samples <- c(tau_init, rep(0, num_samples))
```

## Python

```{python}
num_warmstart = 10
num_mcmc = 100
num_samples = num_warmstart + num_mcmc
global_var_samples = np.concatenate([[global_variance_init], np.zeros(num_samples)])
leaf_scale_samples = np.concatenate([[tau_init], np.zeros(num_samples)])
```

::::

Run the grow-from-root sampler to "warm-start" BART (@he2023stochastic).

::::{.panel-tabset group="language"}

## R

```{r}
for (i in 1:num_warmstart) {
    # Sample forest
    forest_model$sample_one_iteration(
        forest_dataset, outcome, forest_samples, active_forest, rng,
        forest_model_config, global_model_config, keep_forest = T, gfr = T
    )

    # Sample global variance parameter
    current_sigma2 <- sampleGlobalErrorVarianceOneIteration(
        outcome, forest_dataset, rng, nu, lambda
    )
    global_var_samples[i+1] <- current_sigma2
    global_model_config$update_global_error_variance(current_sigma2)

    # Sample leaf node variance parameter and update `leaf_prior_scale`
    leaf_scale_samples[i+1] <- sampleLeafVarianceOneIteration(
        active_forest, rng, a_leaf, b_leaf
    )
    leaf_prior_scale[1,1] <- leaf_scale_samples[i+1]
    forest_model_config$update_leaf_model_scale(leaf_prior_scale)
}
```

## Python

```{python}
for i in range(num_warmstart):
    forest_sampler.sample_one_iteration(
        forest_container, active_forest, dataset, residual, cpp_rng,
        global_model_config, forest_model_config, True, True, 1,  # keep_forest=True, gfr=True, num_threads=1
    )
    current_sigma2 = global_var_model.sample_one_iteration(residual, cpp_rng, a_global, b_global)
    global_var_samples[i + 1] = current_sigma2
    leaf_scale_samples[i + 1] = leaf_var_model.sample_one_iteration(active_forest, cpp_rng, a_leaf, b_leaf)
    leaf_prior_scale[0, 0] = leaf_scale_samples[i + 1]
    forest_model_config.update_leaf_model_scale(leaf_prior_scale)
```

::::

Pick up from the last GFR forest (and associated global variance / leaf scale
parameters) with an MCMC sampler.

::::{.panel-tabset group="language"}

## R

```{r}
for (i in (num_warmstart+1):num_samples) {
    # Sample forest
    forest_model$sample_one_iteration(
        forest_dataset, outcome, forest_samples, active_forest, rng,
        forest_model_config, global_model_config, keep_forest = T, gfr = F
    )

    # Sample global variance parameter
    current_sigma2 <- sampleGlobalErrorVarianceOneIteration(
        outcome, forest_dataset, rng, nu, lambda
    )
    global_var_samples[i+1] <- current_sigma2
    global_model_config$update_global_error_variance(current_sigma2)

    # Sample leaf node variance parameter and update `leaf_prior_scale`
    leaf_scale_samples[i+1] <- sampleLeafVarianceOneIteration(
        active_forest, rng, a_leaf, b_leaf
    )
    leaf_prior_scale[1,1] <- leaf_scale_samples[i+1]
    forest_model_config$update_leaf_model_scale(leaf_prior_scale)
}
```

## Python

```{python}
for i in range(num_warmstart, num_samples):
    forest_sampler.sample_one_iteration(
        forest_container, active_forest, dataset, residual, cpp_rng,
        global_model_config, forest_model_config, True, False, 1,  # keep_forest=True, gfr=False, num_threads=1
    )
    current_sigma2 = global_var_model.sample_one_iteration(residual, cpp_rng, a_global, b_global)
    global_var_samples[i + 1] = current_sigma2
    leaf_scale_samples[i + 1] = leaf_var_model.sample_one_iteration(active_forest, cpp_rng, a_leaf, b_leaf)
    leaf_prior_scale[0, 0] = leaf_scale_samples[i + 1]
    forest_model_config.update_leaf_model_scale(leaf_prior_scale)
```

::::

Predict and rescale samples.

::::{.panel-tabset group="language"}

## R

```{r}
# Forest predictions
preds <- forest_samples$predict(forest_dataset)*y_std + y_bar

# Global error variance
sigma_samples <- sqrt(global_var_samples)*y_std
```

## Python

```{python}
# Forest predictions: shape (n, num_samples); rescale to original y scale
preds = forest_container.predict(dataset) * y_std + y_bar

# Global error variance (sigma, not sigma^2)
sigma_samples = np.sqrt(global_var_samples) * y_std
```

::::

## Results

Inspect the initial samples obtained via "grow-from-root" (@he2023stochastic).

::::{.panel-tabset group="language"}

## R

```{r}
plot(sigma_samples[1:num_warmstart], ylab="sigma")
plot(rowMeans(preds[,1:num_warmstart]), y, pch=16,
     cex=0.75, xlab = "pred", ylab = "actual")
abline(0,1,col="red",lty=2,lwd=2.5)
```

## Python

```{python}
gfr_preds = preds[:, :num_warmstart]
fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 4))
ax1.plot(sigma_samples[1:num_warmstart + 1])
ax1.set_ylabel("sigma")
ax1.set_title("GFR: Global Error Scale")
ax2.scatter(gfr_preds.mean(axis=1), y, s=10, alpha=0.5)
lo = min(gfr_preds.mean(axis=1).min(), y.min())
hi = max(gfr_preds.mean(axis=1).max(), y.max())
ax2.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
ax2.set_xlabel("pred")
ax2.set_ylabel("actual")
ax2.set_title("GFR: Predicted vs Actual")
plt.tight_layout()
plt.show()
```

::::

Inspect the BART samples obtained after "warm-starting".

::::{.panel-tabset group="language"}

## R

```{r}
plot(sigma_samples[(num_warmstart+1):num_samples], ylab="sigma")
plot(rowMeans(preds[,(num_warmstart+1):num_samples]), y, pch=16,
     cex=0.75, xlab = "pred", ylab = "actual")
abline(0,1,col="red",lty=2,lwd=2.5)
```

## Python

```{python}
mcmc_preds = preds[:, num_warmstart:num_samples]
fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 4))
ax1.plot(sigma_samples[num_warmstart + 1:num_samples + 1])
ax1.set_ylabel("sigma")
ax1.set_title("MCMC: Global Error Scale")
ax2.scatter(mcmc_preds.mean(axis=1), y, s=10, alpha=0.5)
lo = min(mcmc_preds.mean(axis=1).min(), y.min())
hi = max(mcmc_preds.mean(axis=1).max(), y.max())
ax2.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
ax2.set_xlabel("pred")
ax2.set_ylabel("actual")
ax2.set_title("MCMC: Predicted vs Actual")
plt.tight_layout()
plt.show()
```

::::

# References
