---
title: "Custom Sampling Routine"
---

Building a custom Gibbs sampler using the low-level `stochtree` primitives —
`Dataset`, `Residual`, `ForestSampler`, `GlobalVarianceModel`, and
`LeafVarianceModel` — to implement non-standard models and sampling schemes.

::: {.callout-note}
This vignette is under construction. Content will be ported from:

- **Python**: `stochtree_repo/demo/notebooks/prototype_interface.ipynb`
- **R**: pkgdown article `CustomSamplingRoutine`
:::

## Background

The high-level `BARTModel` and `BCFModel` classes handle data preparation,
prior calibration, and sampling internally. The low-level interface exposes
these building blocks so you can:

- Customize which parameters are sampled and in what order
- Insert additional Gibbs steps (e.g. for auxiliary variables)
- Implement novel models that share `stochtree`'s tree ensemble samplers

## Setup

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
library(stochtree)
```

## Python

```{python}
#| eval: false
from stochtree import (
    Dataset,
    Residual,
    ForestSampler,
    GlobalVarianceModel,
    LeafVarianceModel,
    Forest,
    ForestContainer,
)
import numpy as np
```

::::

## Preparing Data Structures

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Construct Dataset and Residual objects
```

## Python

```{python}
#| eval: false
# Construct Dataset and Residual objects
```

::::

## Initializing Forests

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Initialize Forest and ForestContainer
```

## Python

```{python}
#| eval: false
# Initialize Forest and ForestContainer
```

::::

## Custom Gibbs Loop

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Implement a full Gibbs sampler loop
```

## Python

```{python}
#| eval: false
# Implement a full Gibbs sampler loop
```

::::

## Extracting Results

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Extract posterior samples from ForestContainer
```

## Python

```{python}
#| eval: false
# Extract posterior samples from ForestContainer
```

::::
