---
title: "Using Shared Leaf Membership as a Kernel"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(here::here(), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

A trained tree ensemble with strong out-of-sample performance admits a natural
motivation for the "distance" between two samples: shared leaf membership.
This vignette demonstrates how to extract a kernel matrix from a fitted `stochtree`
ensemble and use it for Gaussian process inference.

# Motivation

We number the leaves in an ensemble from 1 to $s$ (that is, if tree 1 has 3 leaves,
it reserves the numbers 1 - 3, and in turn if tree 2 has 5 leaves, it reserves the
numbers 4 - 8 to label its leaves, and so on). For a dataset with $n$ observations,
we construct the matrix $W$ as follows:

|    Initialize $W$ as a matrix of all zeroes with $n$ rows and as many columns as leaves in the ensemble
|    Let `s` = 0
|    **FOR** $j$ **IN** $\left\{1,\dots,m\right\}$:
|       Let `num_leaves` be the number of leaves in tree $j$
|       **FOR** $i$ **IN** $\left\{1,\dots,n\right\}$:
|          Let `k` be the leaf to which tree $j$ maps observation $i$
|          Set element $W_{i,k+s} = 1$
|       Let `s` = `s + num_leaves`

This sparse matrix $W$ is a matrix representation of the basis predictions of an
ensemble (i.e. integrating out the leaf parameters and just analyzing the leaf
indices). For an ensemble with $m$ trees, we can determine the proportion of trees
that map each observation to the same leaf by computing $W W^T / m$. This can form
the basis for a kernel function used in a Gaussian process regression, as we
demonstrate below.

# Setup

Load necessary packages

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
library(tgp)
library(MASS)
library(Matrix)
library(mvtnorm)
```

## Python

```{python}
import numpy as np
import matplotlib.pyplot as plt
from scipy.sparse import csr_matrix
from sklearn.gaussian_process import GaussianProcessRegressor
from sklearn.gaussian_process.kernels import RBF, WhiteKernel
from sklearn.datasets import make_friedman1

from stochtree import BARTModel, compute_forest_leaf_indices
```

::::

Set a seed for reproducibility

::::{.panel-tabset group="language"}

## R

```{r}
random_seed <- 101
set.seed(random_seed)
```

## Python

```{python}
random_seed = 101
rng = np.random.default_rng(random_seed)
```

::::


# Demo 1: Univariate Supervised Learning

We begin with a non-stationary simulated DGP with a single numeric covariate,
originally described in @gramacy2010categorical. We define a training set and test
set and evaluate various approaches to modeling the out-of-sample outcome.

## Traditional Gaussian Process

::::{.panel-tabset group="language"}

## R

```{r}
#| results: hide
# Generate the data
X_train <- seq(0,20,length=100)
X_test <- seq(0,20,length=99)
y_train <- (sin(pi*X_train/5) + 0.2*cos(4*pi*X_train/5)) * (X_train <= 9.6)
lin_train <- X_train>9.6;
y_train[lin_train] <- -1 + X_train[lin_train]/10
y_train <- y_train + rnorm(length(y_train), sd=0.1)
y_test <- (sin(pi*X_test/5) + 0.2*cos(4*pi*X_test/5)) * (X_test <= 9.6)
lin_test <- X_test>9.6;
y_test[lin_test] <- -1 + X_test[lin_test]/10

# Fit the GP
model_gp <- bgp(X=X_train, Z=y_train, XX=X_test)
plot(model_gp$ZZ.mean, y_test, xlab = "predicted", ylab = "actual", main = "Gaussian process")
abline(0,1,lwd=2.5,lty=3,col="red")
```

## Python

```{python}
# Generate the data
X_train_1d = np.linspace(0, 20, 100)
X_test_1d  = np.linspace(0, 20, 99)

y_train_1 = (np.sin(np.pi * X_train_1d / 5) + 0.2 * np.cos(4 * np.pi * X_train_1d / 5)) * (X_train_1d <= 9.6)
y_train_1[X_train_1d > 9.6] = -1 + X_train_1d[X_train_1d > 9.6] / 10
y_train_1 = y_train_1 + rng.normal(0, 0.1, len(X_train_1d))

y_test_1 = (np.sin(np.pi * X_test_1d / 5) + 0.2 * np.cos(4 * np.pi * X_test_1d / 5)) * (X_test_1d <= 9.6)
y_test_1[X_test_1d > 9.6] = -1 + X_test_1d[X_test_1d > 9.6] / 10

# sklearn's GaussianProcessRegressor is used here in place of R's tgp::bgp
X_train_2d = X_train_1d.reshape(-1, 1)
X_test_2d  = X_test_1d.reshape(-1, 1)
gp_kernel = RBF(length_scale=1.0) + WhiteKernel(noise_level=0.01)
model_gp_1 = GaussianProcessRegressor(kernel=gp_kernel, n_restarts_optimizer=5,
                                       random_state=random_seed)
model_gp_1.fit(X_train_2d, y_train_1)
gp_pred_1 = model_gp_1.predict(X_test_2d)

plt.scatter(gp_pred_1, y_test_1, alpha=0.5)
lo, hi = min(gp_pred_1.min(), y_test_1.min()), max(gp_pred_1.max(), y_test_1.max())
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted")
plt.ylabel("Actual")
plt.title("Gaussian process")
plt.show()
```

::::

Assess the RMSE

::::{.panel-tabset group="language"}

## R

```{r}
sqrt(mean((model_gp$ZZ.mean - y_test)^2))
```

## Python

```{python}
print(f"RMSE: {np.sqrt(np.mean((gp_pred_1 - y_test_1)**2)):.4f}")
```

::::

## BART-based Gaussian Process

::::{.panel-tabset group="language"}

## R

```{r}
# Run BART on the data
num_trees <- 200
sigma_leaf <- 1/num_trees
X_train <- as.data.frame(X_train)
X_test <- as.data.frame(X_test)
colnames(X_train) <- colnames(X_test) <- "x1"
general_params <- list(num_threads=1)
mean_forest_params <- list(num_trees=num_trees, sigma2_leaf_init=sigma_leaf)
bart_model <- bart(X_train=X_train, y_train=y_train, X_test=X_test, 
                   general_params = general_params, mean_forest_params = mean_forest_params)

# Extract kernels needed for kriging
leaf_mat_train <- computeForestLeafIndices(bart_model, X_train, forest_type = "mean",
                                           forest_inds = bart_model$model_params$num_samples - 1)
leaf_mat_test <- computeForestLeafIndices(bart_model, X_test, forest_type = "mean",
                                           forest_inds = bart_model$model_params$num_samples - 1)
W_train <- sparseMatrix(i=rep(1:length(y_train),num_trees), j=leaf_mat_train + 1, x=1)
W_test <- sparseMatrix(i=rep(1:length(y_test),num_trees), j=leaf_mat_test + 1, x=1)
Sigma_11 <- tcrossprod(W_test) / num_trees
Sigma_12 <- tcrossprod(W_test, W_train) / num_trees
Sigma_22 <- tcrossprod(W_train) / num_trees
Sigma_22_inv <- ginv(as.matrix(Sigma_22))
Sigma_21 <- t(Sigma_12)

# Compute mean and covariance for the test set posterior
mu_tilde <- Sigma_12 %*% Sigma_22_inv %*% y_train
Sigma_tilde <- as.matrix((sigma_leaf)*(Sigma_11 - Sigma_12 %*% Sigma_22_inv %*% Sigma_21))

# Sample from f(X_test) | X_test, X_train, f(X_train)
gp_samples <- mvtnorm::rmvnorm(1000, mean = mu_tilde, sigma = Sigma_tilde)

# Compute posterior mean predictions for f(X_test)
yhat_mean_test <- colMeans(gp_samples)
plot(yhat_mean_test, y_test, xlab = "predicted", ylab = "actual", main = "BART Gaussian process")
abline(0,1,lwd=2.5,lty=3,col="red")
```

## Python

```{python}
# Run BART on the data
num_trees = 200
sigma_leaf = 1 / num_trees
general_params = {"num_threads": 1}
mean_forest_params = {"num_trees": num_trees, "sigma2_leaf_init": sigma_leaf}
bart_model_1 = BARTModel()
bart_model_1.sample(X_train=X_train_2d, y_train=y_train_1, X_test=X_test_2d,
                    general_params=general_params, mean_forest_params=mean_forest_params)

# Extract leaf indices for the last retained forest sample
last_sample = bart_model_1.num_samples - 1
leaf_mat_train_1 = compute_forest_leaf_indices(bart_model_1, X_train_2d,
                                               forest_type="mean", forest_inds=last_sample)
leaf_mat_test_1  = compute_forest_leaf_indices(bart_model_1, X_test_2d,
                                               forest_type="mean", forest_inds=last_sample)

# Build sparse W matrices (rows = observations, cols = global leaf indices)
n_train_1, n_test_1 = len(y_train_1), len(y_test_1)
col_inds_train = leaf_mat_train_1.flatten()
col_inds_test  = leaf_mat_test_1.flatten()
max_col = max(col_inds_train.max(), col_inds_test.max()) + 1
W_train_1 = csr_matrix(
    (np.ones(len(col_inds_train)), (np.tile(np.arange(n_train_1), num_trees), col_inds_train)),
    shape=(n_train_1, max_col),
)
W_test_1 = csr_matrix(
    (np.ones(len(col_inds_test)), (np.tile(np.arange(n_test_1), num_trees), col_inds_test)),
    shape=(n_test_1, max_col),
)

# Compute kernel matrices
W_tr = W_train_1.toarray()
W_te = W_test_1.toarray()
Sigma_22   = (W_tr @ W_tr.T) / num_trees
Sigma_11   = (W_te @ W_te.T) / num_trees
Sigma_12   = (W_te @ W_tr.T) / num_trees
Sigma_21   = Sigma_12.T
Sigma_22_inv = np.linalg.pinv(Sigma_22)

# Compute GP posterior mean and covariance
mu_tilde    = Sigma_12 @ Sigma_22_inv @ y_train_1
Sigma_tilde = sigma_leaf * (Sigma_11 - Sigma_12 @ Sigma_22_inv @ Sigma_21)
Sigma_tilde += 1e-8 * np.eye(n_test_1)  # small jitter for numerical stability

# Sample from f(X_test) | X_test, X_train, f(X_train)
gp_samples_1 = rng.multivariate_normal(mu_tilde, Sigma_tilde, size=1000, method="eigh")

# Posterior mean predictions
yhat_mean_test_1 = gp_samples_1.mean(axis=0)
lo = min(yhat_mean_test_1.min(), y_test_1.min())
hi = max(yhat_mean_test_1.max(), y_test_1.max())
plt.scatter(yhat_mean_test_1, y_test_1, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted")
plt.ylabel("Actual")
plt.title("BART Gaussian process")
plt.show()
```

::::

Assess the RMSE

::::{.panel-tabset group="language"}

## R

```{r}
sqrt(mean((yhat_mean_test - y_test)^2))
```

## Python

```{python}
print(f"RMSE: {np.sqrt(np.mean((yhat_mean_test_1 - y_test_1)**2)):.4f}")
```

::::

# Demo 2: Multivariate Supervised Learning

We proceed to the simulated "Friedman" dataset (@friedman1991multivariate). In R, this
is accessed via `tgp::friedman.1.data`; in Python we use
`sklearn.datasets.make_friedman1`, which implements the same DGP.

## Traditional Gaussian Process

::::{.panel-tabset group="language"}

## R

```{r}
#| results: hide
# Generate the data, add many "noise variables"
n <- 100
friedman.df <- friedman.1.data(n=n)
train_inds <- sort(sample(1:n, floor(0.8*n), replace = FALSE))
test_inds <- (1:n)[!((1:n) %in% train_inds)]
X <- as.matrix(friedman.df)[,1:10]
X <- cbind(X, matrix(runif(n*10), ncol = 10))
y <- as.matrix(friedman.df)[,12] + rnorm(n,0,1)*(sd(as.matrix(friedman.df)[,11])/2)
X_train <- X[train_inds,]
X_test <- X[test_inds,]
y_train <- y[train_inds]
y_test <- y[test_inds]

# Fit the GP
model_gp <- bgp(X=X_train, Z=y_train, XX=X_test)
plot(model_gp$ZZ.mean, y_test, xlab = "predicted", ylab = "actual", main = "Gaussian process")
abline(0,1,lwd=2.5,lty=3,col="red")
```

## Python

```{python}
# Generate the data: 10 Friedman features + 10 noise features
n = 100
X_raw, y_friedman = make_friedman1(n_samples=n, n_features=10, noise=1.0,
                                    random_state=random_seed)
X_2 = np.hstack([X_raw, rng.uniform(size=(n, 10))])  # 20 features total
y_2 = y_friedman

train_inds_2 = rng.choice(n, int(0.8 * n), replace=False)
test_inds_2  = np.setdiff1d(np.arange(n), train_inds_2)
X_train_2, X_test_2 = X_2[train_inds_2], X_2[test_inds_2]
y_train_2, y_test_2 = y_2[train_inds_2], y_2[test_inds_2]

# sklearn's GaussianProcessRegressor is used here in place of R's tgp::bgp
gp_kernel_2 = RBF(length_scale=1.0) + WhiteKernel(noise_level=1.0)
model_gp_2 = GaussianProcessRegressor(kernel=gp_kernel_2, n_restarts_optimizer=1,
                                       random_state=random_seed)
model_gp_2.fit(X_train_2, y_train_2)
gp_pred_2 = model_gp_2.predict(X_test_2)

lo = min(gp_pred_2.min(), y_test_2.min())
hi = max(gp_pred_2.max(), y_test_2.max())
plt.scatter(gp_pred_2, y_test_2, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted")
plt.ylabel("Actual")
plt.title("Gaussian process")
plt.show()
```

::::

Assess the RMSE

::::{.panel-tabset group="language"}

## R

```{r}
sqrt(mean((model_gp$ZZ.mean - y_test)^2))
```

## Python

```{python}
print(f"RMSE: {np.sqrt(np.mean((gp_pred_2 - y_test_2)**2)):.4f}")
```

::::

## BART-based Gaussian Process

::::{.panel-tabset group="language"}

## R

```{r}
# Run BART on the data
num_trees <- 200
sigma_leaf <- 1/num_trees
X_train <- as.data.frame(X_train)
X_test <- as.data.frame(X_test)
general_params <- list(num_threads=1)
mean_forest_params <- list(num_trees=num_trees, sigma2_leaf_init=sigma_leaf)
bart_model <- bart(X_train=X_train, y_train=y_train, X_test=X_test, 
                   general_params = general_params, mean_forest_params = mean_forest_params)

# Extract kernels needed for kriging
leaf_mat_train <- computeForestLeafIndices(bart_model, X_train, forest_type = "mean",
                                           forest_inds = bart_model$model_params$num_samples - 1)
leaf_mat_test <- computeForestLeafIndices(bart_model, X_test, forest_type = "mean",
                                           forest_inds = bart_model$model_params$num_samples - 1)
W_train <- sparseMatrix(i=rep(1:length(y_train),num_trees), j=leaf_mat_train + 1, x=1)
W_test <- sparseMatrix(i=rep(1:length(y_test),num_trees), j=leaf_mat_test + 1, x=1)
Sigma_11 <- tcrossprod(W_test) / num_trees
Sigma_12 <- tcrossprod(W_test, W_train) / num_trees
Sigma_22 <- tcrossprod(W_train) / num_trees
Sigma_22_inv <- ginv(as.matrix(Sigma_22))
Sigma_21 <- t(Sigma_12)

# Compute mean and covariance for the test set posterior
mu_tilde <- Sigma_12 %*% Sigma_22_inv %*% y_train
Sigma_tilde <- as.matrix((sigma_leaf)*(Sigma_11 - Sigma_12 %*% Sigma_22_inv %*% Sigma_21))

# Sample from f(X_test) | X_test, X_train, f(X_train)
gp_samples <- mvtnorm::rmvnorm(1000, mean = mu_tilde, sigma = Sigma_tilde)

# Compute posterior mean predictions for f(X_test)
yhat_mean_test <- colMeans(gp_samples)
plot(yhat_mean_test, y_test, xlab = "predicted", ylab = "actual", main = "BART Gaussian process")
abline(0,1,lwd=2.5,lty=3,col="red")
```

## Python

```{python}
num_trees = 200
sigma_leaf = 1 / num_trees
general_params = {"num_threads": 1}
mean_forest_params = {"num_trees": num_trees, "sigma2_leaf_init": sigma_leaf}
bart_model_2 = BARTModel()
bart_model_2.sample(X_train=X_train_2, y_train=y_train_2, X_test=X_test_2,
                    general_params=general_params, mean_forest_params=mean_forest_params)

last_sample_2 = bart_model_2.num_samples - 1
leaf_mat_train_2 = compute_forest_leaf_indices(bart_model_2, X_train_2,
                                               forest_type="mean", forest_inds=last_sample_2)
leaf_mat_test_2  = compute_forest_leaf_indices(bart_model_2, X_test_2,
                                               forest_type="mean", forest_inds=last_sample_2)

n_train_2, n_test_2 = len(y_train_2), len(y_test_2)
col_inds_train_2 = leaf_mat_train_2.flatten()
col_inds_test_2  = leaf_mat_test_2.flatten()
max_col_2 = max(col_inds_train_2.max(), col_inds_test_2.max()) + 1
W_train_2 = csr_matrix(
    (np.ones(len(col_inds_train_2)),
     (np.tile(np.arange(n_train_2), num_trees), col_inds_train_2)),
    shape=(n_train_2, max_col_2),
)
W_test_2 = csr_matrix(
    (np.ones(len(col_inds_test_2)),
     (np.tile(np.arange(n_test_2), num_trees), col_inds_test_2)),
    shape=(n_test_2, max_col_2),
)

W_tr2 = W_train_2.toarray()
W_te2 = W_test_2.toarray()
Sigma_22_2   = (W_tr2 @ W_tr2.T) / num_trees
Sigma_11_2   = (W_te2 @ W_te2.T) / num_trees
Sigma_12_2   = (W_te2 @ W_tr2.T) / num_trees
Sigma_21_2   = Sigma_12_2.T
Sigma_22_inv_2 = np.linalg.pinv(Sigma_22_2)

mu_tilde_2    = Sigma_12_2 @ Sigma_22_inv_2 @ y_train_2
Sigma_tilde_2 = sigma_leaf * (Sigma_11_2 - Sigma_12_2 @ Sigma_22_inv_2 @ Sigma_21_2)
Sigma_tilde_2 += 1e-8 * np.eye(n_test_2)

gp_samples_2 = rng.multivariate_normal(mu_tilde_2, Sigma_tilde_2, size=1000, method="eigh")
yhat_mean_test_2 = gp_samples_2.mean(axis=0)

lo = min(yhat_mean_test_2.min(), y_test_2.min())
hi = max(yhat_mean_test_2.max(), y_test_2.max())
plt.scatter(yhat_mean_test_2, y_test_2, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted")
plt.ylabel("Actual")
plt.title("BART Gaussian process")
plt.show()
```

::::

Assess the RMSE

::::{.panel-tabset group="language"}

## R

```{r}
sqrt(mean((yhat_mean_test - y_test)^2))
```

## Python

```{python}
print(f"RMSE: {np.sqrt(np.mean((yhat_mean_test_2 - y_test_2)**2)):.4f}")
```

::::

While the use case of a BART kernel for classical kriging is perhaps unclear without
more empirical investigation, the kernel approach can be very beneficial for causal
inference applications.

# References
