---
title: "Ensemble Kernel"
---

Using the `stochtree` tree ensemble as a kernel function — extracting the
posterior similarity matrix induced by leaf co-membership across trees and
MCMC samples.

::: {.callout-note}
This vignette is under construction. Content will be ported from:

- **R**: pkgdown article `EnsembleKernel`
:::

## Background

Each tree in a BART ensemble partitions the covariate space into leaf regions.
Two observations $i$ and $j$ that fall in the same leaf of tree $t$ at sample $s$
are "similar" according to that tree. Averaging over trees and posterior samples
gives a data-driven kernel:

$$
K(x_i, x_j) = \frac{1}{Sm} \sum_{s=1}^S \sum_{t=1}^m \mathbf{1}[\ell_t^{(s)}(x_i) = \ell_t^{(s)}(x_j)]
$$

This kernel can be used for clustering, Gaussian process inference, or as a
similarity measure for downstream tasks.

## Setup

```{r}
#| eval: false
library(stochtree)
```

## Fitting a Model

```{r}
#| eval: false
# Fit a BART model
```

## Extracting the Kernel Matrix

```{r}
#| eval: false
# Compute leaf co-membership kernel
```

## Visualizing the Kernel

```{r}
#| eval: false
# Heatmap of kernel matrix
```

## Downstream Applications

```{r}
#| eval: false
# Clustering or GP prediction with the ensemble kernel
```
