---
title: "Heteroskedastic BART"
---

BART with a forest-modeled error variance, where the residual variance
$\sigma^2(X_i)$ is itself a function of covariates learned via a separate
tree ensemble.

::: {.callout-note}
This vignette is under construction. Content will be ported from:

- **Python**: `stochtree_repo/demo/notebooks/heteroskedastic_supervised_learning.ipynb`
- **R**: pkgdown article `Heteroskedasticity`
:::

## Background

Standard BART assumes homoskedastic errors $\epsilon_i \sim \mathcal{N}(0, \sigma^2)$.
When error variance varies with covariates, modeling $\sigma^2(X_i)$ jointly with
the mean function improves inference. The model is:

$$
Y_i = \mu(X_i) + \epsilon_i, \quad \epsilon_i \sim \mathcal{N}(0, \sigma^2(X_i))
$$

where both $\mu(\cdot)$ and $\log \sigma^2(\cdot)$ are modeled with BART ensembles.

## Setup

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
library(stochtree)
```

## Python

```{python}
#| eval: false
import stochtree
import numpy as np
```

::::

## Data Simulation

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Simulate data with non-constant variance
```

## Python

```{python}
#| eval: false
# Simulate data with non-constant variance
```

::::

## Model Fitting

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Fit heteroskedastic BART
```

## Python

```{python}
#| eval: false
# Fit heteroskedastic BART
```

::::

## Posterior Summaries

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Mean and variance posterior summaries
```

## Python

```{python}
#| eval: false
# Mean and variance posterior summaries
```

::::
