---
title: "BART with a Forest-based Variance Model"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(here::here(), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

This vignette demonstrates how to use the `bart()` function for Bayesian supervised
learning (@chipman2010bart) and causal inference (@hahn2020bayesian), with an
additional "variance forest," for modeling conditional variance (see @murray2021log).

# Setup

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
```

## Python

```{python}
import numpy as np
import matplotlib.pyplot as plt
from stochtree import BARTModel

rng = np.random.default_rng(101)
```

::::

# Section 1: Supervised Learning

## Demo 1: Variance-Only Simulation (Simple DGP)

Here, we generate data with a constant (zero) mean and a relatively simple
covariate-modified variance function.

\begin{equation*}
\begin{aligned}
y &= 0 + \sigma(X) \epsilon\\
\sigma^2(X) &= \begin{cases}
0.5 & X_1 \geq 0 \text{ and } X_1 < 0.25\\
1 & X_1 \geq 0.25 \text{ and } X_1 < 0.5\\
2 & X_1 \geq 0.5 \text{ and } X_1 < 0.75\\
3 & X_1 \geq 0.75 \text{ and } X_1 < 1\\
\end{cases}\\
X_1,\dots,X_p &\sim \text{U}\left(0,1\right)\\
\epsilon &\sim \mathcal{N}\left(0,1\right)
\end{aligned}
\end{equation*}

### Simulation

::::{.panel-tabset group="language"}

## R

```{r}
# Generate the data
n <- 500
p_x <- 10
X <- matrix(runif(n*p_x), ncol = p_x)
f_XW <- 0
s_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (0.5) +
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (1) +
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2) +
    ((0.75 <= X[,1]) & (1 > X[,1])) * (3)
)
y <- f_XW + rnorm(n, 0, 1)*s_XW

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
y_test <- y[test_inds]
y_train <- y[train_inds]
s_x_test <- s_XW[test_inds]
s_x_train <- s_XW[train_inds]
```

## Python

```{python}
n, p_x = 500, 10
X = rng.uniform(size=(n, p_x))
# Note: R's X[,1] = Python's X[:,0]
s_XW = (
    ((X[:, 0] >= 0)    & (X[:, 0] < 0.25)) * 0.5 +
    ((X[:, 0] >= 0.25) & (X[:, 0] < 0.5))  * 1.0 +
    ((X[:, 0] >= 0.5)  & (X[:, 0] < 0.75)) * 2.0 +
    ((X[:, 0] >= 0.75) & (X[:, 0] < 1.0))  * 3.0
)
y = rng.normal(size=n) * s_XW

test_set_pct = 0.2
n_test = round(test_set_pct * n)
test_inds  = rng.choice(n, n_test, replace=False)
train_inds = np.setdiff1d(np.arange(n), test_inds)
X_test,  X_train  = X[test_inds],  X[train_inds]
y_test,  y_train  = y[test_inds],  y[train_inds]
s_x_test, s_x_train = s_XW[test_inds], s_XW[train_inds]
```

::::

### Sampling and Analysis

#### Warmstart

We first sample the $\sigma^2(X)$ ensemble using "warm-start" initialization
(@he2023stochastic). This is the default in `stochtree`.

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
num_trees <- 20
num_samples <- num_gfr + num_burnin + num_mcmc
general_params <- list(sample_sigma2_global = F)
mean_forest_params <- list(sample_sigma2_leaf = F, num_trees = 0)
variance_forest_params <- list(num_trees = num_trees)
bart_model_warmstart <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test,
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    general_params = general_params, mean_forest_params = mean_forest_params,
    variance_forest_params = variance_forest_params
)
```

## Python

```{python}
num_trees = 20
bart_model_warmstart = BARTModel()
bart_model_warmstart.sample(
    X_train=X_train, y_train=y_train, X_test=X_test,
    num_gfr=10, num_burnin=0, num_mcmc=100,
    general_params={"num_threads": 1, "sample_sigma2_global": False},
    mean_forest_params={"sample_sigma2_leaf": False, "num_trees": 0},
    variance_forest_params={"num_trees": num_trees},
)
```

::::

Inspect the MCMC samples.

::::{.panel-tabset group="language"}

## R

```{r}
plot(rowMeans(bart_model_warmstart$sigma2_x_hat_test), s_x_test^2,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "variance function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

## Python

```{python}
var_pred = bart_model_warmstart.sigma2_x_test.mean(axis=1)
lo, hi = min(var_pred.min(), (s_x_test**2).min()), max(var_pred.max(), (s_x_test**2).max())
plt.scatter(var_pred, s_x_test**2, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Variance function")
plt.show()
```

::::

#### MCMC

We now sample the $\sigma^2(X)$ ensemble using MCMC with root initialization (as in
@chipman2010bart).

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 0
num_burnin <- 1000
num_mcmc <- 100
general_params <- list(sample_sigma2_global = F)
mean_forest_params <- list(sample_sigma2_leaf = F, num_trees = 0)
variance_forest_params <- list(num_trees = num_trees)
bart_model_mcmc <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test,
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    general_params = general_params, mean_forest_params = mean_forest_params,
    variance_forest_params = variance_forest_params
)
```

## Python

```{python}
bart_model_mcmc = BARTModel()
bart_model_mcmc.sample(
    X_train=X_train, y_train=y_train, X_test=X_test,
    num_gfr=0, num_burnin=1000, num_mcmc=100,
    general_params={"num_threads": 1, "sample_sigma2_global": False},
    mean_forest_params={"sample_sigma2_leaf": False, "num_trees": 0},
    variance_forest_params={"num_trees": num_trees},
)
```

::::

Inspect the MCMC samples.

::::{.panel-tabset group="language"}

## R

```{r}
plot(rowMeans(bart_model_mcmc$sigma2_x_hat_test), s_x_test^2,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "variance function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

## Python

```{python}
var_pred = bart_model_mcmc.sigma2_x_test.mean(axis=1)
lo, hi = min(var_pred.min(), (s_x_test**2).min()), max(var_pred.max(), (s_x_test**2).max())
plt.scatter(var_pred, s_x_test**2, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Variance function")
plt.show()
```

::::

## Demo 2: Variance-Only Simulation (Complex DGP)

Here, we generate data with a constant (zero) mean and a more complex
covariate-modified variance function.

\begin{equation*}
\begin{aligned}
y &= 0 + \sigma(X) \epsilon\\
\sigma^2(X) &= \begin{cases}
0.25X_3^2 & X_1 \geq 0 \text{ and } X_1 < 0.25\\
1X_3^2 & X_1 \geq 0.25 \text{ and } X_1 < 0.5\\
4X_3^2 & X_1 \geq 0.5 \text{ and } X_1 < 0.75\\
9X_3^2 & X_1 \geq 0.75 \text{ and } X_1 < 1\\
\end{cases}\\
X_1,\dots,X_p &\sim \text{U}\left(0,1\right)\\
\epsilon &\sim \mathcal{N}\left(0,1\right)
\end{aligned}
\end{equation*}

### Simulation

::::{.panel-tabset group="language"}

## R

```{r}
# Generate the data
n <- 500
p_x <- 10
X <- matrix(runif(n*p_x), ncol = p_x)
f_XW <- 0
s_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (0.5*X[,3]) +
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (1*X[,3]) +
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2*X[,3]) +
    ((0.75 <= X[,1]) & (1 > X[,1])) * (3*X[,3])
)
y <- f_XW + rnorm(n, 0, 1)*s_XW

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
y_test <- y[test_inds]
y_train <- y[train_inds]
s_x_test <- s_XW[test_inds]
s_x_train <- s_XW[train_inds]
```

## Python

```{python}
n, p_x = 500, 10
X = rng.uniform(size=(n, p_x))
# R's X[,3] = Python's X[:,2]
s_XW = (
    ((X[:, 0] >= 0)    & (X[:, 0] < 0.25)) * (0.5 * X[:, 2]) +
    ((X[:, 0] >= 0.25) & (X[:, 0] < 0.5))  * (1.0 * X[:, 2]) +
    ((X[:, 0] >= 0.5)  & (X[:, 0] < 0.75)) * (2.0 * X[:, 2]) +
    ((X[:, 0] >= 0.75) & (X[:, 0] < 1.0))  * (3.0 * X[:, 2])
)
y = rng.normal(size=n) * s_XW

test_set_pct = 0.2
n_test = round(test_set_pct * n)
test_inds  = rng.choice(n, n_test, replace=False)
train_inds = np.setdiff1d(np.arange(n), test_inds)
X_test,  X_train  = X[test_inds],  X[train_inds]
y_test,  y_train  = y[test_inds],  y[train_inds]
s_x_test, s_x_train = s_XW[test_inds], s_XW[train_inds]
```

::::

### Sampling and Analysis

#### Warmstart

::::{.panel-tabset group="language"}

## R

```{r}
num_trees <- 20
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
general_params <- list(sample_sigma2_global = F)
mean_forest_params <- list(sample_sigma2_leaf = F, num_trees = 0,
                           alpha = 0.95, beta = 2, min_samples_leaf = 5)
variance_forest_params <- list(num_trees = num_trees, alpha = 0.95,
                               beta = 1.25, min_samples_leaf = 1)
bart_model_warmstart <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test,
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    general_params = general_params, mean_forest_params = mean_forest_params,
    variance_forest_params = variance_forest_params
)
```

## Python

```{python}
num_trees = 20
bart_model_warmstart = BARTModel()
bart_model_warmstart.sample(
    X_train=X_train, y_train=y_train, X_test=X_test,
    num_gfr=10, num_burnin=0, num_mcmc=100,
    general_params={"num_threads": 1, "sample_sigma2_global": False},
    mean_forest_params={"sample_sigma2_leaf": False, "num_trees": 0,
                        "alpha": 0.95, "beta": 2, "min_samples_leaf": 5},
    variance_forest_params={"num_trees": num_trees, "alpha": 0.95,
                            "beta": 1.25, "min_samples_leaf": 1},
)
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
plot(rowMeans(bart_model_warmstart$sigma2_x_hat_test), s_x_test^2,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "variance function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

## Python

```{python}
var_pred = bart_model_warmstart.sigma2_x_test.mean(axis=1)
lo, hi = min(var_pred.min(), (s_x_test**2).min()), max(var_pred.max(), (s_x_test**2).max())
plt.scatter(var_pred, s_x_test**2, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Variance function")
plt.show()
```

::::

#### MCMC

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 0
num_burnin <- 1000
num_mcmc <- 100
general_params <- list(sample_sigma2_global = F)
mean_forest_params <- list(sample_sigma2_leaf = F, num_trees = 0,
                           alpha = 0.95, beta = 2, min_samples_leaf = 5)
variance_forest_params <- list(num_trees = num_trees, alpha = 0.95,
                               beta = 1.25, min_samples_leaf = 1)
bart_model_mcmc <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test,
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    general_params = general_params, mean_forest_params = mean_forest_params,
    variance_forest_params = variance_forest_params
)
```

## Python

```{python}
bart_model_mcmc = BARTModel()
bart_model_mcmc.sample(
    X_train=X_train, y_train=y_train, X_test=X_test,
    num_gfr=0, num_burnin=1000, num_mcmc=100,
    general_params={"num_threads": 1, "sample_sigma2_global": False},
    mean_forest_params={"sample_sigma2_leaf": False, "num_trees": 0,
                        "alpha": 0.95, "beta": 2, "min_samples_leaf": 5},
    variance_forest_params={"num_trees": num_trees, "alpha": 0.95,
                            "beta": 1.25, "min_samples_leaf": 1},
)
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
plot(rowMeans(bart_model_mcmc$sigma2_x_hat_test), s_x_test^2,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "variance function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

## Python

```{python}
var_pred = bart_model_mcmc.sigma2_x_test.mean(axis=1)
lo, hi = min(var_pred.min(), (s_x_test**2).min()), max(var_pred.max(), (s_x_test**2).max())
plt.scatter(var_pred, s_x_test**2, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Variance function")
plt.show()
```

::::

## Demo 3: Mean and Variance Simulation (Simple DGP)

Here, we generate data with (relatively simple) covariate-modified mean and variance
functions.

\begin{equation*}
\begin{aligned}
y &= f(X) + \sigma(X) \epsilon\\
f(X) &= \begin{cases}
-6 & X_2 \geq 0 \text{ and } X_2 < 0.25\\
-2 & X_2 \geq 0.25 \text{ and } X_2 < 0.5\\
2 & X_2 \geq 0.5 \text{ and } X_2 < 0.75\\
6 & X_2 \geq 0.75 \text{ and } X_2 < 1\\
\end{cases}\\
\sigma^2(X) &= \begin{cases}
0.25 & X_1 \geq 0 \text{ and } X_1 < 0.25\\
1 & X_1 \geq 0.25 \text{ and } X_1 < 0.5\\
4 & X_1 \geq 0.5 \text{ and } X_1 < 0.75\\
9 & X_1 \geq 0.75 \text{ and } X_1 < 1\\
\end{cases}\\
X_1,\dots,X_p &\sim \text{U}\left(0,1\right)\\
\epsilon &\sim \mathcal{N}\left(0,1\right)
\end{aligned}
\end{equation*}

### Simulation

::::{.panel-tabset group="language"}

## R

```{r}
# Generate the data
n <- 500
p_x <- 10
X <- matrix(runif(n*p_x), ncol = p_x)
f_XW <- (
    ((0 <= X[,2]) & (0.25 > X[,2])) * (-6) +
    ((0.25 <= X[,2]) & (0.5 > X[,2])) * (-2) +
    ((0.5 <= X[,2]) & (0.75 > X[,2])) * (2) +
    ((0.75 <= X[,2]) & (1 > X[,2])) * (6)
)
s_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (0.5) +
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (1) +
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2) +
    ((0.75 <= X[,1]) & (1 > X[,1])) * (3)
)
y <- f_XW + rnorm(n, 0, 1)*s_XW

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
y_test <- y[test_inds]
y_train <- y[train_inds]
f_x_test <- f_XW[test_inds]
s_x_test <- s_XW[test_inds]
```

## Python

```{python}
n, p_x = 500, 10
X = rng.uniform(size=(n, p_x))
# R's X[,2] = Python's X[:,1]; R's X[,1] = Python's X[:,0]
f_XW = (
    ((X[:, 1] >= 0)    & (X[:, 1] < 0.25)) * (-6) +
    ((X[:, 1] >= 0.25) & (X[:, 1] < 0.5))  * (-2) +
    ((X[:, 1] >= 0.5)  & (X[:, 1] < 0.75)) * (2)  +
    ((X[:, 1] >= 0.75) & (X[:, 1] < 1.0))  * (6)
)
s_XW = (
    ((X[:, 0] >= 0)    & (X[:, 0] < 0.25)) * 0.5 +
    ((X[:, 0] >= 0.25) & (X[:, 0] < 0.5))  * 1.0 +
    ((X[:, 0] >= 0.5)  & (X[:, 0] < 0.75)) * 2.0 +
    ((X[:, 0] >= 0.75) & (X[:, 0] < 1.0))  * 3.0
)
y = f_XW + rng.normal(size=n) * s_XW

test_set_pct = 0.2
n_test = round(test_set_pct * n)
test_inds  = rng.choice(n, n_test, replace=False)
train_inds = np.setdiff1d(np.arange(n), test_inds)
X_test,  X_train  = X[test_inds],  X[train_inds]
y_test,  y_train  = y[test_inds],  y[train_inds]
f_x_test = f_XW[test_inds]
s_x_test = s_XW[test_inds]
```

::::

### Sampling and Analysis

#### Warmstart

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
general_params <- list(sample_sigma2_global = F)
mean_forest_params <- list(sample_sigma2_leaf = F, num_trees = 50,
                           alpha = 0.95, beta = 2, min_samples_leaf = 5)
variance_forest_params <- list(num_trees = 50, alpha = 0.95,
                               beta = 1.25, min_samples_leaf = 5)
bart_model_warmstart <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test,
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    general_params = general_params, mean_forest_params = mean_forest_params,
    variance_forest_params = variance_forest_params
)
```

## Python

```{python}
bart_model_warmstart = BARTModel()
bart_model_warmstart.sample(
    X_train=X_train, y_train=y_train, X_test=X_test,
    num_gfr=10, num_burnin=0, num_mcmc=100,
    general_params={"num_threads": 1, "sample_sigma2_global": False},
    mean_forest_params={"sample_sigma2_leaf": False, "num_trees": 50,
                        "alpha": 0.95, "beta": 2, "min_samples_leaf": 5},
    variance_forest_params={"num_trees": 50, "alpha": 0.95,
                            "beta": 1.25, "min_samples_leaf": 5},
)
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
plot(rowMeans(bart_model_warmstart$y_hat_test), y_test,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "mean function")
abline(0,1,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_warmstart$sigma2_x_hat_test), s_x_test^2,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "variance function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

## Python

```{python}
mean_pred = bart_model_warmstart.y_hat_test.mean(axis=1)
lo, hi = min(mean_pred.min(), y_test.min()), max(mean_pred.max(), y_test.max())
plt.scatter(mean_pred, y_test, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Mean function")
plt.show()

var_pred = bart_model_warmstart.sigma2_x_test.mean(axis=1)
lo, hi = min(var_pred.min(), (s_x_test**2).min()), max(var_pred.max(), (s_x_test**2).max())
plt.scatter(var_pred, s_x_test**2, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Variance function")
plt.show()
```

::::

#### MCMC

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 0
num_burnin <- 1000
num_mcmc <- 100
general_params <- list(sample_sigma2_global = F)
mean_forest_params <- list(sample_sigma2_leaf = F, num_trees = 50,
                           alpha = 0.95, beta = 2, min_samples_leaf = 5)
variance_forest_params <- list(num_trees = 50, alpha = 0.95,
                               beta = 1.25, min_samples_leaf = 5)
bart_model_mcmc <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test,
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    general_params = general_params, mean_forest_params = mean_forest_params,
    variance_forest_params = variance_forest_params
)
```

## Python

```{python}
bart_model_mcmc = BARTModel()
bart_model_mcmc.sample(
    X_train=X_train, y_train=y_train, X_test=X_test,
    num_gfr=0, num_burnin=1000, num_mcmc=100,
    general_params={"num_threads": 1, "sample_sigma2_global": False},
    mean_forest_params={"sample_sigma2_leaf": False, "num_trees": 50,
                        "alpha": 0.95, "beta": 2, "min_samples_leaf": 5},
    variance_forest_params={"num_trees": 50, "alpha": 0.95,
                            "beta": 1.25, "min_samples_leaf": 5},
)
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
plot(rowMeans(bart_model_mcmc$y_hat_test), y_test,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "mean function")
abline(0,1,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_mcmc$sigma2_x_hat_test), s_x_test^2,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "variance function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

## Python

```{python}
mean_pred = bart_model_mcmc.y_hat_test.mean(axis=1)
lo, hi = min(mean_pred.min(), y_test.min()), max(mean_pred.max(), y_test.max())
plt.scatter(mean_pred, y_test, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Mean function")
plt.show()

var_pred = bart_model_mcmc.sigma2_x_test.mean(axis=1)
lo, hi = min(var_pred.min(), (s_x_test**2).min()), max(var_pred.max(), (s_x_test**2).max())
plt.scatter(var_pred, s_x_test**2, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Variance function")
plt.show()
```

::::

## Demo 4: Mean and Variance Simulation (Complex DGP)

Here, we generate data with more complex covariate-modified mean and variance
functions.

\begin{equation*}
\begin{aligned}
y &= f(X) + \sigma(X) \epsilon\\
f(X) &= \begin{cases}
-6X_4 & X_2 \geq 0 \text{ and } X_2 < 0.25\\
-2X_4 & X_2 \geq 0.25 \text{ and } X_2 < 0.5\\
2X_4 & X_2 \geq 0.5 \text{ and } X_2 < 0.75\\
6X_4 & X_2 \geq 0.75 \text{ and } X_2 < 1\\
\end{cases}\\
\sigma^2(X) &= \begin{cases}
0.25X_3^2 & X_1 \geq 0 \text{ and } X_1 < 0.25\\
1X_3^2 & X_1 \geq 0.25 \text{ and } X_1 < 0.5\\
4X_3^2 & X_1 \geq 0.5 \text{ and } X_1 < 0.75\\
9X_3^2 & X_1 \geq 0.75 \text{ and } X_1 < 1\\
\end{cases}\\
X_1,\dots,X_p &\sim \text{U}\left(0,1\right)\\
\epsilon &\sim \mathcal{N}\left(0,1\right)
\end{aligned}
\end{equation*}

### Simulation

::::{.panel-tabset group="language"}

## R

```{r}
# Generate the data
n <- 500
p_x <- 10
X <- matrix(runif(n*p_x), ncol = p_x)
f_XW <- (
    ((0 <= X[,2]) & (0.25 > X[,2])) * (-6*X[,4]) +
    ((0.25 <= X[,2]) & (0.5 > X[,2])) * (-2*X[,4]) +
    ((0.5 <= X[,2]) & (0.75 > X[,2])) * (2*X[,4]) +
    ((0.75 <= X[,2]) & (1 > X[,2])) * (6*X[,4])
)
s_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (0.5*X[,3]) +
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (1*X[,3]) +
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2*X[,3]) +
    ((0.75 <= X[,1]) & (1 > X[,1])) * (3*X[,3])
)
y <- f_XW + rnorm(n, 0, 1)*s_XW

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
y_test <- y[test_inds]
y_train <- y[train_inds]
f_x_test <- f_XW[test_inds]
s_x_test <- s_XW[test_inds]
```

## Python

```{python}
n, p_x = 500, 10
X = rng.uniform(size=(n, p_x))
# R's X[,2]=X[:,1], X[,4]=X[:,3], X[,1]=X[:,0], X[,3]=X[:,2]
f_XW = (
    ((X[:, 1] >= 0)    & (X[:, 1] < 0.25)) * (-6 * X[:, 3]) +
    ((X[:, 1] >= 0.25) & (X[:, 1] < 0.5))  * (-2 * X[:, 3]) +
    ((X[:, 1] >= 0.5)  & (X[:, 1] < 0.75)) * (2  * X[:, 3]) +
    ((X[:, 1] >= 0.75) & (X[:, 1] < 1.0))  * (6  * X[:, 3])
)
s_XW = (
    ((X[:, 0] >= 0)    & (X[:, 0] < 0.25)) * (0.5 * X[:, 2]) +
    ((X[:, 0] >= 0.25) & (X[:, 0] < 0.5))  * (1.0 * X[:, 2]) +
    ((X[:, 0] >= 0.5)  & (X[:, 0] < 0.75)) * (2.0 * X[:, 2]) +
    ((X[:, 0] >= 0.75) & (X[:, 0] < 1.0))  * (3.0 * X[:, 2])
)
y = f_XW + rng.normal(size=n) * s_XW

test_set_pct = 0.2
n_test = round(test_set_pct * n)
test_inds  = rng.choice(n, n_test, replace=False)
train_inds = np.setdiff1d(np.arange(n), test_inds)
X_test,  X_train  = X[test_inds],  X[train_inds]
y_test,  y_train  = y[test_inds],  y[train_inds]
f_x_test = f_XW[test_inds]
s_x_test = s_XW[test_inds]
```

::::

### Sampling and Analysis

#### Warmstart

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
general_params <- list(sample_sigma2_global = F)
mean_forest_params <- list(sample_sigma2_leaf = F, num_trees = 50,
                           alpha = 0.95, beta = 2, min_samples_leaf = 5)
variance_forest_params <- list(num_trees = 50, alpha = 0.95,
                               beta = 1.25, min_samples_leaf = 5)
bart_model_warmstart <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test,
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    general_params = general_params, mean_forest_params = mean_forest_params,
    variance_forest_params = variance_forest_params
)
```

## Python

```{python}
bart_model_warmstart = BARTModel()
bart_model_warmstart.sample(
    X_train=X_train, y_train=y_train, X_test=X_test,
    num_gfr=10, num_burnin=0, num_mcmc=100,
    general_params={"num_threads": 1, "sample_sigma2_global": False},
    mean_forest_params={"sample_sigma2_leaf": False, "num_trees": 50,
                        "alpha": 0.95, "beta": 2, "min_samples_leaf": 5},
    variance_forest_params={"num_trees": 50, "alpha": 0.95,
                            "beta": 1.25, "min_samples_leaf": 5},
)
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
plot(rowMeans(bart_model_warmstart$y_hat_test), y_test,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "mean function")
abline(0,1,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_warmstart$sigma2_x_hat_test), s_x_test^2,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "variance function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

## Python

```{python}
mean_pred = bart_model_warmstart.y_hat_test.mean(axis=1)
lo, hi = min(mean_pred.min(), y_test.min()), max(mean_pred.max(), y_test.max())
plt.scatter(mean_pred, y_test, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Mean function")
plt.show()

var_pred = bart_model_warmstart.sigma2_x_test.mean(axis=1)
lo, hi = min(var_pred.min(), (s_x_test**2).min()), max(var_pred.max(), (s_x_test**2).max())
plt.scatter(var_pred, s_x_test**2, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Variance function")
plt.show()
```

::::

#### MCMC

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 0
num_burnin <- 1000
num_mcmc <- 100
general_params <- list(sample_sigma2_global = F)
mean_forest_params <- list(sample_sigma2_leaf = F, num_trees = 50,
                           alpha = 0.95, beta = 2, min_samples_leaf = 5)
variance_forest_params <- list(num_trees = 50, alpha = 0.95,
                               beta = 1.25, min_samples_leaf = 5)
bart_model_mcmc <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test,
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    general_params = general_params, mean_forest_params = mean_forest_params,
    variance_forest_params = variance_forest_params
)
```

## Python

```{python}
bart_model_mcmc = BARTModel()
bart_model_mcmc.sample(
    X_train=X_train, y_train=y_train, X_test=X_test,
    num_gfr=0, num_burnin=1000, num_mcmc=100,
    general_params={"num_threads": 1, "sample_sigma2_global": False},
    mean_forest_params={"sample_sigma2_leaf": False, "num_trees": 50,
                        "alpha": 0.95, "beta": 2, "min_samples_leaf": 5},
    variance_forest_params={"num_trees": 50, "alpha": 0.95,
                            "beta": 1.25, "min_samples_leaf": 5},
)
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
plot(rowMeans(bart_model_mcmc$y_hat_test), y_test,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "mean function")
abline(0,1,col="red",lty=2,lwd=2.5)
plot(rowMeans(bart_model_mcmc$sigma2_x_hat_test), s_x_test^2,
     pch=16, cex=0.75, xlab = "pred", ylab = "actual", main = "variance function")
abline(0,1,col="red",lty=2,lwd=2.5)
```

## Python

```{python}
mean_pred = bart_model_mcmc.y_hat_test.mean(axis=1)
lo, hi = min(mean_pred.min(), y_test.min()), max(mean_pred.max(), y_test.max())
plt.scatter(mean_pred, y_test, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Mean function")
plt.show()

var_pred = bart_model_mcmc.sigma2_x_test.mean(axis=1)
lo, hi = min(var_pred.min(), (s_x_test**2).min()), max(var_pred.max(), (s_x_test**2).max())
plt.scatter(var_pred, s_x_test**2, s=10, alpha=0.6)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual"); plt.title("Variance function")
plt.show()
```

::::

# References
