---
title: "Multi-Chain Inference"
bibliography: vignettes.bib
---

# Motivation

Mixing of an MCMC sampler is a perennial concern for complex Bayesian models. BART
and BCF are no exception. One common way to address such concerns is to run multiple
independent "chains" of an MCMC sampler, so that if each chain gets stuck in a
different region of the posterior, their combined samples attain better coverage of
the full posterior.

This idea works with the classic "root-initialized" MCMC sampler of @chipman2010bart,
but a key insight of @he2023stochastic and @krantsevich2023stochastic is that the GFR
algorithm may be used to warm-start initialize multiple chains of the BART / BCF MCMC
sampler.

Operationally, the above two approaches have the same implementation (setting
`num_gfr > 0` if warm-start initialization is desired), so this vignette will
demonstrate how to run a multi-chain sampler sequentially or in parallel.

# Setup

::::{.panel-tabset group="language"}

## R

```{r}
#| warning: false
#| message: false
library(stochtree)
library(ggplot2)
library(coda)
library(bayesplot)
library(foreach)
library(doParallel)
```

## Python

```{python}
# Python implementation coming soon
```

::::

# Demo 1: Supervised Learning

## Data Simulation

Simulate a simple partitioned linear model.

::::{.panel-tabset group="language"}

## R

```{r}
# Generate the data
set.seed(1111)
n <- 500
p_x <- 10
p_w <- 1
snr <- 3
X <- matrix(runif(n * p_x), ncol = p_x)
leaf_basis <- matrix(runif(n * p_w), ncol = p_w)
f_XW <- (((0 <= X[, 1]) & (0.25 > X[, 1])) *
  (-7.5 * leaf_basis[, 1]) +
  ((0.25 <= X[, 1]) & (0.5 > X[, 1])) * (-2.5 * leaf_basis[, 1]) +
  ((0.5 <= X[, 1]) & (0.75 > X[, 1])) * (2.5 * leaf_basis[, 1]) +
  ((0.75 <= X[, 1]) & (1 > X[, 1])) * (7.5 * leaf_basis[, 1]))
noise_sd <- sd(f_XW) / snr
y <- f_XW + rnorm(n, 0, 1) * noise_sd

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct * n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- X[test_inds, ]
X_train <- X[train_inds, ]
leaf_basis_test <- leaf_basis[test_inds, ]
leaf_basis_train <- leaf_basis[train_inds, ]
y_test <- y[test_inds]
y_train <- y[train_inds]
```

## Python

```{python}
# Python implementation coming soon
```

::::

## Sampling Multiple Chains Sequentially from Scratch

The simplest way to sample multiple chains of a stochtree model is to do so
"sequentially," that is, after chain 1 is sampled, chain 2 is sampled from a
different starting state, and similarly for each of the requested chains. This is
supported internally in both the `bart()` and `bcf()` functions, with the
`num_chains` parameter in the `general_params` list.

Define some high-level parameters, including number of chains to run and number of
samples per chain. Here we run 4 independent chains with 2000 MCMC iterations, each
of which is burned in for 1000 iterations.

::::{.panel-tabset group="language"}

## R

```{r}
num_chains <- 4
num_gfr <- 0
num_burnin <- 1000
num_mcmc <- 2000
```

## Python

```{python}
# Python implementation coming soon
```

::::

Run the sampler.

::::{.panel-tabset group="language"}

## R

```{r}
bart_model <- stochtree::bart(
  X_train = X_train,
  leaf_basis_train = leaf_basis_train,
  y_train = y_train,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = list(num_chains = num_chains)
)
```

## Python

```{python}
# Python implementation coming soon
```

::::

Now we have a `bartmodel` object with `num_chains * num_mcmc` samples stored
internally. These samples are arranged sequentially, with the first `num_mcmc`
samples corresponding to chain 1, the next `num_mcmc` samples to chain 2, etc.

Since each chain is a set of samples of the same model, we can analyze the samples
collectively, for example, by looking at out-of-sample predictions.

::::{.panel-tabset group="language"}

## R

```{r}
y_hat_test <- predict(
  bart_model,
  X = X_test,
  leaf_basis = leaf_basis_test,
  type = "mean",
  terms = "y_hat"
)
plot(y_hat_test, y_test, xlab = "Predicted", ylab = "Actual")
abline(0, 1, col = "red", lty = 3, lwd = 3)
```

## Python

```{python}
# Python implementation coming soon
```

::::

Now, suppose we want to analyze each of the chains separately to assess mixing /
convergence. We can construct an `mcmc.list` in the `coda` package to perform various
diagnostics.

::::{.panel-tabset group="language"}

## R

```{r}
sigma2_coda_list <- coda::as.mcmc.list(lapply(
  1:num_chains,
  function(chain_idx) {
    offset <- (chain_idx - 1) * num_mcmc
    inds_start <- offset + 1
    inds_end <- offset + num_mcmc
    coda::mcmc(bart_model$sigma2_global_samples[inds_start:inds_end])
  }
))
traceplot(sigma2_coda_list, ylab = expression(sigma^2))
abline(h = noise_sd^2, col = "black", lty = 3, lwd = 3)
acf <- autocorr.diag(sigma2_coda_list)
ess <- effectiveSize(sigma2_coda_list)
rhat <- gelman.diag(sigma2_coda_list, autoburnin = F)
cat(paste0(
  "Average autocorrelation across chains:\n",
  paste0(paste0(rownames(acf), ": ", round(acf, 3)), collapse = ", "),
  "\nTotal effective sample size across chains: ",
  paste0(round(ess, 1), collapse = ", "),
  "\n'R-hat' potential scale reduction factor of Gelman and Rubin (1992)): ",
  paste0(round(rhat$psrf[, 1], 3), collapse = ", ")
))
```

## Python

```{python}
# Python implementation coming soon
```

::::

We can convert this to an array to be consumed by the `bayesplot` package.

::::{.panel-tabset group="language"}

## R

```{r}
coda_array <- as.array(sigma2_coda_list)
dim(coda_array) <- c(nrow(coda_array), ncol(coda_array), 1)
dimnames(coda_array) <- list(
  Iteration = paste0("iter", 1:num_mcmc),
  Chain = paste0("chain", 1:num_chains),
  Parameter = "sigma2_global"
)
```

## Python

```{python}
# Python implementation coming soon
```

::::

From here, we can visualize the posterior of $\sigma^2$ for each chain, comparing
to the true simulated value.

::::{.panel-tabset group="language"}

## R

```{r}
#| warning: false
#| message: false
bayesplot::mcmc_hist_by_chain(
  coda_array,
  pars = "sigma2_global"
) +
  ggplot2::labs(
    title = "Global error scale posterior by chain",
    x = expression(sigma^2)
  ) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(hjust = 0.5)
  ) +
  ggplot2::geom_vline(
    xintercept = noise_sd^2,
    color = "black",
    linetype = "dashed",
    size = 1
  )
```

## Python

```{python}
# Python implementation coming soon
```

::::

## Sampling Multiple Chains Sequentially from XBART Forests

In the example above, each chain was initialized from "root". If we sample a model
using a small number of 'grow-from-root' iterations, we can use these forests to
initialize MCMC chains.

::::{.panel-tabset group="language"}

## R

```{r}
num_chains <- 4
num_gfr <- 5
num_burnin <- 1000
num_mcmc <- 2000
```

## Python

```{python}
# Python implementation coming soon
```

::::

Run the initial GFR sampler.

::::{.panel-tabset group="language"}

## R

```{r}
xbart_model <- stochtree::bart(
  X_train = X_train,
  leaf_basis_train = leaf_basis_train,
  y_train = y_train,
  num_gfr = num_gfr,
  num_burnin = 0,
  num_mcmc = 0
)
xbart_model_string <- stochtree::saveBARTModelToJsonString(xbart_model)
```

## Python

```{python}
# Python implementation coming soon
```

::::

Run the multi-chain BART sampler, with each chain initialized from a different GFR
forest.

::::{.panel-tabset group="language"}

## R

```{r}
bart_model <- stochtree::bart(
  X_train = X_train,
  leaf_basis_train = leaf_basis_train,
  y_train = y_train,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = list(num_chains = num_chains),
  previous_model_json = xbart_model_string,
  previous_model_warmstart_sample_num = num_gfr
)
```

## Python

```{python}
# Python implementation coming soon
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
y_hat_test <- predict(
  bart_model,
  X = X_test,
  leaf_basis = leaf_basis_test,
  type = "mean",
  terms = "y_hat"
)
plot(y_hat_test, y_test, xlab = "Predicted", ylab = "Actual")
abline(0, 1, col = "red", lty = 3, lwd = 3)
```

## Python

```{python}
# Python implementation coming soon
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
sigma2_coda_list <- coda::as.mcmc.list(lapply(
  1:num_chains,
  function(chain_idx) {
    offset <- (chain_idx - 1) * num_mcmc
    inds_start <- offset + 1
    inds_end <- offset + num_mcmc
    coda::mcmc(bart_model$sigma2_global_samples[inds_start:inds_end])
  }
))
traceplot(sigma2_coda_list, ylab = expression(sigma^2))
abline(h = noise_sd^2, col = "black", lty = 3, lwd = 3)
acf <- autocorr.diag(sigma2_coda_list)
ess <- effectiveSize(sigma2_coda_list)
rhat <- gelman.diag(sigma2_coda_list, autoburnin = F)
cat(paste0(
  "Average autocorrelation across chains:\n",
  paste0(paste0(rownames(acf), ": ", round(acf, 3)), collapse = ", "),
  "\nTotal effective sample size across chains: ",
  paste0(round(ess, 1), collapse = ", "),
  "\n'R-hat' potential scale reduction factor of Gelman and Rubin (1992)): ",
  paste0(round(rhat$psrf[, 1], 3), collapse = ", ")
))
```

## Python

```{python}
# Python implementation coming soon
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
coda_array <- as.array(sigma2_coda_list)
dim(coda_array) <- c(nrow(coda_array), ncol(coda_array), 1)
dimnames(coda_array) <- list(
  Iteration = paste0("iter", 1:num_mcmc),
  Chain = paste0("chain", 1:num_chains),
  Parameter = "sigma2_global"
)
```

## Python

```{python}
# Python implementation coming soon
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
#| warning: false
#| message: false
bayesplot::mcmc_hist_by_chain(
  coda_array,
  pars = "sigma2_global"
) +
  ggplot2::labs(
    title = "Global error scale posterior by chain",
    x = expression(sigma^2)
  ) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(hjust = 0.5)
  ) +
  ggplot2::geom_vline(
    xintercept = noise_sd^2,
    color = "black",
    linetype = "dashed",
    size = 1
  )
```

## Python

```{python}
# Python implementation coming soon
```

::::

## Sampling Multiple Chains in Parallel

While the above examples used sequential multi-chain sampling internally in `bart()`
and `bcf()`, it is also possible to run chains in parallel via `doParallel`. While
`bartmodel` or `bcfmodel` objects contain external pointers to C++ data structures
which are not reachable by other processes, we can serialize `stochtree` models to
JSON for cross-process communication. After `num_chains` models have been run in
parallel and their JSON representations have been collated in the primary R session,
we can combine these into a single `bartmodel` or `bcfmodel` object via
`createBARTModelFromCombinedJsonString()` or `createBCFModelFromCombinedJsonString()`.

In order to run multiple parallel stochtree chains, a parallel backend must be
registered in your R environment. Note that we do not evaluate the cluster setup
code below in order to interact nicely with CRAN / GitHub Actions environments.

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
ncores <- parallel::detectCores()
cl <- makeCluster(ncores)
registerDoParallel(cl)
```

## Python

```{python}
# Python implementation coming soon
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
num_chains <- 4
num_gfr <- 0
num_burnin <- 100
num_mcmc <- 100
```

## Python

```{python}
# Python implementation coming soon
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
bart_model_outputs <- foreach(i = 1:num_chains) %dopar%
  {
    random_seed <- i
    general_params <- list(sample_sigma2_global = T, random_seed = random_seed)
    mean_forest_params <- list(sample_sigma2_leaf = F)
    bart_model <- stochtree::bart(
      X_train = X_train,
      leaf_basis_train = leaf_basis_train,
      y_train = y_train,
      X_test = X_test,
      leaf_basis_test = leaf_basis_test,
      num_gfr = num_gfr,
      num_burnin = num_burnin,
      num_mcmc = num_mcmc,
      general_params = general_params,
      mean_forest_params = mean_forest_params
    )
    bart_model_string <- stochtree::saveBARTModelToJsonString(bart_model)
    y_hat_test <- bart_model$y_hat_test
    list(model = bart_model_string, yhat = y_hat_test)
  }
```

## Python

```{python}
# Python implementation coming soon
```

::::

Close the parallel cluster (not evaluated here).

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
stopCluster(cl)
```

## Python

```{python}
# Python implementation coming soon
```

::::

Combine the forests from each BART model into a single forest.

::::{.panel-tabset group="language"}

## R

```{r}
bart_model_strings <- list()
bart_model_yhats <- matrix(NA, nrow = length(y_test), ncol = num_chains)
for (i in 1:length(bart_model_outputs)) {
  bart_model_strings[[i]] <- bart_model_outputs[[i]]$model
  bart_model_yhats[, i] <- rowMeans(bart_model_outputs[[i]]$yhat)
}
combined_bart <- createBARTModelFromCombinedJsonString(bart_model_strings)
```

## Python

```{python}
# Python implementation coming soon
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
yhat_combined <- predict(combined_bart, X_test, leaf_basis_test)$y_hat
```

## Python

```{python}
# Python implementation coming soon
```

::::

Compare average predictions from each chain to the original predictions and to
the true $y$ values.

::::{.panel-tabset group="language"}

## R

```{r}
par(mfrow = c(1, 2))
for (i in 1:num_chains) {
  offset <- (i - 1) * num_mcmc
  inds_start <- offset + 1
  inds_end <- offset + num_mcmc
  plot(
    rowMeans(yhat_combined[, inds_start:inds_end]),
    bart_model_yhats[, i],
    xlab = "deserialized",
    ylab = "original",
    main = paste0("Chain ", i, "\nPredictions")
  )
  abline(0, 1, col = "red", lty = 3, lwd = 3)
}
par(mfrow = c(1, 1))
```

## Python

```{python}
# Python implementation coming soon
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
par(mfrow = c(1, 2))
for (i in 1:num_chains) {
  offset <- (i - 1) * num_mcmc
  inds_start <- offset + 1
  inds_end <- offset + num_mcmc
  plot(
    rowMeans(yhat_combined[, inds_start:inds_end]),
    y_test,
    xlab = "predicted",
    ylab = "actual",
    main = paste0("Chain ", i, "\nPredictions")
  )
  abline(0, 1, col = "red", lty = 3, lwd = 3)
}
par(mfrow = c(1, 1))
```

## Python

```{python}
# Python implementation coming soon
```

::::

# References
