---
title: "Multivariate Treatment BCF"
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(here::here(), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

BCF extended to vector-valued (multivariate) treatments, estimating heterogeneous
effects for multiple treatment arms simultaneously.

## Background

When treatments are multivariate — such as continuous dose vectors or multiple
binary arms — the standard BCF model extends to:

$$
Y_i = \mu(X_i) + \tau(X_i)^\top Z_i + \epsilon_i
$$

where $Z_i \in \mathbb{R}^p$ and $\tau(X_i) \in \mathbb{R}^p$ is a vector of
covariate-varying treatment effects.

## Setup

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
```

## Python

```{python}
import matplotlib.pyplot as plt
import numpy as np
import pandas as pd
import seaborn as sns
from sklearn.model_selection import train_test_split

from stochtree import BCFModel
```

::::

## Data Simulation

::::{.panel-tabset group="language"}

## R

```{r}
# Generate covariates and propensity scores
n <- 500
p_X <- 5
X <- matrix(runif(n * p_X), nrow = n, ncol = p_X)
pi_X <- cbind(0.25 + 0.5 * X[, 1], 0.75 - 0.5 * X[, 2])
Z <- cbind(rbinom(n, 1, pi_X[, 1]), rbinom(n, 1, pi_X[, 2]))

# Define outcome mean functions (prognostic and treatment effects)
mu_X <- pi_X[, 1] * 5 + pi_X[, 2] * 2 + 2 * X[, 3]
tau_X <- cbind(X[, 2], X[, 3])

# Generate outcome
treatment_term <- rowSums(tau_X * Z)
y <- mu_X + treatment_term + rnorm(n)
```

## Python

```{python}
# RNG
rng = np.random.default_rng()

# Generate covariates and basis
n = 500
p_X = 5
X = rng.uniform(0, 1, (n, p_X))
pi_X = np.c_[0.25 + 0.5 * X[:, 0], 0.75 - 0.5 * X[:, 1]]
Z = rng.binomial(1, pi_X, (n, 2))

# Define the outcome mean functions (prognostic and treatment effects)
mu_X = pi_X[:, 0] * 5 + pi_X[:, 1] * 2 + 2 * X[:, 2]
tau_X = np.stack((X[:, 1], X[:, 2]), axis=-1)

# Generate outcome
epsilon = rng.normal(0, 1, n)
treatment_term = np.multiply(tau_X, Z).sum(axis=1)
y = mu_X + treatment_term + epsilon
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
n_test <- round(n * 0.5)
test_inds <- sort(sample(seq_len(n), n_test, replace = FALSE))
train_inds <- setdiff(seq_len(n), test_inds)
X_train <- X[train_inds, ]
X_test  <- X[test_inds, ]
Z_train <- Z[train_inds, ]
Z_test  <- Z[test_inds, ]
y_train <- y[train_inds]
y_test  <- y[test_inds]
mu_train  <- mu_X[train_inds]
mu_test   <- mu_X[test_inds]
tau_train <- tau_X[train_inds, ]
tau_test  <- tau_X[test_inds, ]
```

## Python

```{python}
sample_inds = np.arange(n)
train_inds, test_inds = train_test_split(sample_inds, test_size=0.5)
X_train = X[train_inds, :]
X_test = X[test_inds, :]
Z_train = Z[train_inds, :]
Z_test = Z[test_inds, :]
y_train = y[train_inds]
y_test = y[test_inds]
pi_train = pi_X[train_inds]
pi_test = pi_X[test_inds]
mu_train = mu_X[train_inds]
mu_test = mu_X[test_inds]
tau_train = tau_X[train_inds, :]
tau_test = tau_X[test_inds, :]
```

::::

## Model Fitting

::::{.panel-tabset group="language"}

## R

```{r}
# Note: propensity adjustment is not supported for multivariate treatment in bcf()
bcf_model <- bcf(
  X_train = X_train,
  Z_train = Z_train,
  y_train = y_train,
  X_test = X_test,
  Z_test = Z_test,
  num_gfr = 10,
  num_mcmc = 100
)
```

## Python

```{python}
bcf_model = BCFModel()
bcf_model.sample(
    X_train=X_train,
    Z_train=Z_train,
    y_train=y_train,
    propensity_train=pi_train,
    X_test=X_test,
    Z_test=Z_test,
    propensity_test=pi_test,
    num_gfr=10,
    num_mcmc=100,
)
```

::::

## Posterior Summaries

### Outcome

::::{.panel-tabset group="language"}

## R

```{r}
# y_hat_test is (n_test x num_samples)
plot(
  rowMeans(bcf_model$y_hat_test), y_test,
  xlab = "Average estimated outcome", ylab = "True outcome"
)
abline(0, 1, col = "black", lty = 3)
```

```{r}
sqrt(mean((rowMeans(bcf_model$y_hat_test) - y_test)^2))
```

## Python

```{python}
forest_preds_y_mcmc = bcf_model.y_hat_test
y_avg_mcmc = np.squeeze(forest_preds_y_mcmc).mean(axis=1, keepdims=True)
y_df_mcmc = pd.DataFrame(
    np.concatenate((np.expand_dims(y_test, 1), y_avg_mcmc), axis=1),
    columns=["True outcome", "Average estimated outcome"],
)
sns.scatterplot(data=y_df_mcmc, x="Average estimated outcome", y="True outcome")
plt.axline((0, 0), slope=1, color="black", linestyle=(0, (3, 3)))
plt.show()
```

```{python}
np.sqrt(np.mean(np.power(y_avg_mcmc - y_test, 2)))
```

::::

### Treatment Effects

::::{.panel-tabset group="language"}

## R

```{r}
# tau_hat_test is (n_test x p_Z x num_samples) for multivariate treatment
tau_avg_1 <- rowMeans(bcf_model$tau_hat_test[, 1, ])
plot(
  tau_test[, 1], tau_avg_1,
  xlab = "True tau", ylab = "Average estimated tau",
  main = "Treatment 1"
)
abline(0, 1, col = "black", lty = 3)
```

```{r}
tau_avg_2 <- rowMeans(bcf_model$tau_hat_test[, 2, ])
plot(
  tau_test[, 2], tau_avg_2,
  xlab = "True tau", ylab = "Average estimated tau",
  main = "Treatment 2"
)
abline(0, 1, col = "black", lty = 3)
```

## Python

```{python}
treatment_idx = 0
forest_preds_tau_mcmc = np.squeeze(bcf_model.tau_hat_test[:, :, treatment_idx])
tau_avg_mcmc = np.squeeze(forest_preds_tau_mcmc).mean(axis=1, keepdims=True)
tau_df_mcmc = pd.DataFrame(
    np.concatenate(
        (np.expand_dims(tau_test[:, treatment_idx], 1), tau_avg_mcmc), axis=1
    ),
    columns=["True tau", "Average estimated tau"],
)
sns.scatterplot(data=tau_df_mcmc, x="True tau", y="Average estimated tau")
plt.axline((0, 0), slope=1, color="black", linestyle=(0, (3, 3)))
plt.show()
```

```{python}
treatment_idx = 1
forest_preds_tau_mcmc = np.squeeze(bcf_model.tau_hat_test[:, :, treatment_idx])
tau_avg_mcmc = np.squeeze(forest_preds_tau_mcmc).mean(axis=1, keepdims=True)
tau_df_mcmc = pd.DataFrame(
    np.concatenate(
        (np.expand_dims(tau_test[:, treatment_idx], 1), tau_avg_mcmc), axis=1
    ),
    columns=["True tau", "Average estimated tau"],
)
sns.scatterplot(data=tau_df_mcmc, x="True tau", y="Average estimated tau")
plt.axline((0, 0), slope=1, color="black", linestyle=(0, (3, 3)))
plt.show()
```

::::

### Treatment Term

::::{.panel-tabset group="language"}

## R

```{r}
# Compute sum_j(tau_hat(X)_j * Z_j) per observation per sample
treatment_term_mcmc <- apply(bcf_model$tau_hat_test, 3, function(tau_s) {
  rowSums(tau_s * Z_test)
})
true_treatment_term <- rowSums(tau_test * Z_test)
plot(
  true_treatment_term, rowMeans(treatment_term_mcmc),
  xlab = "True treatment term", ylab = "Average estimated treatment term"
)
abline(0, 1, col = "black", lty = 3)
```

## Python

```{python}
treatment_term_mcmc_test = np.multiply(
    np.atleast_3d(Z_test).swapaxes(1, 2), bcf_model.tau_hat_test
).sum(axis=2)
treatment_term_test = np.multiply(tau_test, Z_test).sum(axis=1)
treatment_term_mcmc_avg = np.squeeze(treatment_term_mcmc_test).mean(
    axis=1, keepdims=True
)
mu_df_mcmc = pd.DataFrame(
    np.concatenate(
        (np.expand_dims(treatment_term_test, 1), treatment_term_mcmc_avg), axis=1
    ),
    columns=["True treatment term", "Average estimated treatment term"],
)
sns.scatterplot(
    data=mu_df_mcmc, x="True treatment term", y="Average estimated treatment term"
)
plt.axline((0, 0), slope=1, color="black", linestyle=(0, (3, 3)))
plt.show()
```

::::

### Prognostic Function

::::{.panel-tabset group="language"}

## R

```{r}
plot(
  mu_test, rowMeans(bcf_model$mu_hat_test),
  xlab = "True mu", ylab = "Average estimated mu"
)
abline(0, 1, col = "black", lty = 3)
```

## Python

```{python}
forest_preds_mu_mcmc = bcf_model.mu_hat_test
mu_avg_mcmc = np.squeeze(forest_preds_mu_mcmc).mean(axis=1, keepdims=True)
mu_df_mcmc = pd.DataFrame(
    np.concatenate((np.expand_dims(mu_test, 1), mu_avg_mcmc), axis=1),
    columns=["True mu", "Average estimated mu"],
)
sns.scatterplot(data=mu_df_mcmc, x="True mu", y="Average estimated mu")
plt.axline((0, 0), slope=1, color="black", linestyle=(0, (3, 3)))
plt.show()
```

::::

### Global Error Variance

::::{.panel-tabset group="language"}

## R

```{r}
plot(
  bcf_model$sigma2_global_samples,
  xlab = "Sample", ylab = expression(sigma^2)
)
```

## Python

```{python}
sigma_df_mcmc = pd.DataFrame(
    np.concatenate(
        (
            np.expand_dims(
                np.arange(bcf_model.num_samples - bcf_model.num_gfr), axis=1
            ),
            np.expand_dims(bcf_model.global_var_samples[bcf_model.num_gfr:], axis=1),
        ),
        axis=1,
    ),
    columns=["Sample", "Sigma"],
)
sns.scatterplot(data=sigma_df_mcmc, x="Sample", y="Sigma")
plt.show()
```

::::
