---
title: "Multi-Chain Inference"
---

Running multiple independent MCMC chains with `stochtree` and combining
their results for improved mixing diagnostics and more robust posterior summaries.

::: {.callout-note}
This vignette is under construction. Content will be ported from:

- **Python**: `stochtree_repo/demo/notebooks/multi_chain.ipynb`
- **R**: pkgdown article `MultiChain`
:::

## Background

Running a single MCMC chain can be fragile. Multiple independent chains allow:

- **Convergence diagnostics**: $\hat{R}$ (potential scale reduction factor) detects
  non-convergence across chains.
- **Parallelism**: Chains can run concurrently on multiple cores.
- **Ensemble pooling**: Posterior samples from all chains can be pooled for final inference.

## Setup

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
library(stochtree)
library(doParallel)
```

## Python

```{python}
#| eval: false
import stochtree
import numpy as np
from multiprocessing import Pool
```

::::

## Data Simulation

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Simulate data
```

## Python

```{python}
#| eval: false
# Simulate data
```

::::

## Running Multiple Chains

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Run chains in parallel with doParallel
```

## Python

```{python}
#| eval: false
# Run chains sequentially or with multiprocessing
```

::::

## Convergence Diagnostics

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Compute R-hat, trace plots
```

## Python

```{python}
#| eval: false
# Compute R-hat, trace plots
```

::::

## Combining Chain Results

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Pool posterior samples across chains
```

## Python

```{python}
#| eval: false
# Pool posterior samples across chains
```

::::
