---
title: "Running and Combining Multiple MCMC Chains"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(rprojroot::find_root(rprojroot::has_file(".here")), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

# Motivation

Mixing of an MCMC sampler is a perennial concern for complex Bayesian models. BART
and BCF are no exception. One common way to address such concerns is to run multiple
independent "chains" of an MCMC sampler, so that if each chain gets stuck in a
different region of the posterior, their combined samples attain better coverage of
the full posterior.

This idea works with the classic "root-initialized" MCMC sampler of @chipman2010bart,
but a key insight of @he2023stochastic and @krantsevich2023stochastic is that the GFR
algorithm may be used to warm-start initialize multiple chains of the BART / BCF MCMC
sampler.

Operationally, the above two approaches have the same implementation (setting
`num_gfr > 0` if warm-start initialization is desired), so this vignette will
demonstrate how to run a multi-chain sampler sequentially or in parallel.

# Setup

::::{.panel-tabset group="language"}

## R

```{r}
#| warning: false
#| message: false
library(stochtree)
library(ggplot2)
library(coda)
library(bayesplot)
library(foreach)
library(doParallel)
```

## Python

```{python}
import numpy as np
import matplotlib.pyplot as plt
import arviz as az
from stochtree import BARTModel

rng = np.random.default_rng(1111)
```

::::

# Demo 1: Supervised Learning

## Data Simulation

Simulate a simple partitioned linear model.

::::{.panel-tabset group="language"}

## R

```{r}
# Generate the data
set.seed(1111)
n <- 500
p_x <- 10
p_w <- 1
snr <- 3
X <- matrix(runif(n * p_x), ncol = p_x)
leaf_basis <- matrix(runif(n * p_w), ncol = p_w)
f_XW <- (((0 <= X[, 1]) & (0.25 > X[, 1])) *
  (-7.5 * leaf_basis[, 1]) +
  ((0.25 <= X[, 1]) & (0.5 > X[, 1])) * (-2.5 * leaf_basis[, 1]) +
  ((0.5 <= X[, 1]) & (0.75 > X[, 1])) * (2.5 * leaf_basis[, 1]) +
  ((0.75 <= X[, 1]) & (1 > X[, 1])) * (7.5 * leaf_basis[, 1]))
noise_sd <- sd(f_XW) / snr
y <- f_XW + rnorm(n, 0, 1) * noise_sd

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct * n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- X[test_inds, ]
X_train <- X[train_inds, ]
leaf_basis_test <- leaf_basis[test_inds, ]
leaf_basis_train <- leaf_basis[train_inds, ]
y_test <- y[test_inds]
y_train <- y[train_inds]
```

## Python

```{python}
n, p_x, p_w, snr = 500, 10, 1, 3
X = rng.uniform(size=(n, p_x))
leaf_basis = rng.uniform(size=(n, p_w))
f_XW = (((0 <= X[:, 0]) & (0.25 > X[:, 0])) * (-7.5 * leaf_basis[:, 0]) +
        ((0.25 <= X[:, 0]) & (0.5 > X[:, 0]))  * (-2.5 * leaf_basis[:, 0]) +
        ((0.5 <= X[:, 0]) & (0.75 > X[:, 0]))  * (2.5  * leaf_basis[:, 0]) +
        ((0.75 <= X[:, 0]) & (1 > X[:, 0]))     * (7.5  * leaf_basis[:, 0]))
noise_sd = np.std(f_XW) / snr
y = f_XW + rng.normal(0, noise_sd, size=n)

test_set_pct = 0.2
n_test = round(test_set_pct * n)
test_inds  = rng.choice(n, n_test, replace=False)
train_inds = np.setdiff1d(np.arange(n), test_inds)
X_test,          X_train          = X[test_inds],          X[train_inds]
leaf_basis_test, leaf_basis_train = leaf_basis[test_inds], leaf_basis[train_inds]
y_test,          y_train          = y[test_inds],          y[train_inds]
```

::::

## Sampling Multiple Chains Sequentially from Scratch

The simplest way to sample multiple chains of a stochtree model is to do so
"sequentially," that is, after chain 1 is sampled, chain 2 is sampled from a
different starting state, and similarly for each of the requested chains. This is
supported internally in both the `bart()` and `bcf()` functions, with the
`num_chains` parameter in the `general_params` list.

Define some high-level parameters, including number of chains to run and number of
samples per chain. Here we run 4 independent chains with 2000 MCMC iterations, each
of which is burned in for 1000 iterations.

::::{.panel-tabset group="language"}

## R

```{r}
num_chains <- 4
num_gfr <- 0
num_burnin <- 1000
num_mcmc <- 2000
```

## Python

```{python}
num_chains = 4
num_gfr    = 0
num_burnin = 1000
num_mcmc   = 2000
```

::::

Run the sampler.

::::{.panel-tabset group="language"}

## R

```{r}
bart_model <- stochtree::bart(
  X_train = X_train,
  leaf_basis_train = leaf_basis_train,
  y_train = y_train,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = list(num_chains = num_chains)
)
```

## Python

```{python}
bart_model = BARTModel()
bart_model.sample(
    X_train=X_train, leaf_basis_train=leaf_basis_train, y_train=y_train,
    num_gfr=num_gfr, num_burnin=num_burnin, num_mcmc=num_mcmc,
    general_params={"num_threads": 1, "num_chains": num_chains},
)
```

::::

Now we have a model with `num_chains * num_mcmc` samples stored internally. These
samples are arranged sequentially, with the first `num_mcmc` samples corresponding
to chain 1, the next `num_mcmc` samples to chain 2, etc.

Since each chain is a set of samples of the same model, we can analyze the samples
collectively, for example, by looking at out-of-sample predictions.

::::{.panel-tabset group="language"}

## R

```{r}
y_hat_test <- predict(
  bart_model,
  X = X_test,
  leaf_basis = leaf_basis_test,
  type = "mean",
  terms = "y_hat"
)
plot(y_hat_test, y_test, xlab = "Predicted", ylab = "Actual")
abline(0, 1, col = "red", lty = 3, lwd = 3)
```

## Python

```{python}
y_hat_test = bart_model.predict(
    X=X_test, leaf_basis=leaf_basis_test, type="mean", terms="y_hat"
)
lo, hi = min(y_hat_test.min(), y_test.min()), max(y_hat_test.max(), y_test.max())
plt.scatter(y_hat_test, y_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual")
plt.show()
```

::::

Now, suppose we want to analyze each of the chains separately to assess mixing /
convergence. We can construct an `mcmc.list` in the `coda` package to perform various
diagnostics.

::::{.panel-tabset group="language"}

## R

```{r}
sigma2_coda_list <- coda::as.mcmc.list(lapply(
  1:num_chains,
  function(chain_idx) {
    offset <- (chain_idx - 1) * num_mcmc
    inds_start <- offset + 1
    inds_end <- offset + num_mcmc
    coda::mcmc(extractParameter(bart_model, "sigma2_global")[inds_start:inds_end])
  }
))
traceplot(sigma2_coda_list, ylab = expression(sigma^2))
abline(h = noise_sd^2, col = "black", lty = 3, lwd = 3)
acf <- autocorr.diag(sigma2_coda_list)
ess <- effectiveSize(sigma2_coda_list)
rhat <- gelman.diag(sigma2_coda_list, autoburnin = F)
cat(paste0(
  "Average autocorrelation across chains:\n",
  paste0(paste0(rownames(acf), ": ", round(acf, 3)), collapse = ", "),
  "\nTotal effective sample size across chains: ",
  paste0(round(ess, 1), collapse = ", "),
  "\n'R-hat' potential scale reduction factor of Gelman and Rubin (1992)): ",
  paste0(round(rhat$psrf[, 1], 3), collapse = ", ")
))
```

## Python

```{python}
# Reshape flat sigma2 samples into (num_chains, num_mcmc) for per-chain diagnostics
# az.from_dict requires nested dict: {"posterior": {"var": array(chains, draws)}}
idata = az.from_dict({"posterior": {"sigma2": bart_model.global_var_samples.reshape(num_chains, num_mcmc)}})

az.plot_trace(idata)
plt.axhline(noise_sd**2, color="black", linestyle="dashed", linewidth=1.5)
plt.show()

print("ESS:  ", az.ess(idata))
print("R-hat:", az.rhat(idata))
az.plot_autocorr(idata)
plt.show()
```

::::

We can convert this to an array to be consumed by the `bayesplot` package.

::::{.panel-tabset group="language"}

## R

```{r}
coda_array <- as.array(sigma2_coda_list)
dim(coda_array) <- c(nrow(coda_array), ncol(coda_array), 1)
dimnames(coda_array) <- list(
  Iteration = paste0("iter", 1:num_mcmc),
  Chain = paste0("chain", 1:num_chains),
  Parameter = "sigma2_global"
)
```

## Python

```{python}
# sigma2_by_chain already has shape (num_chains, num_mcmc) — ready for per-chain plots
sigma2_chains = bart_model.global_var_samples.reshape(num_chains, num_mcmc)
```

::::

From here, we can visualize the posterior of $\sigma^2$ for each chain, comparing
to the true simulated value.

::::{.panel-tabset group="language"}

## R

```{r}
#| warning: false
#| message: false
bayesplot::mcmc_hist_by_chain(
  coda_array,
  pars = "sigma2_global"
) +
  ggplot2::labs(
    title = "Global error scale posterior by chain",
    x = expression(sigma^2)
  ) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(hjust = 0.5)
  ) +
  ggplot2::geom_vline(
    xintercept = noise_sd^2,
    color = "black",
    linetype = "dashed",
    size = 1
  )
```

## Python

```{python}
fig, axes = plt.subplots(1, num_chains, figsize=(12, 3), sharey=True)
for i, ax in enumerate(axes):
    ax.hist(sigma2_chains[i], bins=30)
    ax.axvline(noise_sd**2, color="black", linestyle="dashed", linewidth=1.5)
    ax.set_title(f"Chain {i+1}")
    ax.set_xlabel(r"$\sigma^2$")
fig.suptitle("Global error scale posterior by chain")
plt.tight_layout()
plt.show()
```

::::

## Sampling Multiple Chains Sequentially from XBART Forests

In the example above, each chain was initialized from "root". If we sample a model
using a small number of 'grow-from-root' iterations, we can use these forests to
initialize MCMC chains.

::::{.panel-tabset group="language"}

## R

```{r}
num_chains <- 4
num_gfr <- 5
num_burnin <- 1000
num_mcmc <- 2000
```

## Python

```{python}
num_chains = 4
num_gfr    = 5
num_burnin = 1000
num_mcmc   = 2000
```

::::

Run the initial GFR sampler.

::::{.panel-tabset group="language"}

## R

```{r}
xbart_model <- stochtree::bart(
  X_train = X_train,
  leaf_basis_train = leaf_basis_train,
  y_train = y_train,
  num_gfr = num_gfr,
  num_burnin = 0,
  num_mcmc = 0
)
xbart_model_string <- stochtree::saveBARTModelToJsonString(xbart_model)
```

## Python

```{python}
xbart_model = BARTModel()
xbart_model.sample(
    X_train=X_train, leaf_basis_train=leaf_basis_train, y_train=y_train,
    num_gfr=num_gfr, num_burnin=0, num_mcmc=0,
    general_params={"num_threads": 1},
)
xbart_model_json = xbart_model.to_json()
```

::::

Run the multi-chain BART sampler, with each chain initialized from a different GFR
forest.

::::{.panel-tabset group="language"}

## R

```{r}
bart_model <- stochtree::bart(
  X_train = X_train,
  leaf_basis_train = leaf_basis_train,
  y_train = y_train,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = list(num_chains = num_chains),
  previous_model_json = xbart_model_string,
  previous_model_warmstart_sample_num = num_gfr
)
```

## Python

```{python}
bart_model = BARTModel()
bart_model.sample(
    X_train=X_train, leaf_basis_train=leaf_basis_train, y_train=y_train,
    num_gfr=0, num_burnin=num_burnin, num_mcmc=num_mcmc,
    general_params={"num_threads": 1, "num_chains": num_chains},
    previous_model_json=xbart_model_json,
    previous_model_warmstart_sample_num=num_gfr - 1,  # 0-indexed
)
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
y_hat_test <- predict(
  bart_model,
  X = X_test,
  leaf_basis = leaf_basis_test,
  type = "mean",
  terms = "y_hat"
)
plot(y_hat_test, y_test, xlab = "Predicted", ylab = "Actual")
abline(0, 1, col = "red", lty = 3, lwd = 3)
```

## Python

```{python}
y_hat_test = bart_model.predict(
    X=X_test, leaf_basis=leaf_basis_test, type="mean", terms="y_hat"
)
lo, hi = min(y_hat_test.min(), y_test.min()), max(y_hat_test.max(), y_test.max())
plt.scatter(y_hat_test, y_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted"); plt.ylabel("Actual")
plt.show()
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
sigma2_coda_list <- coda::as.mcmc.list(lapply(
  1:num_chains,
  function(chain_idx) {
    offset <- (chain_idx - 1) * num_mcmc
    inds_start <- offset + 1
    inds_end <- offset + num_mcmc
    coda::mcmc(extractParameter(bart_model, "sigma2_global")[inds_start:inds_end])
  }
))
traceplot(sigma2_coda_list, ylab = expression(sigma^2))
abline(h = noise_sd^2, col = "black", lty = 3, lwd = 3)
acf <- autocorr.diag(sigma2_coda_list)
ess <- effectiveSize(sigma2_coda_list)
rhat <- gelman.diag(sigma2_coda_list, autoburnin = F)
cat(paste0(
  "Average autocorrelation across chains:\n",
  paste0(paste0(rownames(acf), ": ", round(acf, 3)), collapse = ", "),
  "\nTotal effective sample size across chains: ",
  paste0(round(ess, 1), collapse = ", "),
  "\n'R-hat' potential scale reduction factor of Gelman and Rubin (1992)): ",
  paste0(round(rhat$psrf[, 1], 3), collapse = ", ")
))
```

## Python

```{python}
idata = az.from_dict({"posterior": {"sigma2": bart_model.global_var_samples.reshape(num_chains, num_mcmc)}})

az.plot_trace(idata)
plt.axhline(noise_sd**2, color="black", linestyle="dashed", linewidth=1.5)
plt.show()

print("ESS:  ", az.ess(idata))
print("R-hat:", az.rhat(idata))
az.plot_autocorr(idata)
plt.show()
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
coda_array <- as.array(sigma2_coda_list)
dim(coda_array) <- c(nrow(coda_array), ncol(coda_array), 1)
dimnames(coda_array) <- list(
  Iteration = paste0("iter", 1:num_mcmc),
  Chain = paste0("chain", 1:num_chains),
  Parameter = "sigma2_global"
)
```

## Python

```{python}
sigma2_chains = bart_model.global_var_samples.reshape(num_chains, num_mcmc)
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
#| warning: false
#| message: false
bayesplot::mcmc_hist_by_chain(
  coda_array,
  pars = "sigma2_global"
) +
  ggplot2::labs(
    title = "Global error scale posterior by chain",
    x = expression(sigma^2)
  ) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(hjust = 0.5)
  ) +
  ggplot2::geom_vline(
    xintercept = noise_sd^2,
    color = "black",
    linetype = "dashed",
    size = 1
  )
```

## Python

```{python}
fig, axes = plt.subplots(1, num_chains, figsize=(12, 3), sharey=True)
for i, ax in enumerate(axes):
    ax.hist(sigma2_chains[i], bins=30)
    ax.axvline(noise_sd**2, color="black", linestyle="dashed", linewidth=1.5)
    ax.set_title(f"Chain {i+1}")
    ax.set_xlabel(r"$\sigma^2$")
fig.suptitle("Global error scale posterior by chain")
plt.tight_layout()
plt.show()
```

::::

## Sampling Multiple Chains in Parallel

While the above examples used sequential multi-chain sampling internally, it is also
possible to run chains in parallel. In R, this is done via `doParallel` / `foreach`;
in Python, via `concurrent.futures.ProcessPoolExecutor`. In both cases, each chain
is serialized to JSON for cross-process communication, then combined into a single
model via `createBARTModelFromCombinedJsonString()` (R) or
`BARTModel.from_json_string_list()` (Python).

In order to run multiple parallel stochtree chains in R, a parallel backend must be
registered. Note that we do not evaluate the cluster setup code below in order to
interact nicely with GitHub Actions.

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
ncores <- parallel::detectCores()
cl <- makeCluster(ncores)
registerDoParallel(cl)
```

## Python

```{python}
#| eval: false
# Worker function must be defined at module level for pickling
from concurrent.futures import ProcessPoolExecutor

def _run_bart_chain(args):
    X_tr, lb_tr, y_tr, X_te, lb_te, num_burnin, num_mcmc, seed = args
    from stochtree import BARTModel
    m = BARTModel()
    m.sample(
        X_train=X_tr, leaf_basis_train=lb_tr, y_train=y_tr,
        X_test=X_te, leaf_basis_test=lb_te,
        num_gfr=0, num_burnin=num_burnin, num_mcmc=num_mcmc,
        general_params={"num_threads": 1, "random_seed": seed},
        mean_forest_params={"sample_sigma2_leaf": False},
    )
    return m.to_json(), m.y_hat_test
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
num_chains <- 4
num_gfr <- 0
num_burnin <- 100
num_mcmc <- 100
```

## Python

```{python}
num_chains = 4
num_gfr    = 0
num_burnin = 100
num_mcmc   = 100
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
bart_model_outputs <- foreach(i = 1:num_chains) %dopar%
  {
    random_seed <- i
    general_params <- list(sample_sigma2_global = T, random_seed = random_seed)
    mean_forest_params <- list(sample_sigma2_leaf = F)
    bart_model <- stochtree::bart(
      X_train = X_train,
      leaf_basis_train = leaf_basis_train,
      y_train = y_train,
      X_test = X_test,
      leaf_basis_test = leaf_basis_test,
      num_gfr = num_gfr,
      num_burnin = num_burnin,
      num_mcmc = num_mcmc,
      general_params = general_params,
      mean_forest_params = mean_forest_params
    )
    bart_model_string <- stochtree::saveBARTModelToJsonString(bart_model)
    y_hat_test <- extractParameter(bart_model, "y_hat_test")
    list(model = bart_model_string, yhat = y_hat_test)
  }
```

## Python

```{python}
# Sequential loop — replace the loop body with ProcessPoolExecutor for true parallelism
bart_model_outputs = []
for i in range(num_chains):
    m = BARTModel()
    m.sample(
        X_train=X_train, leaf_basis_train=leaf_basis_train, y_train=y_train,
        X_test=X_test, leaf_basis_test=leaf_basis_test,
        num_gfr=0, num_burnin=num_burnin, num_mcmc=num_mcmc,
        general_params={"num_threads": 1, "sample_sigma2_global": True, "random_seed": i + 1},
        mean_forest_params={"sample_sigma2_leaf": False},
    )
    bart_model_outputs.append({"model": m.to_json(), "yhat": m.y_hat_test})
```

::::

Close the parallel cluster (not evaluated here).

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
stopCluster(cl)
```

## Python

```{python}
# No explicit teardown required when using concurrent.futures context manager
```

::::

Combine the forests from each BART model into a single forest.

::::{.panel-tabset group="language"}

## R

```{r}
bart_model_strings <- list()
bart_model_yhats <- matrix(NA, nrow = length(y_test), ncol = num_chains)
for (i in 1:length(bart_model_outputs)) {
  bart_model_strings[[i]] <- bart_model_outputs[[i]]$model
  bart_model_yhats[, i] <- rowMeans(bart_model_outputs[[i]]$yhat)
}
combined_bart <- createBARTModelFromCombinedJsonString(bart_model_strings)
```

## Python

```{python}
bart_model_strings = [out["model"] for out in bart_model_outputs]
bart_model_yhats = np.column_stack([
    out["yhat"].mean(axis=1) for out in bart_model_outputs
])  # shape: (n_test, num_chains)
combined_bart = BARTModel()
combined_bart.from_json_string_list(bart_model_strings)
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
yhat_combined <- predict(combined_bart, X_test, leaf_basis_test)$y_hat
```

## Python

```{python}
# type="posterior" (default) returns the full n_test × (num_chains * num_mcmc) matrix
yhat_combined = combined_bart.predict(X=X_test, leaf_basis=leaf_basis_test, terms="y_hat")
```

::::

Compare average predictions from each chain to the original predictions and to
the true $y$ values.

::::{.panel-tabset group="language"}

## R

```{r}
par(mfrow = c(1, 2))
for (i in 1:num_chains) {
  offset <- (i - 1) * num_mcmc
  inds_start <- offset + 1
  inds_end <- offset + num_mcmc
  plot(
    rowMeans(yhat_combined[, inds_start:inds_end]),
    bart_model_yhats[, i],
    xlab = "deserialized",
    ylab = "original",
    main = paste0("Chain ", i, "\nPredictions")
  )
  abline(0, 1, col = "red", lty = 3, lwd = 3)
}
par(mfrow = c(1, 1))
```

## Python

```{python}
fig, axes = plt.subplots(2, 2, figsize=(8, 8))
for i, ax in enumerate(axes.flat):
    chain_combined = yhat_combined[:, i * num_mcmc:(i + 1) * num_mcmc].mean(axis=1)
    chain_orig = bart_model_yhats[:, i]
    lo = min(chain_combined.min(), chain_orig.min())
    hi = max(chain_combined.max(), chain_orig.max())
    ax.scatter(chain_combined, chain_orig, alpha=0.4, s=10)
    ax.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=1.5)
    ax.set_xlabel("Deserialized"); ax.set_ylabel("Original")
    ax.set_title(f"Chain {i+1} Predictions")
plt.tight_layout()
plt.show()
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
par(mfrow = c(1, 2))
for (i in 1:num_chains) {
  offset <- (i - 1) * num_mcmc
  inds_start <- offset + 1
  inds_end <- offset + num_mcmc
  plot(
    rowMeans(yhat_combined[, inds_start:inds_end]),
    y_test,
    xlab = "predicted",
    ylab = "actual",
    main = paste0("Chain ", i, "\nPredictions")
  )
  abline(0, 1, col = "red", lty = 3, lwd = 3)
}
par(mfrow = c(1, 1))
```

## Python

```{python}
fig, axes = plt.subplots(2, 2, figsize=(8, 8))
for i, ax in enumerate(axes.flat):
    chain_pred = yhat_combined[:, i * num_mcmc:(i + 1) * num_mcmc].mean(axis=1)
    lo = min(chain_pred.min(), y_test.min())
    hi = max(chain_pred.max(), y_test.max())
    ax.scatter(chain_pred, y_test, alpha=0.4, s=10)
    ax.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=1.5)
    ax.set_xlabel("Predicted"); ax.set_ylabel("Actual")
    ax.set_title(f"Chain {i+1} Predictions")
plt.tight_layout()
plt.show()
```

::::

# References
