---
title: "Ordinal Outcome Modeling"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(here::here(), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

This vignette demonstrates how to use BART to model ordinal outcomes with a
complementary log-log (cloglog) link function (@alam2025unified).

## Introduction to Ordinal BART with CLogLog Link

Ordinal data refers to outcomes that have a natural ordering but undefined distances
between categories. Examples include survey responses (strongly disagree, disagree,
neutral, agree, strongly agree), severity ratings (mild, moderate, severe), or
educational levels (elementary, high school, college, graduate).

The complementary log-log (cloglog) model uses the cumulative link function
$$
\text{cloglog}(p) = \log(-\log(1-p))
$$
to express cumulative category probabilities as a function of covariates
$$
\text{cloglog}(P(Y \leq k \mid X = x)) = \log(-\log(1-P(Y \leq k \mid X = x))) = \gamma_k + \lambda(x)
$$

This link function is asymmetric and particularly appropriate when the probability of
being in higher categories changes rapidly at certain thresholds, making it different
from the symmetric probit or logit links commonly used in ordinal regression.

In `stochtree`, we let $\lambda(x)$ be represented by a stochastic tree ensemble.

## Setup

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
```

## Python

```{python}
import numpy as np
import matplotlib.pyplot as plt
from stochtree import BARTModel, OutcomeModel
```

::::

## Data Simulation

We simulate a dataset with an ordinal outcome with three categories,
$y_i \in \left\{1,2,3\right\}$ whose probabilities depend on covariates, $X$.

::::{.panel-tabset group="language"}

## R

```{r}
# Set seed
random_seed <- 2026
set.seed(random_seed)

# Sample size and number of predictors
n <- 2000
p <- 5

# Design matrix and true lambda function
X <- matrix(rnorm(n * p), n, p)
beta <- rep(1 / sqrt(p), p)
true_lambda_function <- X %*% beta

# Set cutpoints for ordinal categories (3 categories: 1, 2, 3)
n_categories <- 3
gamma_true <- c(-2, 1)
ordinal_cutpoints <- log(cumsum(exp(gamma_true)))

# True ordinal class probabilities
true_probs <- matrix(0, nrow = n, ncol = n_categories)
for (j in 1:n_categories) {
  if (j == 1) {
    true_probs[, j] <- 1 - exp(-exp(gamma_true[j] + true_lambda_function))
  } else if (j == n_categories) {
    true_probs[, j] <- 1 - rowSums(true_probs[, 1:(j - 1), drop = FALSE])
  } else {
    true_probs[, j] <- exp(-exp(gamma_true[j - 1] + true_lambda_function)) *
      (1 - exp(-exp(gamma_true[j] + true_lambda_function)))
  }
}

# Generate ordinal outcomes
y <- sapply(1:nrow(X), function(i) {
  sample(1:n_categories, 1, prob = true_probs[i, ])
})
cat("Outcome distribution:", table(y), "\n")

# Train test split
train_idx <- sample(1:n, size = floor(0.8 * n))
test_idx <- setdiff(1:n, train_idx)
X_train <- X[train_idx, ]
y_train <- y[train_idx]
X_test <- X[test_idx, ]
y_test <- y[test_idx]
```

## Python

```{python}
random_seed = 2026
rng = np.random.default_rng(random_seed)

# Sample size and number of predictors
n = 2000
p = 5

# Design matrix and true lambda function
X = rng.standard_normal((n, p))
beta = np.ones(p) / np.sqrt(p)
true_lambda = X @ beta

# Set cutpoints for ordinal categories (3 categories: 1, 2, 3)
n_categories = 3
gamma_true = np.array([-2.0, 1.0])

# True ordinal class probabilities
true_probs = np.zeros((n, n_categories))
true_probs[:, 0] = 1 - np.exp(-np.exp(gamma_true[0] + true_lambda))
for j in range(1, n_categories - 1):
    true_probs[:, j] = (
        np.exp(-np.exp(gamma_true[j - 1] + true_lambda))
        * (1 - np.exp(-np.exp(gamma_true[j] + true_lambda)))
    )
true_probs[:, n_categories - 1] = 1 - true_probs[:, :-1].sum(axis=1)

# Generate ordinal outcomes (1-indexed integers)
y = np.array(
    [rng.choice(np.arange(1, n_categories + 1), p=true_probs[i]) for i in range(n)],
    dtype=float,
)
unique, counts = np.unique(y, return_counts=True)
print("Outcome distribution:", dict(zip(unique.astype(int), counts)))

# Train-test split
n_test = round(0.2 * n)
n_train = n - n_test
test_inds = rng.choice(n, n_test, replace=False)
train_inds = np.setdiff1d(np.arange(n), test_inds)
X_train = X[train_inds]
X_test = X[test_inds]
y_train = y[train_inds]
y_test = y[test_inds]
```

::::

## Model Fitting

We specify the cloglog link function for modeling an ordinal outcome by setting
`outcome_model=OutcomeModel(outcome="ordinal", link="cloglog")` in the
`general_params` argument list. Since ordinal outcomes are incompatible with the
Gaussian global error variance model, we also set `sample_sigma2_global=FALSE`.

We also override the default `num_trees` for the mean forest (200) in favor of
greater regularization for the ordinal model and set `sample_sigma2_leaf=FALSE`.

::::{.panel-tabset group="language"}

## R

```{r}
# Sample the cloglog ordinal BART model
bart_model <- bart(
  X_train = X_train,
  y_train = y_train,
  X_test = X_test,
  num_gfr = 0,
  num_burnin = 1000,
  num_mcmc = 1000,
  general_params = list(
    cutpoint_grid_size = 100,
    sample_sigma2_global = FALSE,
    keep_every = 1,
    num_chains = 1,
    verbose = FALSE,
    random_seed = random_seed,
    outcome_model = OutcomeModel(outcome = 'ordinal', link = 'cloglog')
  ),
  mean_forest_params = list(num_trees = 50, sample_sigma2_leaf = FALSE)
)
```

## Python

```{python}
bart_model = BARTModel()
bart_model.sample(
    X_train=X_train,
    y_train=y_train,
    X_test=X_test,
    num_gfr=0,
    num_burnin=1000,
    num_mcmc=1000,
    general_params={
        "num_threads": 1,
        "cutpoint_grid_size": 100,
        "sample_sigma2_global": False,
        "keep_every": 1,
        "num_chains": 1,
        "random_seed": random_seed,
        "outcome_model": OutcomeModel(outcome="ordinal", link="cloglog"),
    },
    mean_forest_params={"num_trees": 50, "sample_sigma2_leaf": False},
)
```

::::

## Prediction

As with any other BART model in `stochtree`, we can use the `predict` function on
our ordinal model. Specifying `scale = "linear"` and `terms = "y_hat"` will simply
return predictions from the estimated $\lambda(x)$ function, but users can estimate
class probabilities via `scale = "probability"`, which by default returns an array of
dimension (`num_observations`, `num_categories`, `num_samples`). Specifying
`type = "mean"` collapses the output to a `num_observations` x `num_categories`
matrix with the average posterior class probability for each observation. Users can
also specify `type = "class"` for the maximum a posteriori (MAP) class label estimate
for each draw of each observation.

Below we compute the posterior class probabilities for the train and test sets.

::::{.panel-tabset group="language"}

## R

```{r}
est_probs_train <- predict(
  bart_model,
  X = X_train,
  scale = "probability",
  terms = "y_hat"
)
est_probs_test <- predict(
  bart_model,
  X = X_test,
  scale = "probability",
  terms = "y_hat"
)
```

## Python

```{python}
# predict returns (n_obs, n_categories) posterior mean class probabilities
est_probs_train = bart_model.predict(X=X_train, scale="probability", terms="y_hat", type="mean")
est_probs_test = bart_model.predict(X=X_test, scale="probability", terms="y_hat", type="mean")
```

::::

## Model Results and Interpretation

Since one of the "cutpoints" is fixed for identifiability, we plot the posterior
distributions of the other two cutpoints and compare them to their true simulated
values (blue dotted lines).

The cutpoint samples are accessed via `extractParameter(bart_model, "cloglog_cutpoints")`
(shape: `(n_categories - 1, num_samples)`) and are shifted by the per-sample mean of
the training predictions to account for the non-identifiable intercept.

::::{.panel-tabset group="language"}

## R

```{r}
y_hat_train_post <- predict(
  bart_model,
  X = X_train,
  scale = "linear",
  terms = "y_hat",
  type = "posterior"
)
cutpoint_samples <- extractParameter(bart_model, "cloglog_cutpoints")
gamma1 <- cutpoint_samples[1, ] + colMeans(y_hat_train_post)
hist(
  gamma1,
  main = "Posterior Distribution of Cutpoint 1",
  xlab = "Cutpoint 1",
  freq = FALSE
)
abline(v = gamma_true[1], col = 'blue', lty = 3, lwd = 3)
gamma2 <- cutpoint_samples[2, ] + colMeans(y_hat_train_post)
hist(
  gamma2,
  main = "Posterior Distribution of Cutpoint 2",
  xlab = "Cutpoint 2",
  freq = FALSE
)
abline(v = gamma_true[2], col = 'blue', lty = 3, lwd = 3)
```

## Python

```{python}
# cutpoint_samples shape: (n_categories - 1, num_samples)
# shifted by per-sample mean of train predictions to remove non-identifiable intercept
cutpoint_samples = bart_model.extract_parameter("cloglog_cutpoints")
y_hat_train_post = bart_model.predict(X=X_train, scale="linear", terms="y_hat", type="posterior")
gamma1 = cutpoint_samples[0, :] + y_hat_train_post.mean(axis=0)
gamma2 = cutpoint_samples[1, :] + y_hat_train_post.mean(axis=0)

fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 4))
ax1.hist(gamma1, density=True, bins=40)
ax1.axvline(gamma_true[0], color="blue", linestyle="dotted", linewidth=2)
ax1.set_title("Posterior Distribution of Cutpoint 1")
ax1.set_xlabel("Cutpoint 1")
ax2.hist(gamma2, density=True, bins=40)
ax2.axvline(gamma_true[1], color="blue", linestyle="dotted", linewidth=2)
ax2.set_title("Posterior Distribution of Cutpoint 2")
ax2.set_xlabel("Cutpoint 2")
plt.tight_layout()
plt.show()
```

::::

We can compare the true value of the latent "utility function" $\lambda(x)$ to the
(mean-shifted) BART forest predictions.

::::{.panel-tabset group="language"}

## R

```{r}
# Train set predicted versus actual
y_hat_train <- predict(
  bart_model,
  X = X_train,
  scale = "linear",
  terms = "y_hat",
  type = "mean"
)
lambda_pred_train <- y_hat_train - mean(y_hat_train)
plot(
  lambda_pred_train,
  true_lambda_function[train_idx],
  main = "Train Set: Predicted vs Actual",
  xlab = "Predicted",
  ylab = "Actual"
)
abline(a = 0, b = 1, col = 'blue', lwd = 2)
cor_train <- cor(true_lambda_function[train_idx], lambda_pred_train)
text(
  min(lambda_pred_train),
  max(true_lambda_function[train_idx]),
  paste('Correlation:', round(cor_train, 3)),
  adj = 0,
  col = 'red'
)

# Test set predicted versus actual
y_hat_test <- predict(
  bart_model,
  X = X_test,
  scale = "linear",
  terms = "y_hat",
  type = "mean"
)
lambda_pred_test <- y_hat_test - mean(y_hat_test)
plot(
  lambda_pred_test,
  true_lambda_function[test_idx],
  main = "Test Set: Predicted vs Actual",
  xlab = "Predicted",
  ylab = "Actual"
)
abline(a = 0, b = 1, col = 'blue', lwd = 2)
cor_test <- cor(true_lambda_function[test_idx], lambda_pred_test)
text(
  min(lambda_pred_test),
  max(true_lambda_function[test_idx]),
  paste('Correlation:', round(cor_test, 3)),
  adj = 0,
  col = 'red'
)
```

## Python

```{python}
y_hat_train = bart_model.predict(X=X_train, scale="linear", terms="y_hat", type="mean")
y_hat_test = bart_model.predict(X=X_test, scale="linear", terms="y_hat", type="mean")
lambda_pred_train = y_hat_train - y_hat_train.mean()
lambda_pred_test = y_hat_test - y_hat_test.mean()
corr_train = np.corrcoef(true_lambda[train_inds], lambda_pred_train)[0, 1]
corr_test = np.corrcoef(true_lambda[test_inds], lambda_pred_test)[0, 1]

fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 5))
ax1.scatter(lambda_pred_train, true_lambda[train_inds], alpha=0.3, s=10)
ax1.axline((0, 0), slope=1, color="blue", linewidth=2)
ax1.set_title("Train Set: Predicted vs Actual")
ax1.set_xlabel("Predicted")
ax1.set_ylabel("Actual")
ax1.text(0.05, 0.95, f"Correlation: {corr_train:.3f}", transform=ax1.transAxes,
         color="red", verticalalignment="top")
ax2.scatter(lambda_pred_test, true_lambda[test_inds], alpha=0.3, s=10)
ax2.axline((0, 0), slope=1, color="blue", linewidth=2)
ax2.set_title("Test Set: Predicted vs Actual")
ax2.set_xlabel("Predicted")
ax2.set_ylabel("Actual")
ax2.text(0.05, 0.95, f"Correlation: {corr_test:.3f}", transform=ax2.transAxes,
         color="red", verticalalignment="top")
plt.tight_layout()
plt.show()
```

::::

Finally, we compare the estimated class probabilities with their true simulated values
for each class on the training set.

::::{.panel-tabset group="language"}

## R

```{r}
for (j in 1:n_categories) {
  mean_probs <- rowMeans(est_probs_train[, j, ])
  plot(
    true_probs[train_idx, j],
    mean_probs,
    main = paste("Training Set: True vs Estimated Probability, Class", j),
    xlab = "True Class Probability",
    ylab = "Estimated Class Probability"
  )
  abline(a = 0, b = 1, col = 'blue', lwd = 2)
  cor_train_prob <- cor(true_probs[train_idx, j], mean_probs)
  text(
    min(true_probs[train_idx, j]),
    max(mean_probs),
    paste('Correlation:', round(cor_train_prob, 3)),
    adj = 0,
    col = 'red'
  )
}
```

## Python

```{python}
fig, axes = plt.subplots(1, n_categories, figsize=(15, 5))
for j in range(n_categories):
    corr = np.corrcoef(true_probs[train_inds, j], est_probs_train[:, j])[0, 1]
    axes[j].scatter(true_probs[train_inds, j], est_probs_train[:, j], alpha=0.3, s=10)
    axes[j].axline((0, 0), slope=1, color="blue", linewidth=2)
    axes[j].set_title(f"Training Set: True vs Estimated Probability, Class {j + 1}")
    axes[j].set_xlabel("True Class Probability")
    axes[j].set_ylabel("Estimated Class Probability")
    axes[j].text(0.05, 0.95, f"Correlation: {corr:.3f}", transform=axes[j].transAxes,
                 color="red", verticalalignment="top")
plt.tight_layout()
plt.show()
```

::::

And the same comparison on the test set.

::::{.panel-tabset group="language"}

## R

```{r}
for (j in 1:n_categories) {
  mean_probs <- rowMeans(est_probs_test[, j, ])
  plot(
    true_probs[test_idx, j],
    mean_probs,
    main = paste("Test Set: True vs Estimated Probability, Class", j),
    xlab = "True Class Probability",
    ylab = "Estimated Class Probability"
  )
  abline(a = 0, b = 1, col = 'blue', lwd = 2)
  cor_test_prob <- cor(true_probs[test_idx, j], mean_probs)
  text(
    min(true_probs[test_idx, j]),
    max(mean_probs),
    paste('Correlation:', round(cor_test_prob, 3)),
    adj = 0,
    col = 'red'
  )
}
```

## Python

```{python}
fig, axes = plt.subplots(1, n_categories, figsize=(15, 5))
for j in range(n_categories):
    corr = np.corrcoef(true_probs[test_inds, j], est_probs_test[:, j])[0, 1]
    axes[j].scatter(true_probs[test_inds, j], est_probs_test[:, j], alpha=0.3, s=10)
    axes[j].axline((0, 0), slope=1, color="blue", linewidth=2)
    axes[j].set_title(f"Test Set: True vs Estimated Probability, Class {j + 1}")
    axes[j].set_xlabel("True Class Probability")
    axes[j].set_ylabel("Estimated Class Probability")
    axes[j].text(0.05, 0.95, f"Correlation: {corr:.3f}", transform=axes[j].transAxes,
                 color="red", verticalalignment="top")
plt.tight_layout()
plt.show()
```

::::

# References
