---
title: "Model Prediction and Parameter Extraction"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
code-annotations: hover
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(
      rprojroot::find_root(rprojroot::has_file(".here")),
      ".venv",
      "bin",
      "python"
    )
  ),
  required = TRUE
)
```

# Introduction

The vignettes in @sec-core-models demonstrate how to **fit** different types of models in `stochtree`. 
This vignette walks through the interfaces for querying and extracting parameters from fitted `stochtree` models.

There are two primary interfaces for interacting with trained BART / BCF models:

1. **Prediction**: obtaining estimates for model terms that depend on covariates or other input data
    - In R, this is governed by an overloaded `predict()` function that specializes to BART and BCF so that `predict(bart_model, ...)` and `predict(bcf_model, ...)` return the requested predictions from the provided model
    - In Python, the `BARTModel` and `BCFModel` classes define `predict()` methods that can be called directly to obtain model predictions
2. **Parameter Extraction**: obtaining samples / traces of parametric model terms (such as the global error scale, $\sigma^2$)
    - In R, this is governed by an overloaded `extractParameter()` function that specializes to BART and BCF so that `extractParameter(bart_model, ...)` and `extractParameter(bcf_model, ...)` return the requested samples from the provided model
    - In Python, the `BARTModel` and `BCFModel` classes define `extract_parameter()` methods that can be called directly to obtain parameter estimates

## Prediction

The prediction interface is designed to be as flexible as the class of models that `stochtree` supports. The broad pattern that input data is provided for a given model and the corresponding predict method returns predictions for one or more model terms, with some optional user-requested transformations.

### BART

The BART Prediction API differs stylistically in R and Python but provides the same functionality, which we annotate below

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
predict(
  object, # <1>
  X, # <2>
  leaf_basis = NULL, # <3>
  rfx_group_ids = NULL, # <4>
  rfx_basis = NULL, # <5>
  type = "posterior", # <6>
  terms = "all", # <7>
  scale = "linear" # <8>
)
```

1. Reference to the sampled BART model
2. Covariates: all BART models require covariates so these are not optional in the predict method
3. Leaf basis: required if the original model was fit with a leaf basis
4. Random effects group labels: required if the original model was fit with an additive random effects term
5. Random effects basis: required if the original model was fit with an additive random effects term with custom basis (i.e. `model_spec` was not `"intercept_only"`)
6. Whether predictions should return each posterior draw (`'posterior'`) or be averaged across the posterior (`'mean'`)
7. Which model terms to return predictions for. `'all'` returns every model term available, `'y_hat'` returns only outcome predictions, `'mean_forest'` returns mean forest predictions (if available), `'rfx'` returns random effects predictions (if available), and `'variance_forest'` returns variance forest predictions (if available)
8. Whether predictions should be returned as-is from forest / RFX terms (`scale = 'linear'`), transformed to probabilities (`scale = 'probability'`) or class labels (`scale = 'class'`)

## Python

```{python}
#| eval: false
bart.BARTModel.predict(     # <1>
  X,                        # <2>
  leaf_basis=None,          # <3>
  rfx_group_ids=None,       # <4>
  rfx_basis=None,           # <5>
  type='posterior',         # <6>
  terms='all',              # <7>
  scale='linear',           # <8>
)
```

1. `.predict()` called from a reference to the sampled BART model
2. Covariates: all BART models require covariates so these are not optional in the predict method
3. Leaf basis: required if the original model was fit with a leaf basis
4. Random effects group labels: required if the original model was fit with an additive random effects term
5. Random effects basis: required if the original model was fit with an additive random effects term with custom basis (i.e. `model_spec` was not `"intercept_only"`)
6. Whether predictions should return each posterior draw (`'posterior'`) or be averaged across the posterior (`'mean'`)
7. Which model terms to return predictions for. `'all'` returns every model term available, `'y_hat'` returns only outcome predictions, `'mean_forest'` returns mean forest predictions (if available), `'rfx'` returns random effects predictions (if available), and `'variance_forest'` returns variance forest predictions (if available)
8. Whether predictions should be returned as-is from forest / RFX terms (`scale = 'linear'`), transformed to probabilities (`scale = 'probability'`) or class labels (`scale = 'class'`)

::::

The data inputs to this API are fairly straightforward, but the `type`, `terms` and `scale` arguments merit some detailed (language-agnostic) explanation.

The [BART overview page](../about/bart.qmd) has a more in-depth review of the BART model, but the most complex version a BART model that `stochtree` supports is, mathematically,
$$
\begin{aligned}
y_{i,j} \mid \Psi_{i,j} = \psi_{i,j}, X_{i,j} = x_{i,j}, W_{i,j} = w_{i,j} &\sim \mathcal{N}\left(w_{i,j} \vec{\gamma}_j + f(x_{i,j}, \psi_{i,j}), \sigma^2_0 \exp{h(x_{i,j})}\right),\\
f &\sim \text{BART}(\alpha_f, \beta_f, m_f), \\
h &\sim \text{logBART}(\alpha_h, \beta_h, m_h), \\
\vec{\gamma}_j &\sim \mathrm{N}(\vec{0}, \sigma^2_{\gamma} I).
\end{aligned}
$$

This model has:

1. Linear leaf regression on a basis $\Psi$
2. Forest-based heteroskedasticity, defined by a $\text{logBART}$ prior on $h$
3. Additive random effects on basis $W$

#### Prediction Terms

Each of the individual options for the `terms` argument can be mapped to the equation above:

| Term | Mathematical Representation |
| --- | --- |
| `y_hat` | $w_{i,j} \vec{\gamma}_j + f(x_{i,j}, \psi_{i,j})$ |
| `mean_forest` | $f(x_{i,j}, \psi_{i,j})$ |
| `rfx` | $w_{i,j} \vec{\gamma}_j$ |
| `variance_forest` | $\sigma^2_0 \exp{h(x_{i,j})}$ |

Requesting `terms = 'all'` will yield a list / dictionary with each of the terms above stored as separate entries. For models without a variance forest, `terms = 'variance_forest'` will be ignored. For models without random effects, `terms = 'rfx'` will be ignored and `y_hat` collapses to $f(x_{i}, \psi_{i})$. For models without leaf regression, `mean_forest` and `y_hat` collapse to $f(x_i)$.

#### Prediction Type

If a BART model has been sampled for $mc$ iterations, then most of the above terms will be $n \times mc$ matrices, where $n$ is the number of rows in `X`, representing a posterior distribution over $f$, $h$ or the conditional outcome mean. This can be converted to a "simple" $n$-dimensional posterior mean (useful for prediction when `terms = 'y_hat'`, for example) by averaging over the $mc$ columns. This use case is common enough that `stochtree` will perform the aggregation before returning to users if `type = 'mean'` is specified.

#### Prediction Scale

In addition to the Gaussian model for continuous outcomes in the equation above, `stochtree` supports modeling binary outcomes through a probit link and ordinal outcomes through a cloglog link. Both of these models apply transformations to model discrete outcomes with continuous forest predictions, and the `scale` argument lets users determine which transformation (if any) to apply to model predictions.

**Probit**

The probit model precludes a variance forest, but works with both random effects and leaf regression. Specifying `scale = 'linear'` returns predictions as returned by the mean forest and random effects terms, `scale = 'probability'` converts the predictions into that term's contribution to the conditional probability of `y = 1`, and `scale = 'class'` returns the category index with the highest predicted probability.

| Term | `scale = 'linear'` | `scale = 'probability'` |
| --- | --- | --- |
| `y_hat` | $w_{i,j} \vec{\gamma}_j + f(x_{i,j}, \psi_{i,j})$ | $\Phi\left(w_{i,j} \vec{\gamma}_j + f(x_{i,j}, \psi_{i,j})\right)$ |
| `mean_forest` | $f(x_{i,j}, \psi_{i,j})$ | $\Phi\left(f(x_{i,j}, \psi_{i,j})\right)$ |
| `rfx` | $w_{i,j} \vec{\gamma}_j$ | $\Phi\left(w_{i,j} \vec{\gamma}_j\right)$ |

where $\Phi(\cdot)$ is the standard normal CDF.

**Cloglog**

The cloglog model precludes a variance forest, random effects, or leaf regression. Specifying `scale = 'linear'` returns predictions as returned by the mean forest, `scale = 'probability'` converts the predictions into the conditional probability of each outcome category, and `scale = 'class'` returns the category index with the highest predicted probability.

| Term | `scale = 'linear'` | `scale = 'probability'` |
| --- | --- | --- |
| `y_hat` | $f(x_{i})$ | $\exp(-\exp(c_{k-1}+f(x_i))) - \exp(-\exp(c_{k}+f(x_i)))$ |
| `mean_forest` | $f(x_{i})$ | $\exp(-\exp(c_{k-1}+f(x_i))) - \exp(-\exp(c_{k}+f(x_i)))$ |

Note that $k$ denotes a category index and $c_k$ is a category-specific "cutpoint" parameter unique to the cloglog model.

### BCF

The multilingual BCF Prediction API is annotated below as with the BART API

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
predict(
  object, # <1>
  X, # <2>
  Z, # <3>
  propensity = NULL, # <4>
  rfx_group_ids = NULL, # <5>
  rfx_basis = NULL, # <6>
  type = "posterior", # <7>
  terms = "all", # <8>
  scale = "linear" # <9>
)
```

1. Reference to the sampled BCF model
2. Covariates: all BCF models require covariates so these are not optional in the predict method
3. Treatment: all BCF models require treatment so this is not optional in the predict method
4. Propensity: required if the original model was fit with a user-supplied propensity score
5. Random effects group labels: required if the original model was fit with an additive random effects term
6. Random effects basis: required if the original model was fit with an additive random effects term with custom basis (i.e. `model_spec` was not `"intercept_only"`)
7. Whether predictions should return each posterior draw (`'posterior'`) or be averaged across the posterior (`'mean'`)
8. Which model terms to return predictions for. `'all'` returns every model term available, `'y_hat'` returns only outcome predictions, `'mu'` returns prognostic forest predictions, `'tau'` returns treatment effect forest predictions (and $\tau_0$ intercept, if available), `'prognostic_forest'` returns prognostic forest predictions and any random intercepts modeled by a random effects `model_spec` of `'intercept_only'` or `'intercept_plus_treatment'`, `'cate'` returns treatment effect forest predictions and any random treatment effect modeled by a random effects `model_spec` of `'intercept_plus_treatment'`, `'rfx'` returns random effects predictions (if available), and `'variance_forest'` returns variance forest predictions (if available)
9. Whether predictions should be returned as-is from forest / RFX terms (`scale = 'linear'`), transformed to probabilities (`scale = 'probability'`) or class labels (`scale = 'class'`)

## Python

```{python}
#| eval: false
bcf.BCFModel.predict(       # <1>
  X,                        # <2>
  Z,                        # <3>
  propensity=None,          # <4>
  rfx_group_ids=None,       # <5>
  rfx_basis=None,           # <6>
  type='posterior',         # <7>
  terms='all',              # <8>
  scale='linear',           # <9>
)
```

1. `.predict()` called from a reference to the sampled BART model
2. Covariates: all BCF models require covariates so these are not optional in the predict method
3. Treatment: all BCF models require treatment so this is not optional in the predict method
4. Propensity: required if the original model was fit with a user-supplied propensity score
5. Random effects group labels: required if the original model was fit with an additive random effects term
6. Random effects basis: required if the original model was fit with an additive random effects term with custom basis (i.e. `model_spec` was not `"intercept_only"`)
7. Whether predictions should return each posterior draw (`'posterior'`) or be averaged across the posterior (`'mean'`)
8. Which model terms to return predictions for. `'all'` returns every model term available, `'y_hat'` returns only outcome predictions, `'mu'` returns prognostic forest predictions, `'tau'` returns treatment effect forest predictions (and $\tau_0$ intercept, if available), `'prognostic_forest'` returns prognostic forest predictions and any random intercepts modeled by a random effects `model_spec` of `'intercept_only'` or `'intercept_plus_treatment'`, `'cate'` returns treatment effect forest predictions and any random treatment effect modeled by a random effects `model_spec` of `'intercept_plus_treatment'`, `'rfx'` returns random effects predictions (if available), and `'variance_forest'` returns variance forest predictions (if available)
9. Whether predictions should be returned as-is from forest / RFX terms (`scale = 'linear'`), transformed to probabilities (`scale = 'probability'`) or class labels (`scale = 'class'`)

::::

The data inputs to this API are fairly straightforward, but the `type`, `terms` and `scale` arguments merit some detailed (language-agnostic) explanation.

The [BCF overview page](../about/bcf.qmd) has a more in-depth review of the BCF model, but the most complex version a BCF model that `stochtree` supports is, mathematically,
$$
\begin{aligned}
y_{i,j} \mid - &\sim \mathcal{N}\left(\gamma_{0,j} + f_0(x_{i,j}) + \gamma_{1,j} z_{i,j} + \left(b_1 z_{i,j} + b_0 (1-z_{i,j}) \right) \left[\tau_0 + t(x_{i,j})\right], \sigma^2_0 \exp{h(x_{i,j})}\right),\\
f_0 &\sim \text{BART}(\alpha_0, \beta_0, m_0), \\
t &\sim \text{BART}(\alpha_t, \beta_t, m_t), \\
h &\sim \text{logBART}(\alpha_h, \beta_h, m_h), \\
\tau_0 &\sim \mathcal{N}(0, \sigma^2_{\tau}), \\
b_0, b_1 &\sim \mathcal{N}(0, 1/2), \\
\gamma_{0,j}, \gamma_{1,j} &\sim \mathrm{N}(0, \sigma^2_{\gamma}).
\end{aligned}
$$

This model has:

1. Binary treatment with adaptive coding
2. Treatment effect function decomposed into parametric ($\tau_0$) and nonparametric ($t(X)$) terms (i.e. the `sample_intercept` option in the `treatment_effect_forest_params` list / dictionary)
3. Forest-based heteroskedasticity
4. Additive random effects on the `intercept_plus_treatment` basis of $[1,Z]$

#### Prediction Terms

Each of the individual options for the `terms` argument can be mapped to the equation above:

| Term | Mathematical Representation |
| --- | --- |
| `y_hat` | $\gamma_{0,j} + f_0(x_{i,j}) + \gamma_{1,j} z_{i,j} + \left(b_1 z_{i,j} + b_0 (1-z_{i,j}) \right) \left[\tau_0 + t(x_{i,j})\right]$ |
| `mu` | $f_0(x_{i,j}) + b_0 \left[\tau_0 + t(x_{i,j})\right]$ |
| `prognostic_function` | $\gamma_{0,j} + f_0(x_{i,j}) + b_0 \left[\tau_0 + t(x_{i,j})\right]$ |
| `tau` | $\left( b_1 - b_0 \right) \left[\tau_0 + t(x_{i,j})\right]$ |
| `cate` | $\gamma_{1,j} + \left( b_1 - b_0 \right) \left[\tau_0 + t(x_{i,j})\right]$ |
| `rfx` | $\gamma_{0,j} + \gamma_{1,j} z_{i,j}$ |
| `variance_forest` | $\sigma^2_0 \exp{h(x_{i,j})}$ |

Requesting `terms = 'all'` will yield a list / dictionary with each of the terms above stored as separate entries. For models without a variance forest, `terms = 'variance_forest'` will be ignored. For models without random effects, `terms = 'rfx'` will be ignored and `y_hat` collapses to $f_0(x_{i,j}) + \left(b_1 z_{i,j} + b_0 (1-z_{i,j}) \right) \left[\tau_0 + t(x_{i,j})\right]$, and both `'mu'` and `'prognostic_function'` as well as `'tau'` and `'cate'` will be the same.

Without adaptive coding, the terms simplify to

| Term | Mathematical Representation |
| --- | --- |
| `y_hat` | $\gamma_{0,j} + f_0(x_{i,j}) + z_{i,j} \left[\gamma_{1,j} + \tau_0 + t(x_{i,j})\right]$ |
| `mu` | $f_0(x_{i,j})$ |
| `prognostic_function` | $\gamma_{0,j} + f_0(x_{i,j})$ |
| `tau` | $\tau_0 + t(x_{i,j})$ |
| `cate` | $\gamma_{1,j} + \tau_0 + t(x_{i,j})$ |
| `rfx` | $\gamma_{0,j} + \gamma_{1,j} z_{i,j}$ |
| `variance_forest` | $\sigma^2_0 \exp{h(x_{i,j})}$ |

Similarly, without a parametric treatment effect, the terms simplify to

| Term | Mathematical Representation |
| --- | --- |
| `y_hat` | $\gamma_{0,j} + f_0(x_{i,j}) + \gamma_{1,j} z_{i,j} + \left(b_1 z_{i,j} + b_0 (1-z_{i,j}) \right) t(x_{i,j})$ |
| `mu` | $f_0(x_{i,j}) + b_0 t(x_{i,j})$ |
| `prognostic_function` | $\gamma_{0,j} + f_0(x_{i,j}) + b_0 t(x_{i,j})$ |
| `tau` | $\left( b_1 - b_0 \right) t(x_{i,j})$ |
| `cate` | $\gamma_{1,j} + \left( b_1 - b_0 \right) t(x_{i,j})$ |
| `rfx` | $\gamma_{0,j} + \gamma_{1,j} z_{i,j}$ |
| `variance_forest` | $\sigma^2_0 \exp{h(x_{i,j})}$ |

#### Prediction Type

As with BART, `stochtree` will average over posterior samples before returning to users if `type = 'mean'` is specified.

#### Prediction Scale

In addition to the Gaussian model for continuous outcomes in the equation above, `stochtree` supports modeling binary outcomes through a probit link, and the `scale` argument lets users determine which transformation (if any) to apply to model predictions.

The probit model precludes a variance forest, but works with random effects and most other BCF modeling features. Specifying `scale = 'linear'` returns predictions as returned by the mean forest and random effects terms, `scale = 'probability'` converts the predictions into that term's contribution to the conditional probability of `y = 1`, and `scale = 'class'` returns the category index with the highest predicted probability.

## Parameter Extraction

Most BART and BCF models include one or more parametric terms, stored internally as vectors / arrays. Similarly, BART and BCF models often cache predictions for forest terms internally during sampling. `stochtree` provides a standard API for extracting these samples from fitted BART / BCF models.

### BART

The BART parameter extraction API differs stylistically in R and Python but provides the same functionality, which we annotate below

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
extractParameter(
  object, # <1>
  term, # <2>
)
```

1. Reference to the sampled BART model
2. Which internal model samples to extract. Options include `"sigma2"`, `"global_error_scale"`, `"sigma2_global"`, `"sigma2_leaf"`, `"leaf_scale"`, `"y_hat_train"`, `"y_hat_test"`, `"sigma2_x_train"`, `"var_x_train"`, `"sigma2_x_test"`, `"var_x_test"`, `"cloglog_cutpoints"`, `"cutpoints"`

## Python

```{python}
#| eval: false
bart.BARTModel.extract_parameter(     # <1>
  term,                               # <2>
)
```

1. `.extract_parameter()` called from a reference to the sampled BART model
2. Which internal model samples to extract. Options include `"sigma2"`, `"global_error_scale"`, `"sigma2_global"`, `"sigma2_leaf"`, `"leaf_scale"`, `"y_hat_train"`, `"y_hat_test"`, `"sigma2_x_train"`, `"var_x_train"`, `"sigma2_x_test"`, `"var_x_test"`, `"cloglog_cutpoints"`, `"cutpoints"`

::::

This method extracts a vector, matrix, or array of samples from a BART model, queried by name. Random effects are handled by a separate `extract_parameter_samples` method attached to the underlying RandomEffectsContainer object due to the complexity of the random effects parameters. If the requested model term is not found, an error is thrown. The following conventions are used for parameter names:

1. Global error variance: `"sigma2"`, `"global_error_scale"`, `"sigma2_global"`
2. Leaf scale: `"sigma2_leaf"`, `"leaf_scale"`
3. In-sample mean function predictions: `"y_hat_train"`
4. Test set mean function predictions: `"y_hat_test"`
5. In-sample variance forest predictions: `"sigma2_x_train"`, `"var_x_train"`
6. Test set variance forest predictions: `"sigma2_x_test"`, `"var_x_test"`
7. Ordinal model cutpoints (valid only for ordinal cloglog models): `"cloglog_cutpoints"`, `"cutpoints"`

### BCF

The multilingual BCF parameter extraction API is annotated below as with the BART API

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
extractParameter(
  object, # <1>
  term, # <2>
)
```

1. Reference to the sampled BCF model
2. Which internal model samples to extract. Options include `"sigma2"`, `"global_error_scale"`, `"sigma2_global"`, `"sigma2_leaf_mu"`, `"leaf_scale_mu"`, `"mu_leaf_scale"`, `"sigma2_leaf_tau"`, `"leaf_scale_tau"`, `"tau_leaf_scale"`, `"adaptive_coding"`, `"y_hat_train"`, `"y_hat_test"`, `"tau_hat_train"`, `"tau_hat_test"`, `"tau_0"`, `"treatment_intercept"`, `"tau_intercept"`, `"sigma2_x_train"`, `"var_x_train"`, `"sigma2_x_test"`, `"var_x_test"`

## Python

```{python}
#| eval: false
bcf.BCFModel.extract_parameter(     # <1>
  term,                             # <2>
)
```

1. `.extract_parameter()` called from a reference to the sampled BCF model
2. Which internal model samples to extract. Options include `"sigma2"`, `"global_error_scale"`, `"sigma2_global"`, `"sigma2_leaf_mu"`, `"leaf_scale_mu"`, `"mu_leaf_scale"`, `"sigma2_leaf_tau"`, `"leaf_scale_tau"`, `"tau_leaf_scale"`, `"adaptive_coding"`, `"y_hat_train"`, `"y_hat_test"`, `"tau_hat_train"`, `"tau_hat_test"`, `"tau_0"`, `"treatment_intercept"`, `"tau_intercept"`, `"sigma2_x_train"`, `"var_x_train"`, `"sigma2_x_test"`, `"var_x_test"`

::::

This method extracts a vector, matrix, or array of samples from a BART model, queried by name. Random effects are handled by a separate `extract_parameter_samples` method attached to the underlying RandomEffectsContainer object due to the complexity of the random effects parameters. If the requested model term is not found, an error is thrown. The following conventions are used for parameter names:

1. Global error variance: `"sigma2"`, `"global_error_scale"`, `"sigma2_global"`
2. Prognostic forest leaf scale: `"sigma2_leaf_mu"`, `"leaf_scale_mu"`, `"mu_leaf_scale"`
3. Treatment effect forest leaf scale: `"sigma2_leaf_tau"`, `"leaf_scale_tau"`, `"tau_leaf_scale"`
4. Adaptive coding parameters: `"adaptive_coding"` (returns both the control and treated parameters jointly, with control in the first row and treated in the second row)
5. In-sample mean function predictions: `"y_hat_train"`
6. Test set mean function predictions: `"y_hat_test"`
7. In-sample treatment effect forest predictions: `"tau_hat_train"`
8. Test set treatment effect forest predictions: `"tau_hat_test"`
9. Treatment effect intercept: `"tau_0"`, `"treatment_intercept"`, `"tau_intercept"`
10. In-sample variance forest predictions: `"sigma2_x_train"`, `"var_x_train"`
11. Test set variance forest predictions: `"sigma2_x_test"`, `"var_x_test"`

# Demos

We now illustrate these APIs with a series of demos. We start by loading the necessary packages.

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
```

## Python

```{python}
import numpy as np
import matplotlib.pyplot as plt
from sklearn.model_selection import train_test_split
from stochtree import BARTModel, BCFModel
```

::::

Set a seed for reproducibility

::::{.panel-tabset group="language"}

## R

```{r}
random_seed = 1234
set.seed(random_seed)
```

## Python

```{python}
random_seed = 1234
rng = np.random.default_rng(random_seed)
```

::::

## Supervised Learning

We begin with a demo of the supervised learning (BART) use case. We simulate from a regression dataset with leaf-wise basis and group random effects.

::::{.panel-tabset group="language"}

## R

```{r}
n <- 1000
p_x <- 10
p_w <- 1
X <- matrix(runif(n * p_x), ncol = p_x)
W <- matrix(runif(n * p_w), ncol = p_w)
f_XW <- (((0 <= X[, 10]) & (0.25 > X[, 10])) *
  (-7.5 * W[, 1]) +
  ((0.25 <= X[, 10]) & (0.5 > X[, 10])) * (-2.5 * W[, 1]) +
  ((0.5 <= X[, 10]) & (0.75 > X[, 10])) * (2.5 * W[, 1]) +
  ((0.75 <= X[, 10]) & (1 > X[, 10])) * (7.5 * W[, 1]))
rfx_group_ids <- sample(1:3, n, replace = T)
rfx_params <- c(-2, 0, 2)
rfx_basis <- runif(n, -1, 1)
f_rfx <- rfx_params[rfx_group_ids] * rfx_basis
noise_sd <- 1
y <- f_XW + f_rfx + rnorm(n, 0, 1) * noise_sd
```

## Python

```{python}
n = 1000
p_x = 10
p_w = 1
X = rng.uniform(size=(n, p_x))
W = rng.uniform(size=(n, p_w))
f_XW = (
    ((X[:, 9] >= 0)    & (X[:, 9] < 0.25)) * (-7.5 * W[:, 0]) +
    ((X[:, 9] >= 0.25) & (X[:, 9] < 0.5))  * (-2.5 * W[:, 0]) +
    ((X[:, 9] >= 0.5)  & (X[:, 9] < 0.75)) * ( 2.5 * W[:, 0]) +
    ((X[:, 9] >= 0.75) & (X[:, 9] < 1.0))  * ( 7.5 * W[:, 0])
)
rfx_group_ids = rng.choice([0,1,2], size=(n,))
rfx_params = np.array([-2,0,2])
rfx_basis = rng.uniform(low=-1,high=1,size=(n,))
f_rfx = rfx_params[rfx_group_ids] * rfx_basis
y = f_XW + f_rfx + rng.standard_normal(n)
```

::::

And we split the data into train and test sets

::::{.panel-tabset group="language"}

## R

```{r}
n_test <- 200
test_inds <- sample(1:n, n_test, replace = T)
train_inds <- (1:n)[!(1:n %in% test_inds)]
X_train <- X[train_inds, ]
X_test <- X[test_inds, ]
W_train <- W[train_inds, ]
W_test <- W[test_inds, ]
y_train <- y[train_inds]
y_test <- y[test_inds]
rfx_group_ids_train <- rfx_group_ids[train_inds]
rfx_group_ids_test <- rfx_group_ids[test_inds]
rfx_basis_train <- as.matrix(rfx_basis[train_inds])
rfx_basis_test <- as.matrix(rfx_basis[test_inds])
```

## Python

```{python}
n_test = 200
train_inds, test_inds = train_test_split(np.arange(n), train_size=n_test / n)
X_train, X_test = X[train_inds,:], X[test_inds,:]
W_train, W_test = W[train_inds,:], W[test_inds,:]
y_train, y_test = y[train_inds], y[test_inds]
rfx_group_ids_train, rfx_group_ids_test = rfx_group_ids[train_inds], rfx_group_ids[test_inds]
rfx_basis_train, rfx_basis_test = rfx_basis[train_inds], rfx_basis[test_inds]
```

::::

### Example 1(a): BART model without Random Effects or Leaf Regression

First, we fit a simple BART model without additive random effects or leaf regression.

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 5
num_burnin <- 0
num_mcmc <- 10
bart_model <- stochtree::bart(
  X_train = X_train,
  y_train = y_train,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = list(num_threads = 1)
)
```

## Python

```{python}
num_gfr=5
num_burnin=0
num_mcmc=10
bart_model = BARTModel()
bart_model.sample(
    X_train=X_train,
    y_train=y_train,
    num_gfr=num_gfr,
    num_burnin=num_burnin,
    num_mcmc=num_mcmc,
    general_params={"num_threads": 1},
)
```

::::

We extract mean forest predictions and `y_hat` predictions and check that they agree

::::{.panel-tabset group="language"}

## R

```{r}
mean_forest_preds_test <- predict(
  bart_model,
  X = X_test,
  terms = "mean_forest",
  type = "mean"
)
y_hat_test <- predict(bart_model, X = X_test, terms = "y_hat", type = "mean")
plot(y_hat_test, mean_forest_preds_test)
```

## Python

```{python}
mean_forest_preds_test = bart_model.predict(X = X_test, terms = "mean_forest", type = "mean")
y_hat_test = bart_model.predict(X = X_test, terms = "y_hat", type = "mean")
plt.scatter(mean_forest_preds_test, y_hat_test)
```

::::

We now extract and plot global error variance ($\sigma^2$) samples

::::{.panel-tabset group="language"}

## R

```{r}
sigma2_samples <- extractParameter(bart_model, term = "sigma2")
plot(sigma2_samples)
```

## Python

```{python}
sigma2_samples = bart_model.extract_parameter(term = "sigma2")
plt.plot(sigma2_samples)
```

::::

### Example 1(b): BART model with Random Effects and Leaf Regression

First, we fit a more complex BART model with additive random effects and leaf regression.

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 5
num_burnin <- 0
num_mcmc <- 10
bart_model <- stochtree::bart(
  X_train = X_train,
  leaf_basis_train = W_train,
  y_train = y_train,
  rfx_group_ids_train = rfx_group_ids_train,
  rfx_basis_train = rfx_basis_train,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = list(num_threads = 1)
)
```

## Python

```{python}
num_gfr=5
num_burnin=0
num_mcmc=10
bart_model = BARTModel()
bart_model.sample(
    X_train=X_train,
    leaf_basis_train=W_train,
    y_train=y_train,
    rfx_group_ids_train=rfx_group_ids_train,
    rfx_basis_train=rfx_basis_train,
    num_gfr=num_gfr,
    num_burnin=num_burnin,
    num_mcmc=num_mcmc,
    general_params={"num_threads": 1},
)
```

::::

We extract mean forest predictions, RFX predictions and `y_hat` predictions and check that they can be combined to agree

::::{.panel-tabset group="language"}

## R

```{r}
mean_forest_preds_test <- predict(
  bart_model,
  X = X_test,
  leaf_basis = W_test,
  rfx_group_ids = rfx_group_ids_test,
  rfx_basis = rfx_basis_test,
  terms = "mean_forest",
  type = "mean"
)
rfx_preds_test <- predict(
  bart_model,
  X = X_test,
  leaf_basis = W_test,
  rfx_group_ids = rfx_group_ids_test,
  rfx_basis = rfx_basis_test,
  terms = "rfx",
  type = "mean"
)
y_hat_test <- predict(
  bart_model,
  X = X_test,
  leaf_basis = W_test,
  rfx_group_ids = rfx_group_ids_test,
  rfx_basis = rfx_basis_test,
  terms = "y_hat",
  type = "mean"
)
plot(y_hat_test, mean_forest_preds_test + rfx_preds_test)
```

## Python

```{python}
mean_forest_preds_test = bart_model.predict(
    X=X_test,
    leaf_basis=W_test,
    rfx_group_ids=rfx_group_ids_test,
    rfx_basis=rfx_basis_test,
    terms="mean_forest",
    type="mean",
)
rfx_preds_test = bart_model.predict(
    X=X_test,
    leaf_basis=W_test,
    rfx_group_ids=rfx_group_ids_test,
    rfx_basis=rfx_basis_test,
    terms="rfx",
    type="mean",
)
y_hat_test = bart_model.predict(
    X=X_test,
    leaf_basis=W_test,
    rfx_group_ids=rfx_group_ids_test,
    rfx_basis=rfx_basis_test,
    terms="y_hat",
    type="mean",
)
plt.scatter(mean_forest_preds_test + rfx_preds_test, y_hat_test)
```

::::

We now extract and plot global error variance ($\sigma^2$) samples

::::{.panel-tabset group="language"}

## R

```{r}
sigma2_samples <- extractParameter(bart_model, term = "sigma2")
plot(sigma2_samples)
```

## Python

```{python}
sigma2_samples = bart_model.extract_parameter(term = "sigma2")
plt.plot(sigma2_samples)
```

::::
