---
title: "Prior Calibration"
---

Choosing and calibrating the BART tree priors — including the branching process
prior on tree structure and the leaf parameter prior — to match prior beliefs
about the scale of the outcome.

::: {.callout-note}
This vignette is under construction. Content will be ported from:

- **R**: pkgdown article `PriorCalibration`
:::

## Background

BART uses two types of priors:

1. **Tree structure prior**: Controls tree depth via $P(\text{split at depth } d) = \alpha (1 + d)^{-\beta}$.
   Defaults: $\alpha = 0.95$, $\beta = 2$.

2. **Leaf prior**: $\mu_{ij} \sim \mathcal{N}(0, \sigma_\mu^2 / m)$ where $m$ is the number
   of trees. The scale $\sigma_\mu$ is typically calibrated so that $m \cdot \sigma_\mu$
   spans the observed outcome range.

3. **Error variance prior**: $\sigma^2 \sim \text{Inv-}\chi^2(\nu, \lambda)$ with $\nu$ and
   $\lambda$ calibrated from a preliminary linear model fit.

## Setup

```{r}
#| eval: false
library(stochtree)
```

## Default Priors

```{r}
#| eval: false
# Examine default prior settings
```

## Calibrating the Leaf Scale

```{r}
#| eval: false
# Automatic calibration of sigma_mu based on outcome range
```

## Calibrating the Error Variance Prior

```{r}
#| eval: false
# Calibrate nu and lambda from a linear model
```

## Sensitivity Analysis

```{r}
#| eval: false
# Compare posteriors under different prior choices
```
