---
title: "Prior Calibration"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(here::here(), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

This vignette demonstrates prior calibration approaches for the parametric components
of stochastic tree ensembles (@chipman2010bart).

# Background

The "classic" BART model of @chipman2010bart

\begin{equation*}
\begin{aligned}
y &= f(X) + \epsilon\\
f(X) &\sim \text{BART}\left(\alpha, \beta\right)\\
\epsilon &\sim \mathcal{N}\left(0,\sigma^2\right)\\
\sigma^2 &\sim \text{IG}\left(a,b\right)
\end{aligned}
\end{equation*}

is semiparametric, with a nonparametric tree ensemble $f(X)$ and a homoskedastic error
variance parameter $\sigma^2$. Note that in @chipman2010bart, $a$ and $b$ are
parameterized with $a = \frac{\nu}{2}$ and $b = \frac{\nu\lambda}{2}$.

# Setting Priors on Variance Parameters in `stochtree`

By default, `stochtree` employs a Jeffreys' prior for $\sigma^2$
\begin{equation*}
\begin{aligned}
\sigma^2 &\propto \frac{1}{\sigma^2}
\end{aligned}
\end{equation*}
which corresponds to an improper prior with $a = 0$ and $b = 0$.

We provide convenience functions for users wishing to set the $\sigma^2$ prior as in
@chipman2010bart. In this case, $\nu$ is set by default to 3 and $\lambda$ is
calibrated as follows:

1. An "overestimate," $\hat{\sigma}^2$, of $\sigma^2$ is obtained via simple linear
   regression of $y$ on $X$
2. $\lambda$ is chosen to ensure that $p(\sigma^2 < \hat{\sigma}^2) = q$ for some value
   $q$, typically set to a default value of 0.9.

# Setup

Load the necessary packages

:::{.panel-tabset group="language"}

## R

```{r}
#| message: false
library(stochtree)
```

## Python

```{python}
import numpy as np
import matplotlib.pyplot as plt
from stochtree import BARTModel, calibrate_global_error_variance
```

:::

Set a seed for reproducibility

:::{.panel-tabset group="language"}

## R

```{r}
#| message: false
random_seed <- 1234
set.seed(random_seed)
```

## Python

```{python}
random_seed = 1234
rng = np.random.default_rng(random_seed)
```

:::

# Data Generation

Generate data for a straightforward supervised learning problem

::::{.panel-tabset group="language"}

## R

```{r}
n <- 500
p <- 5
X <- matrix(runif(n*p), ncol = p)
f_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (-7.5) +
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (-2.5) +
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2.5) +
    ((0.75 <= X[,1]) & (1 > X[,1])) * (7.5)
)
noise_sd <- 1
y <- f_XW + rnorm(n, 0, noise_sd)
```

## Python

```{python}
n = 500
p = 5
X = rng.uniform(size=(n, p))
f_XW = (
    ((X[:, 0] >= 0)    & (X[:, 0] < 0.25)) * (-7.5) +
    ((X[:, 0] >= 0.25) & (X[:, 0] < 0.5))  * (-2.5) +
    ((X[:, 0] >= 0.5)  & (X[:, 0] < 0.75)) * (2.5)  +
    ((X[:, 0] >= 0.75) & (X[:, 0] < 1.0))  * (7.5)
)
noise_sd = 1.0
y = f_XW + rng.normal(0, noise_sd, n)
```

::::

Split into train and test set

::::{.panel-tabset group="language"}

## R

```{r}
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- X[test_inds,]
X_train <- X[train_inds,]
y_test <- y[test_inds]
y_train <- y[train_inds]
```

## Python

```{python}
test_set_pct = 0.2
n_test = round(test_set_pct * n)
n_train = n - n_test
test_inds = rng.choice(n, n_test, replace=False)
train_inds = np.setdiff1d(np.arange(n), test_inds)
X_test = X[test_inds]
X_train = X[train_inds]
y_test = y[test_inds]
y_train = y[train_inds]
```

::::

# Model Sampling

First, we calibrate the scale parameter for the variance term as in Chipman et al (2010)

::::{.panel-tabset group="language"}

## R

```{r}
nu <- 3
lambda <- calibrateInverseGammaErrorVariance(y_train, X_train, nu = nu)
```

## Python

```{python}
nu = 3
lambda_ = calibrate_global_error_variance(X_train, y_train, nu=nu)
```

::::

Then, we run a BART model with this variance parameterization

::::{.panel-tabset group="language"}

## R

```{r}
general_params <- list(sigma2_global_shape = nu/2, sigma2_global_scale = (nu*lambda)/2)
bart_model <- bart(X_train = X_train, y_train = y_train, X_test = X_test,
                   num_gfr = 0, num_burnin = 1000, num_mcmc = 100,
                   general_params = general_params)
```

## Python

```{python}
bart_model = BARTModel()
bart_model.sample(
    X_train=X_train, y_train=y_train, X_test=X_test,
    num_gfr=0, num_burnin=1000, num_mcmc=100,
    general_params={
        "num_threads": 1,
        "sigma2_global_shape": nu / 2,
        "sigma2_global_scale": (nu * lambda_) / 2,
    },
)
```

::::

Inspect the out-of-sample predictions of the model

::::{.panel-tabset group="language"}

## R

```{r}
plot(rowMeans(bart_model$y_hat_test), y_test, xlab = "predicted", ylab = "actual")
abline(0,1,col="red",lty=3,lwd=3)
```

## Python

```{python}
pred_mean = bart_model.y_hat_test.mean(axis=1)
lo = min(pred_mean.min(), y_test.min())
hi = max(pred_mean.max(), y_test.max())
plt.scatter(pred_mean, y_test, alpha=0.5)
plt.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
plt.xlabel("Predicted")
plt.ylabel("Actual")
plt.show()
```

::::

Inspect the posterior samples of $\sigma^2$

::::{.panel-tabset group="language"}

## R

```{r}
plot(bart_model$sigma2_global_samples, ylab = "sigma^2", xlab = "iteration")
abline(h = noise_sd^2, col = "red", lty = 3, lwd = 3)
```

## Python

```{python}
plt.plot(bart_model.global_var_samples)
plt.xlabel("Iteration")
plt.ylabel(r"$\sigma^2$")
plt.axhline(noise_sd**2, color="red", linestyle="dashed", linewidth=2)
plt.show()
```

::::

# References
