---
title: "Saving and Loading Fitted Models"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(rprojroot::find_root(rprojroot::has_file(".here")), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

This vignette demonstrates how to serialize ensemble models to JSON files and
deserialize back to an R or Python session, where the forests and other parameters
can be used for prediction and further analysis.

# Setup

Load necessary packages

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
```

## Python

```{python}
import json
import numpy as np
import matplotlib.pyplot as plt
import pandas as pd
from scipy.stats import norm
from stochtree import BARTModel, BCFModel
```

::::

Define several simple helper functions used in the data generating processes below

::::{.panel-tabset group="language"}

## R

```{r}
g <- function(x) {ifelse(x[,5]==1,2,ifelse(x[,5]==2,-1,-4))}
mu1 <- function(x) {1+g(x)+x[,1]*x[,3]}
mu2 <- function(x) {1+g(x)+6*abs(x[,3]-1)}
tau1 <- function(x) {rep(3,nrow(x))}
tau2 <- function(x) {1+2*x[,2]*x[,4]}
```

## Python

```{python}
def g(x): return np.where(x[:,4]==1, 2, np.where(x[:,4]==2, -1, -4))
def mu1(x): return 1 + g(x) + x[:,0] * x[:,2]
def mu2(x): return 1 + g(x) + 6 * np.abs(x[:,2] - 1)
def tau1(x): return np.full(x.shape[0], 3.0)
def tau2(x): return 1 + 2 * x[:,1] * x[:,3]
```

::::

Set a seed for reproducibility

::::{.panel-tabset group="language"}

## R

```{r}
random_seed = 1234
set.seed(random_seed)
```

## Python

```{python}
random_seed = 1234
rng = np.random.default_rng(random_seed)
```

::::

# BART Serialization

BART models are initially sampled and constructed using the `bart()` function.
Here we show how to save and reload models from JSON files on disk.

## Model Building

Draw from a relatively straightforward heteroskedastic supervised learning DGP.

::::{.panel-tabset group="language"}

## R

```{r}
# Generate the data
n <- 500
p_x <- 10
X <- matrix(runif(n*p_x), ncol = p_x)
f_XW <- 0
s_XW <- (
    ((0 <= X[,1]) & (0.25 > X[,1])) * (0.5*X[,3]) +
    ((0.25 <= X[,1]) & (0.5 > X[,1])) * (1*X[,3]) +
    ((0.5 <= X[,1]) & (0.75 > X[,1])) * (2*X[,3]) +
    ((0.75 <= X[,1]) & (1 > X[,1])) * (3*X[,3])
)
y <- f_XW + rnorm(n, 0, 1)*s_XW

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
y_test <- y[test_inds]
y_train <- y[train_inds]
s_x_test <- s_XW[test_inds]
s_x_train <- s_XW[train_inds]
```

## Python

```{python}
# Note: new rng here so Python Demo 2 is independent of Demo 1
rng2 = np.random.default_rng(5678)

n = 500
p_x = 10
X2 = rng2.uniform(size=(n, p_x))
s_XW = (
    ((X2[:, 0] >= 0)    & (X2[:, 0] < 0.25)) * (0.5 * X2[:, 2]) +
    ((X2[:, 0] >= 0.25) & (X2[:, 0] < 0.5))  * (1.0 * X2[:, 2]) +
    ((X2[:, 0] >= 0.5)  & (X2[:, 0] < 0.75)) * (2.0 * X2[:, 2]) +
    ((X2[:, 0] >= 0.75) & (X2[:, 0] < 1.0))  * (3.0 * X2[:, 2])
)
y2 = rng2.standard_normal(n) * s_XW

n_test2 = round(0.2 * n)
test_inds2 = rng2.choice(n, n_test2, replace=False)
train_inds2 = np.setdiff1d(np.arange(n), test_inds2)
X_test2 = pd.DataFrame(X2[test_inds2])
X_train2 = pd.DataFrame(X2[train_inds2])
y_test2 = y2[test_inds2]
y_train2 = y2[train_inds2]
```

::::

Sample a BART model.

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
general_params <- list(sample_sigma2_global = F)
mean_forest_params <- list(sample_sigma2_leaf = F, num_trees = 100,
                           alpha = 0.95, beta = 2, min_samples_leaf = 5)
variance_forest_params <- list(num_trees = 50, alpha = 0.95,
                               beta = 1.25, min_samples_leaf = 1)
bart_model <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test,
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    general_params = general_params, mean_forest_params = mean_forest_params,
    variance_forest_params = variance_forest_params
)
```

## Python

```{python}
num_gfr = 10
num_burnin = 0
num_mcmc = 100
bart_model = BARTModel()
bart_model.sample(
    X_train=X_train2, y_train=y_train2, X_test=X_test2,
    num_gfr=num_gfr, num_burnin=num_burnin, num_mcmc=num_mcmc,
    general_params={"num_threads": 1, "sample_sigma2_global": False},
    mean_forest_params={"sample_sigma2_leaf": False, "num_trees": 100,
                        "alpha": 0.95, "beta": 2.0, "min_samples_leaf": 5},
    variance_forest_params={"num_trees": 50, "alpha": 0.95,
                            "beta": 1.25, "min_samples_leaf": 1},
)
```

::::

## Serialization

Save the BART model to disk.

::::{.panel-tabset group="language"}

## R

```{r}
saveBARTModelToJsonFile(bart_model, "bart_r.json")
```

## Python

```{python}
bart_json_string = bart_model.to_json()
with open("bart_py.json", "w") as f:
    json.dump(json.loads(bart_json_string), f)
```

::::

## Deserialization

Reload the BART model from disk.

::::{.panel-tabset group="language"}

## R

```{r}
bart_model_reload <- createBARTModelFromJsonFile("bart_r.json")
```

## Python

```{python}
with open("bart_py.json", "r") as f:
    bart_json_reload = json.dumps(json.load(f))
bart_model_reload = BARTModel()
bart_model_reload.from_json(bart_json_reload)
```

::::

Check that the predictions align with those of the original model.

::::{.panel-tabset group="language"}

## R

```{r}
bart_preds_reload <- predict(bart_model_reload, X_train)
plot(rowMeans(extractParameter(bart_model, "y_hat_train")), rowMeans(bart_preds_reload$y_hat),
     xlab = "Original", ylab = "Deserialized", main = "Conditional Mean Estimates")
abline(0,1,col="red",lwd=3,lty=3)
plot(rowMeans(extractParameter(bart_model, "sigma2_x_train")), rowMeans(bart_preds_reload$variance_forest_predictions),
     xlab = "Original", ylab = "Deserialized", main = "Conditional Variance Estimates")
abline(0,1,col="red",lwd=3,lty=3)
```

## Python

```{python}
bart_preds_orig = bart_model.predict(X=X_train2, terms=["y_hat", "variance_forest"])
bart_preds_reload = bart_model_reload.predict(X=X_train2, terms=["y_hat", "variance_forest"])

fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 5))
yhat_orig = bart_preds_orig["y_hat"].mean(axis=1)
yhat_reload = bart_preds_reload["y_hat"].mean(axis=1)
lo, hi = min(yhat_orig.min(), yhat_reload.min()), max(yhat_orig.max(), yhat_reload.max())
ax1.scatter(yhat_orig, yhat_reload, alpha=0.4, s=10)
ax1.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
ax1.set_xlabel("Original")
ax1.set_ylabel("Deserialized")
ax1.set_title("Conditional Mean Estimates")

# multi-term predict returns variance forest under "variance_forest_predictions"
vhat_orig = bart_preds_orig["variance_forest_predictions"].mean(axis=1)
vhat_reload = bart_preds_reload["variance_forest_predictions"].mean(axis=1)
lo, hi = min(vhat_orig.min(), vhat_reload.min()), max(vhat_orig.max(), vhat_reload.max())
ax2.scatter(vhat_orig, vhat_reload, alpha=0.4, s=10)
ax2.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
ax2.set_xlabel("Original")
ax2.set_ylabel("Deserialized")
ax2.set_title("Conditional Variance Estimates")

plt.tight_layout()
plt.show()
```

::::

# Bayesian Causal Forest (BCF) Serialization

BCF models are initially sampled and constructed using the `bcf()` function.
Here we show how to save and reload models from JSON files on disk.

## Model Building

Draw from a modified version of the data generating process defined in
@hahn2020bayesian.

::::{.panel-tabset group="language"}

## R

```{r}
# Generate synthetic data
n <- 1000
snr <- 2
x1 <- rnorm(n)
x2 <- rnorm(n)
x3 <- rnorm(n)
x4 <- as.numeric(rbinom(n,1,0.5))
x5 <- as.numeric(sample(1:3,n,replace=TRUE))
X <- cbind(x1,x2,x3,x4,x5)
p <- ncol(X)
mu_x <- mu1(X)
tau_x <- tau2(X)
pi_x <- 0.8*pnorm((3*mu_x/sd(mu_x)) - 0.5*X[,1]) + 0.05 + runif(n)/10
Z <- rbinom(n,1,pi_x)
E_XZ <- mu_x + Z*tau_x
rfx_group_ids <- rep(c(1,2), n %/% 2)
rfx_coefs <- matrix(c(-1, -1, 1, 1), nrow=2, byrow=TRUE)
rfx_basis <- cbind(1, runif(n, -1, 1))
rfx_term <- rowSums(rfx_coefs[rfx_group_ids,] * rfx_basis)
y <- E_XZ + rfx_term + rnorm(n, 0, 1)*(sd(E_XZ)/snr)
X <- as.data.frame(X)
X$x4 <- factor(X$x4, ordered = TRUE)
X$x5 <- factor(X$x5, ordered = TRUE)

# Split data into test and train sets
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- X[test_inds,]
X_train <- X[train_inds,]
pi_test <- pi_x[test_inds]
pi_train <- pi_x[train_inds]
Z_test <- Z[test_inds]
Z_train <- Z[train_inds]
y_test <- y[test_inds]
y_train <- y[train_inds]
mu_test <- mu_x[test_inds]
mu_train <- mu_x[train_inds]
tau_test <- tau_x[test_inds]
tau_train <- tau_x[train_inds]
rfx_group_ids_test <- rfx_group_ids[test_inds]
rfx_group_ids_train <- rfx_group_ids[train_inds]
rfx_basis_test <- rfx_basis[test_inds,]
rfx_basis_train <- rfx_basis[train_inds,]
rfx_term_test <- rfx_term[test_inds]
rfx_term_train <- rfx_term[train_inds]
```

## Python

```{python}
random_seed = 1234
rng = np.random.default_rng(random_seed)

n = 1000
snr = 2
x1 = rng.standard_normal(n)
x2 = rng.standard_normal(n)
x3 = rng.standard_normal(n)
x4 = rng.binomial(1, 0.5, n).astype(float)
x5 = rng.choice([1, 2, 3], n).astype(float)
X = np.column_stack([x1, x2, x3, x4, x5])
mu_x = mu1(X)
tau_x = tau2(X)
pi_x = 0.8 * norm.cdf((3 * mu_x / np.std(mu_x)) - 0.5 * X[:, 0]) + 0.05 + rng.uniform(size=n) / 10
Z = rng.binomial(1, pi_x)
E_XZ = mu_x + Z * tau_x
rfx_group_ids = np.tile([1, 2], n // 2)  # 1-indexed group IDs
rfx_coefs = np.array([[-1.0, -1.0], [1.0, 1.0]])
rfx_basis = np.column_stack([np.ones(n), rng.uniform(-1, 1, n)])
rfx_term = np.sum(rfx_coefs[rfx_group_ids - 1] * rfx_basis, axis=1)
y = E_XZ + rfx_term + rng.standard_normal(n) * (np.std(E_XZ) / snr)

# Ordered categoricals
X_df = pd.DataFrame(X, columns=["x1", "x2", "x3", "x4", "x5"])
X_df["x4"] = pd.Categorical(X_df["x4"].astype(int), categories=[0, 1], ordered=True)
X_df["x5"] = pd.Categorical(X_df["x5"].astype(int), categories=[1, 2, 3], ordered=True)

# Train/test split
test_set_pct = 0.2
n_test = round(test_set_pct * n)
test_inds = rng.choice(n, n_test, replace=False)
train_inds = np.setdiff1d(np.arange(n), test_inds)
X_test = X_df.iloc[test_inds]
X_train = X_df.iloc[train_inds]
pi_test = pi_x[test_inds]
pi_train = pi_x[train_inds]
Z_test = Z[test_inds]
Z_train = Z[train_inds]
y_test = y[test_inds]
y_train = y[train_inds]
rfx_group_ids_test = rfx_group_ids[test_inds]
rfx_group_ids_train = rfx_group_ids[train_inds]
rfx_basis_test = rfx_basis[test_inds]
rfx_basis_train = rfx_basis[train_inds]
```

::::

Sample a BCF model.

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
prognostic_forest_params <- list(sample_sigma2_leaf = F)
treatment_effect_forest_params <- list(sample_sigma2_leaf = F)
bcf_model <- bcf(
    X_train = X_train, Z_train = Z_train, y_train = y_train, propensity_train = pi_train,
    rfx_group_ids_train = rfx_group_ids_train, rfx_basis_train = rfx_basis_train,
    X_test = X_test, Z_test = Z_test, propensity_test = pi_test,
    rfx_group_ids_test = rfx_group_ids_test, rfx_basis_test = rfx_basis_test,
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    prognostic_forest_params = prognostic_forest_params,
    treatment_effect_forest_params = treatment_effect_forest_params
)
```

## Python

```{python}
num_gfr = 10
num_burnin = 0
num_mcmc = 100
bcf_model = BCFModel()
bcf_model.sample(
    X_train=X_train, Z_train=Z_train, y_train=y_train, propensity_train=pi_train,
    rfx_group_ids_train=rfx_group_ids_train, rfx_basis_train=rfx_basis_train,
    X_test=X_test, Z_test=Z_test, propensity_test=pi_test,
    rfx_group_ids_test=rfx_group_ids_test, rfx_basis_test=rfx_basis_test,
    num_gfr=num_gfr, num_burnin=num_burnin, num_mcmc=num_mcmc,
    general_params={"num_threads": 1},
    prognostic_forest_params={"sample_sigma2_leaf": False},
    treatment_effect_forest_params={"sample_sigma2_leaf": False},
)
```

::::

## Serialization

Save the BCF model to disk.

::::{.panel-tabset group="language"}

## R

```{r}
saveBCFModelToJsonFile(bcf_model, "bcf_r.json")
```

## Python

```{python}
bcf_json_string = bcf_model.to_json()
with open("bcf_py.json", "w") as f:
    json.dump(json.loads(bcf_json_string), f)
```

::::

## Deserialization

Reload the BCF model from disk.

::::{.panel-tabset group="language"}

## R

```{r}
bcf_model_reload <- createBCFModelFromJsonFile("bcf_r.json")
```

## Python

```{python}
with open("bcf_py.json", "r") as f:
    bcf_json_reload = json.dumps(json.load(f))
bcf_model_reload = BCFModel()
bcf_model_reload.from_json(bcf_json_reload)
```

::::

Check that the predictions align with those of the original model.

::::{.panel-tabset group="language"}

## R

```{r}
bcf_preds_reload <- predict(bcf_model_reload, X_train, Z_train, pi_train, rfx_group_ids_train, rfx_basis_train)
plot(rowMeans(extractParameter(bcf_model, "mu_hat_train")), rowMeans(bcf_preds_reload$mu_hat),
     xlab = "Original", ylab = "Deserialized", main = "Prognostic forest")
abline(0,1,col="red",lwd=3,lty=3)
plot(rowMeans(extractParameter(bcf_model, "tau_hat_train")), rowMeans(bcf_preds_reload$tau_hat),
     xlab = "Original", ylab = "Deserialized", main = "Treatment forest")
abline(0,1,col="red",lwd=3,lty=3)
plot(rowMeans(extractParameter(bcf_model, "y_hat_train")), rowMeans(bcf_preds_reload$y_hat),
     xlab = "Original", ylab = "Deserialized", main = "Overall outcome")
abline(0,1,col="red",lwd=3,lty=3)
```

## Python

```{python}
bcf_preds_orig = bcf_model.predict(
    X=X_train, Z=Z_train, propensity=pi_train,
    rfx_group_ids=rfx_group_ids_train, rfx_basis=rfx_basis_train,
    terms=["mu", "tau", "y_hat"],
)
bcf_preds_reload = bcf_model_reload.predict(
    X=X_train, Z=Z_train, propensity=pi_train,
    rfx_group_ids=rfx_group_ids_train, rfx_basis=rfx_basis_train,
    terms=["mu", "tau", "y_hat"],
)

fig, axes = plt.subplots(1, 3, figsize=(15, 5))
for ax, term, title in zip(
    axes,
    ["mu_hat", "tau_hat", "y_hat"],
    ["Prognostic forest", "Treatment forest", "Overall outcome"],
):
    orig = bcf_preds_orig[term].mean(axis=1)
    reload = bcf_preds_reload[term].mean(axis=1)
    lo, hi = min(orig.min(), reload.min()), max(orig.max(), reload.max())
    ax.scatter(orig, reload, alpha=0.4, s=10)
    ax.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
    ax.set_xlabel("Original")
    ax.set_ylabel("Deserialized")
    ax.set_title(title)
plt.tight_layout()
plt.show()
```

::::

# References
