---
title: "Model Serialization"
---

Saving fitted `stochtree` models to disk and reloading them for prediction
or further analysis — without re-running the sampler.

::: {.callout-note}
This vignette is under construction. Content will be ported from:

- **Python**: `stochtree_repo/demo/notebooks/serialization.ipynb`
- **R**: pkgdown article `ModelSerialization`
:::

## Setup

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
library(stochtree)
```

## Python

```{python}
#| eval: false
import stochtree
import json
```

::::

## Fitting a Model

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Fit a BART model to serialize
```

## Python

```{python}
#| eval: false
# Fit a BART model to serialize
```

::::

## Saving to Disk

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Serialize model to JSON / binary
```

## Python

```{python}
#| eval: false
# Serialize model to JSON / binary
```

::::

## Loading from Disk

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Load serialized model
```

## Python

```{python}
#| eval: false
# Load serialized model
```

::::

## Prediction After Loading

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Predict from reloaded model
```

## Python

```{python}
#| eval: false
# Predict from reloaded model
```

::::
