---
title: "Posterior Summary and Visualization Utilities"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(rprojroot::find_root(rprojroot::has_file(".here")), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

This vignette demonstrates the summary and plotting utilities available for
`stochtree` models.

# Setup

Load necessary packages

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
```

## Python

```{python}
import numpy as np
import matplotlib.pyplot as plt
from stochtree import BARTModel, BCFModel, plot_parameter_trace
```

::::

Set a seed for reproducibility

::::{.panel-tabset group="language"}

## R

```{r}
random_seed = 1234
set.seed(random_seed)
```

## Python

```{python}
random_seed = 1234
rng = np.random.default_rng(random_seed)
```

::::

# Supervised Learning

We begin with the supervised learning use case served by the `bart()` function.

Below we simulate a simple regression dataset.

::::{.panel-tabset group="language"}

## R

```{r}
n <- 1000
p_x <- 10
p_w <- 1
X <- matrix(runif(n * p_x), ncol = p_x)
W <- matrix(runif(n * p_w), ncol = p_w)
f_XW <- (((0 <= X[, 10]) & (0.25 > X[, 10])) *
  (-7.5 * W[, 1]) +
  ((0.25 <= X[, 10]) & (0.5 > X[, 10])) * (-2.5 * W[, 1]) +
  ((0.5 <= X[, 10]) & (0.75 > X[, 10])) * (2.5 * W[, 1]) +
  ((0.75 <= X[, 10]) & (1 > X[, 10])) * (7.5 * W[, 1]))
noise_sd <- 1
y <- f_XW + rnorm(n, 0, 1) * noise_sd
```

## Python

```{python}
n = 1000
p_x = 10
p_w = 1
X = rng.uniform(size=(n, p_x))
W = rng.uniform(size=(n, p_w))
# R uses X[,10] (1-indexed) = Python X[:,9]
f_XW = (
    ((X[:, 9] >= 0)    & (X[:, 9] < 0.25)) * (-7.5 * W[:, 0]) +
    ((X[:, 9] >= 0.25) & (X[:, 9] < 0.5))  * (-2.5 * W[:, 0]) +
    ((X[:, 9] >= 0.5)  & (X[:, 9] < 0.75)) * ( 2.5 * W[:, 0]) +
    ((X[:, 9] >= 0.75) & (X[:, 9] < 1.0))  * ( 7.5 * W[:, 0])
)
noise_sd = 1.0
y = f_XW + rng.standard_normal(n) * noise_sd
```

::::

Now we fit a simple BART model to the data.

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 1000
general_params <- list(
  num_threads = 1, 
  num_chains = 3
)
bart_model <- stochtree::bart(
  X_train = X,
  y_train = y,
  leaf_basis_train = W,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = general_params
)
```

## Python

```{python}
bart_model = BARTModel()
bart_model.sample(
    X_train=X,
    y_train=y,
    leaf_basis_train=W,
    num_gfr=10,
    num_burnin=0,
    num_mcmc=1000,
    general_params={
      "num_threads": 1, 
      "num_chains": 3
    },
)
```

::::

We obtain a high level summary of the BART model by running `print()`.

::::{.panel-tabset group="language"}

## R

```{r}
print(bart_model)
```

## Python

```{python}
print(bart_model)
```

::::

For a more detailed summary (including the information above), we use the `summary()`
function.

::::{.panel-tabset group="language"}

## R

```{r}
summary(bart_model)
```

## Python

```{python}
print(bart_model.summary())
```

::::

We can use the `plot()` function to produce a traceplot of model terms like the global
error scale $\sigma^2$ or (if $\sigma^2$ is not sampled) the first observation of
cached train set predictions.

::::{.panel-tabset group="language"}

## R

```{r}
plot(bart_model)
```

## Python

```{python}
ax = plot_parameter_trace(bart_model, term="global_error_scale")
plt.show()
```

::::

For finer-grained control over which parameters to plot, we can also use the
`extractParameter()` function to pull the posterior distribution of any valid model
term (e.g., global error scale $\sigma^2$, leaf scale $\sigma^2_{\ell}$, in-sample
mean function predictions `y_hat_train`) and then plot any subset or transformation
of these values.

::::{.panel-tabset group="language"}

## R

```{r}
y_hat_train_samples <- extractParameter(bart_model, "y_hat_train")
obs_index <- 1
plot(
  y_hat_train_samples[obs_index, ],
  type = "l",
  main = paste0("In-Sample Predictions Traceplot, Observation ", obs_index),
  xlab = "Index",
  ylab = "Parameter Values"
)
```

## Python

```{python}
y_hat_train_samples = bart_model.extract_parameter("y_hat_train")
obs_index = 0
fig, ax = plt.subplots()
ax.plot(y_hat_train_samples[obs_index, :])
ax.set_title(f"In-Sample Predictions Traceplot, Observation {obs_index}")
ax.set_xlabel("Index")
ax.set_ylabel("Parameter Values")
plt.show()
```

::::

# Causal Inference

We now run the same demo for the causal inference use case served by the `bcf()` function in R and the `BCFModel` Python class.

Below we simulate a simple dataset for a causal inference problem with binary treatment and continuous outcome.

::::{.panel-tabset group="language"}

## R

```{r}
# Generate covariates and treatment
n <- 1000
p_X = 5
X = matrix(runif(n * p_X), ncol = p_X)
pi_X = 0.25 + 0.5 * X[, 1]
Z = rbinom(n, 1, pi_X)

# Define the outcome mean functions (prognostic and treatment effects)
mu_X = pi_X * 5 + 2 * X[, 3]
tau_X = X[, 2] * 2 - 1

# Generate outcome
epsilon = rnorm(n, 0, 1)
y = mu_X + tau_X * Z + epsilon
```

## Python

```{python}
# Generate covariates and treatment
n = 1000
p_X = 5
X = rng.uniform(size=(n, p_X))
pi_X = 0.25 + 0.5 * X[:, 0]
Z = rng.binomial(1, pi_X, n).astype(float)

# Define the outcome mean functions (prognostic and treatment effects)
mu_X = pi_X * 5 + 2 * X[:, 2]
tau_X = X[:, 1] * 2 - 1

# Generate outcome
epsilon = rng.standard_normal(n)
y = mu_X + tau_X * Z + epsilon
```

::::

Now we fit a simple BCF model to the data

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 1000
general_params <- list(
  num_threads = 1, 
  num_chains = 3, 
  adaptive_coding = TRUE
)
bcf_model <- stochtree::bcf(
  X_train = X,
  y_train = y,
  Z_train = Z,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = general_params
)
```

## Python

```{python}
bcf_model = BCFModel()
bcf_model.sample(
    X_train=X,
    Z_train=Z,
    y_train=y,
    propensity_train=pi_X,
    num_gfr=10,
    num_burnin=0,
    num_mcmc=1000,
    general_params={
      "num_threads": 1, 
      "num_chains": 3, 
      "adaptive_coding": True
    },
)
```

::::

We obtain a high level summary of the BCF model by running `print()`.

::::{.panel-tabset group="language"}

## R

```{r}
print(bcf_model)
```

## Python

```{python}
print(bcf_model)
```

::::

For a more detailed summary (including the information above), we use the `summary()` function / method.

::::{.panel-tabset group="language"}

## R

```{r}
summary(bcf_model)
```

## Python

```{python}
print(bcf_model.summary())
```

::::

In R, we have a `plot()` that produces a traceplot of model terms like the global error scale $\sigma^2$ or (if $\sigma^2$ is not sampled) the first observation of cached train set predictions.

In Python, we provide a `plot_parameter_trace()` function for requesting a traceplot of a specific model parameter.

::::{.panel-tabset group="language"}

## R

```{r}
plot(bcf_model)
```

## Python

```{python}
ax = plot_parameter_trace(bcf_model, term="global_error_scale")
plt.show()
```

::::

For finer-grained control over which parameters to plot, we can also use the `extractParameter()` function in R or the `extract_parameter()` method in Python to query the posterior distribution of any valid model term (e.g., global error scale $\sigma^2$, prognostic forest leaf scale $\sigma^2_{\mu}$, CATE forest leaf scale $\sigma^2_{\tau}$, adaptive coding parameters $b_0$ and $b_1$ for binary treatment, in-sample mean function predictions `y_hat_train`, in-sample CATE function predictions `tau_hat_train`) and then plot any subset or transformation of these values.

::::{.panel-tabset group="language"}

## R

```{r}
adaptive_coding_samples <- extractParameter(bcf_model, "adaptive_coding")
plot(
  adaptive_coding_samples[1, ],
  type = "l",
  main = "Adaptive Coding Parameter Traceplot",
  xlab = "Index",
  ylab = "Parameter Values",
  ylim = range(adaptive_coding_samples),
  col = "blue"
)
lines(adaptive_coding_samples[2, ], col = "orange")
legend(
  "topright",
  legend = c("Control", "Treated"),
  lty = 1,
  col = c("blue", "orange")
)
```

## Python

```{python}
adaptive_coding_samples = bcf_model.extract_parameter("adaptive_coding")
fig, ax = plt.subplots()
ax.plot(adaptive_coding_samples[0, :], color="blue", label="Control")
ax.plot(adaptive_coding_samples[1, :], color="orange", label="Treated")
ax.set_title("Adaptive Coding Parameter Traceplot")
ax.set_xlabel("Index")
ax.set_ylabel("Parameter Values")
ax.legend(loc="upper right")
plt.show()
```

::::
