---
title: "Summary and Plotting"
bibliography: vignettes.bib
---

This vignette demonstrates the summary and plotting utilities available for
`stochtree` models.

# Setup

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
random_seed = 1234
set.seed(random_seed)
```

## Python

```{python}
# Python implementation coming soon
```

::::

# Supervised Learning

We begin with the supervised learning use case served by the `bart()` function.

Below we simulate a simple regression dataset.

::::{.panel-tabset group="language"}

## R

```{r}
n <- 1000
p_x <- 10
p_w <- 1
X <- matrix(runif(n * p_x), ncol = p_x)
W <- matrix(runif(n * p_w), ncol = p_w)
f_XW <- (((0 <= X[, 10]) & (0.25 > X[, 10])) *
  (-7.5 * W[, 1]) +
  ((0.25 <= X[, 10]) & (0.5 > X[, 10])) * (-2.5 * W[, 1]) +
  ((0.5 <= X[, 10]) & (0.75 > X[, 10])) * (2.5 * W[, 1]) +
  ((0.75 <= X[, 10]) & (1 > X[, 10])) * (7.5 * W[, 1]))
noise_sd <- 1
y <- f_XW + rnorm(n, 0, 1) * noise_sd
```

## Python

```{python}
# Python implementation coming soon
```

::::

Now we fit a simple BART model to the data.

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 1000
general_params <- list(num_chains = 3)
bart_model <- stochtree::bart(
  X_train = X,
  y_train = y,
  leaf_basis_train = W,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = general_params
)
```

## Python

```{python}
# Python implementation coming soon
```

::::

We obtain a high level summary of the BART model by running `print()`.

::::{.panel-tabset group="language"}

## R

```{r}
print(bart_model)
```

## Python

```{python}
# Python implementation coming soon
```

::::

For a more detailed summary (including the information above), we use the `summary()`
function.

::::{.panel-tabset group="language"}

## R

```{r}
summary(bart_model)
```

## Python

```{python}
# Python implementation coming soon
```

::::

We can use the `plot()` function to produce a traceplot of model terms like the global
error scale $\sigma^2$ or (if $\sigma^2$ is not sampled) the first observation of
cached train set predictions.

::::{.panel-tabset group="language"}

## R

```{r}
plot(bart_model)
```

## Python

```{python}
# Python implementation coming soon
```

::::

For finer-grained control over which parameters to plot, we can also use the
`extractParameter()` function to pull the posterior distribution of any valid model
term (e.g., global error scale $\sigma^2$, leaf scale $\sigma^2_{\ell}$, in-sample
mean function predictions `y_hat_train`) and then plot any subset or transformation
of these values.

::::{.panel-tabset group="language"}

## R

```{r}
y_hat_train_samples <- extractParameter(bart_model, "y_hat_train")
obs_index <- 1
plot(
  y_hat_train_samples[obs_index, ],
  type = "l",
  main = paste0("In-Sample Predictions Traceplot, Observation ", obs_index),
  xlab = "Index",
  ylab = "Parameter Values"
)
```

## Python

```{python}
# Python implementation coming soon
```

::::

# Causal Inference

We now run the same demo for the causal inference use case served by the `bcf()`
function.

Below we simulate a simple dataset for a causal inference problem with binary
treatment and continuous outcome.

::::{.panel-tabset group="language"}

## R

```{r}
# Generate covariates and treatment
n <- 1000
p_X = 5
X = matrix(runif(n * p_X), ncol = p_X)
pi_X = 0.25 + 0.5 * X[, 1]
Z = rbinom(n, 1, pi_X)

# Define the outcome mean functions (prognostic and treatment effects)
mu_X = pi_X * 5 + 2 * X[, 3]
tau_X = X[, 2] * 2 - 1

# Generate outcome
epsilon = rnorm(n, 0, 1)
y = mu_X + tau_X * Z + epsilon
```

## Python

```{python}
# Python implementation coming soon
```

::::

Now we fit a simple BCF model to the data.

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 1000
general_params <- list(num_chains = 3)
bcf_model <- stochtree::bcf(
  X_train = X,
  y_train = y,
  Z_train = Z,
  num_gfr = num_gfr,
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = general_params
)
```

## Python

```{python}
# Python implementation coming soon
```

::::

We obtain a high level summary of the BCF model by running `print()`.

::::{.panel-tabset group="language"}

## R

```{r}
print(bcf_model)
```

## Python

```{python}
# Python implementation coming soon
```

::::

For a more detailed summary (including the information above), we use the `summary()`
function.

::::{.panel-tabset group="language"}

## R

```{r}
summary(bcf_model)
```

## Python

```{python}
# Python implementation coming soon
```

::::

We can use the `plot()` function to produce a traceplot of model terms like the global
error scale $\sigma^2$ or (if $\sigma^2$ is not sampled) the first observation of
cached train set predictions.

::::{.panel-tabset group="language"}

## R

```{r}
plot(bcf_model)
```

## Python

```{python}
# Python implementation coming soon
```

::::

For finer-grained control over which parameters to plot, we can also use the
`extractParameter()` function to pull the posterior distribution of any valid model
term (e.g., global error scale $\sigma^2$, prognostic forest leaf scale
$\sigma^2_{\mu}$, CATE forest leaf scale $\sigma^2_{\tau}$, adaptive coding
parameters $b_0$ and $b_1$ for binary treatment, in-sample mean function predictions
`y_hat_train`, in-sample CATE function predictions `tau_hat_train`) and then plot any
subset or transformation of these values.

::::{.panel-tabset group="language"}

## R

```{r}
adaptive_coding_samples <- extractParameter(bcf_model, "adaptive_coding")
plot(
  adaptive_coding_samples[1, ],
  type = "l",
  main = "Adaptive Coding Parameter Traceplot",
  xlab = "Index",
  ylab = "Parameter Values",
  ylim = range(adaptive_coding_samples),
  col = "blue"
)
lines(adaptive_coding_samples[2, ], col = "orange")
legend(
  "topright",
  legend = c("Control", "Treated"),
  lty = 1,
  col = c("blue", "orange")
)
```

## Python

```{python}
# Python implementation coming soon
```

::::
