---
title: "Targeted Smooth BART Approximation via Leaf Regression"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(rprojroot::find_root(rprojroot::has_file(".here")), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

This vignette demonstrates how to approximate **BART with Targeted Smoothing (tsBART)** (@starling2020bart) using `stochtree`'s leaf regression interface. tsBART extends BART by placing a Gaussian Process (GP) in the leaves of each tree, allowing smooth, nonlinear functions of a designated "target" variable $u$ within each leaf node defined by covariates $X$.

The key idea is that a GP on a compact interval can be well approximated by a linear regression on a finite-dimensional basis expansion of $u$.

# Hilbert Space GP Approximation

We omit much of the formalism and derivation of the Hilbert space GP approximation in this vignette. A blog post by @orduz2024hsgp provides an excellent overview, citing both @solin2020hilbert and @riutort2023practical for further academic references.

Let $u \in [-L, L]$ be the target variable, symmetric around zero. A GP with a squared exponential kernel $K$ on $u$ is defined as

$$
\begin{aligned}
g(u) &\sim \mathcal{GP}(0, K)\\
K(u, u') &= \sigma^2 \exp\left(-\frac{(u - u')^2}{2\ell^2}\right)
\end{aligned}
$$

The Hilbert space approximation represents a GP $g(u)$ as

$$
g(u) \approx \sum_{j=1}^{J} \sqrt{S(\lambda_j)} \phi_j(u) \beta_j, \quad \beta_j \sim \mathcal{N}(0, I)
$$

where $J$ is the total number of basis functions in the finite-dimensional expansion, and $\phi_j$, $S(\cdot)$, and $\lambda_j$ are defined as follows:

$$
\begin{aligned}
\phi_j(u) &= \frac{1}{\sqrt{L}} \sin\!\left(\frac{\pi j (u + L)}{2L}\right), \\
S(\sigma, \ell, \omega) &= \sigma^2 \sqrt{2\pi}\, \ell \exp\!\left(-\frac{\ell^2 \omega^2}{2}\right)\\
\lambda_j &= \left(\frac{\pi j}{2L}\right)^2
\end{aligned}
$$

Evaluating $\phi_j$ at the data points and scaling by $\sqrt{S(\lambda_j)}$ gives a basis matrix $\Omega$ such that $\Omega \beta \sim \mathcal{N}(0, K)$ where $K$ is the kernel matrix — so a standard $\mathcal{N}(0, I)$ prior on $\beta$ induces the target GP prior on $g(u)$.

In the tsBART approximation, $X$ partitions the data within each tree (as in standard BART) and $g_{\mathcal{L}}(u) = \Omega_{\mathcal{L}} \beta_{\mathcal{L}}$ is a GP on each leaf $\mathcal{L}$. The combined prediction at $(X_i, u_i)$ is

$$
\hat{y}_i = \sum_{t=1}^{T} g_{\mathcal{L}_t(X_i)}(u_i) = \sum_{t=1}^{T} \Omega(u_i)^\top \beta_{\mathcal{L}_t(X_i)}
$$

where $T$ is the total number of trees in the ensemble.

This can be sampled straightforwardly as a `stochtree` BART model with `leaf_basis_train = Omega`.

# Setup

Load necessary packages

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
library(ggplot2)
```

## Python

```{python}
import matplotlib.pyplot as plt
import numpy as np
from scipy.stats import norm
from stochtree import BARTModel
```

::::

Set a seed for reproducibility

::::{.panel-tabset group="language"}

## R

```{r}
random_seed <- 1234
set.seed(random_seed)
```

## Python

```{python}
random_seed = 1234
rng = np.random.default_rng(random_seed)
```

::::


# Simulation

We simulate a dataset in which the response depends on $u$ and $X$ through an interaction:

$$
f(u, X) = 3 X_1 u + \frac{(X_2 - 1)(1 - u)^2}{3} + \Phi(2 X_3 u)
$$

where $u \sim \text{Uniform}(-0.5, 0.5)$ is the GP target variable, $X \in \mathbb{R}^3$ are additional covariates, and $\Phi$ is the standard normal CDF.

::::{.panel-tabset group="language"}

## R

```{r}
n <- 1000
p <- 5

u <- runif(n, -0.5, 0.5)
X <- matrix(rnorm(n * p), nrow = n)

f_true <- function(row) {
  # row = c(u, x1, x2, x3)
  3 * row[2] * row[1] + (row[3] - 1) * (1 - row[1])^2 / 3 + pnorm(2 * row[4] * row[1])
}

f0 <- apply(cbind(u, X), 1, f_true)
y  <- f0 + rnorm(n, 0, 0.25 * sd(f0))

cat("Signal SD:", round(sd(f0), 3), "\n")
cat("Noise SD: ", round(0.25 * sd(f0), 3), "\n")
```

## Python

```{python}
n = 1000
p = 5

u = rng.uniform(-0.5, 0.5, n)
X = rng.standard_normal((n, p))

def f_true(u, x):
    return 3 * x[0] * u + (x[1] - 1) * (1 - u)**2 / 3 + norm.cdf(2 * x[2] * u)

f0 = np.array([f_true(u[i], X[i]) for i in range(n)])
y  = f0 + rng.normal(0, 0.25 * f0.std(), n)

print(f"Signal SD: {f0.std():.3f}")
print(f"Noise SD:  {0.25 * f0.std():.3f}")
```

::::

# Constructing the Basis Matrix

Two helper functions below construct the basis:

- `Omega(u, J, L)` — evaluates the $J$ eigenfunctions at the points in `u`
- `SDiag(sigma, l, J, L)` — diagonal matrix of spectral density values $S(\lambda_j)$

::::{.panel-tabset group="language"}

## R

```{r}
Omega <- function(u, J, L) {
  phi <- function(u, j, L) sqrt(1 / L) * sin(pi * j * (u + L) / (2 * L))
  omega <- matrix(0, nrow = length(u), ncol = J)
  for (j in 1:J) {
    omega[, j] <- phi(u, j, L)
  }
  return(omega)
}

SDiag <- function(sigma, l, J, L) {
  eigenval <- function(j, L) (pi * j / (2 * L))^2
  s <- function(sigma, l, w) sigma^2 * sqrt(2 * pi) * l * exp(-0.5 * l^2 * w^2)
  return(diag(s(sigma, l, sqrt(eigenval(1:J, L)))))
}
```

## Python

```{python}
def Omega(u, J, L):
    j = np.arange(1, J + 1)
    return np.sqrt(1 / L) * np.sin(np.pi * j[None, :] * (u[:, None] + L) / (2 * L))


def SDiag(sigma, l, J, L):
    j = np.arange(1, J + 1)
    eigenvals = (np.pi * j / (2 * L)) ** 2
    s = sigma**2 * np.sqrt(2 * np.pi) * l * np.exp(-0.5 * l**2 * eigenvals)
    return np.diag(s)
```

::::

We set $L = 5/8$ (the half-width of the $u$ domain with a small buffer), $J = 12$ basis functions, and the GP lengthscale to $\ell = \text{range}(u) / (2\pi)$ following @solin2020hilbert.

::::{.panel-tabset group="language"}

## R

```{r}
L <- 5 / 4 * max(abs(u)) # domain half-width with buffer
J <- 12 # number of basis functions
l_gp <- diff(range(u)) / (2 * pi) # GP lengthscale
sigma_gp <- 1 # marginal SD; handled by the leaf scale in the model

# Basis at training points
scales <- sqrt(SDiag(sigma_gp, l_gp, J, L))
omega <- Omega(u, J, L)
omega_scaled <- omega %*% scales # scaled: N(0,I) prior on coefs induces GP prior

# Basis on a fine grid for posterior curve plots
u_grid <- seq(-0.5, 0.5, length.out = 500)
omega_grid <- Omega(u_grid, J, L)
omega_grid_scaled <- omega_grid %*% scales
```

## Python

```{python}
L = 5 / 4 * np.abs(u).max()  # domain half-width with buffer
J = 12  # number of basis functions
l_gp = (u.max() - u.min()) / (2 * np.pi)  # GP lengthscale
sigma_gp = 1.0  # marginal SD; handled by the leaf scale in the model

# Basis at training points
scales = np.sqrt(SDiag(sigma_gp, l_gp, J, L))
omega = Omega(u, J, L)
omega_scaled = omega @ scales  # scaled: N(0,I) prior on coefs induces GP prior

# Basis on a fine grid for posterior curve plots
u_grid = np.linspace(-0.5, 0.5, 500)
omega_grid = Omega(u_grid, J, L)
omega_grid_scaled = omega_grid @ scales
```

::::

We can visualise the scaled basis functions to get a sense of the smoothness they encode:

::::{.panel-tabset group="language"}

## R

```{r}
#| fig-width: 7
#| fig-height: 3
#| fig-cap: "First six scaled Hilbert-space basis functions."
matplot(
  u_grid,
  omega_grid_scaled[, 1:6],
  type = "l",
  lty = 1,
  xlab = "u",
  ylab = expression(phi[j](u)),
  main = "Scaled basis functions (j = 1–6)"
)
```

## Python

```{python}
#| fig-cap: "First six scaled Hilbert-space basis functions."
fig, ax = plt.subplots(figsize=(7, 3))
for j in range(6):
    ax.plot(u_grid, omega_grid_scaled[:, j], label=f"j={j + 1}")
ax.set_xlabel("u")
ax.set_ylabel(r"$\phi_j(u)$")
ax.set_title("Scaled basis functions (j = 1–6)")
ax.legend(ncol=3, fontsize=8)
plt.tight_layout()
plt.show()
```

::::

# Sampling and Analysis

We pass `omega_scaled` as the leaf basis.

::::{.panel-tabset group="language"}

## R

```{r}
num_trees <- 100
num_gfr <- 10
num_burnin <- 500
num_mcmc <- 500

bart_model <- bart(
  X_train = X,
  y_train = y,
  leaf_basis_train = omega_scaled,
  num_gfr = num_gfr, 
  num_burnin = num_burnin,
  num_mcmc = num_mcmc,
  general_params = list(
    num_threads = 1,
    sample_sigma2_global = TRUE,
    random_seed = random_seed
  ),
  mean_forest_params = list(num_trees = num_trees, sample_sigma2_leaf = FALSE)
)
```

## Python

```{python}
num_trees = 100
num_gfr = 10
num_burnin = 500
num_mcmc = 500

bart_model = BARTModel()
bart_model.sample(
    X_train=X,
    y_train=y,
    leaf_basis_train=omega_scaled,
    num_gfr=num_gfr,
    num_burnin=num_burnin,
    num_mcmc=num_mcmc,
    general_params={
        "num_threads": 1,
        "sample_sigma2_global": True,
        "random_seed": random_seed,
    },
    mean_forest_params={"num_trees": num_trees, "sample_sigma2_leaf": False},
)
```

::::

## In-Sample Fit

We compare the true outcomes to this model's estimated conditional posterior mean

::::{.panel-tabset group="language"}

## R

```{r}
#| fig-width: 5
#| fig-height: 5
#| fig-cap: "Posterior mean in-sample predictions versus the true function."
y_hat_train <- extractParameter(bart_model, "y_hat_train")
y_hat_mean <- rowMeans(y_hat_train)

plot(
  f0,
  y_hat_mean,
  xlab = "True f(u, X)",
  ylab = "Posterior mean prediction",
  main = "In-sample fit",
  pch = 20,
  col = "#1a7a9c80"
)
abline(0, 1, col = "firebrick", lwd = 1.5)
```

```{r}
cat("RMSE:", round(sqrt(mean((y_hat_mean - f0)^2)), 4), "\n")
```

## Python

```{python}
#| fig-cap: "Posterior mean in-sample predictions versus the true function."
y_hat_train = bart_model.extract_parameter("y_hat_train")
y_hat_mean = y_hat_train.mean(axis=1)

fig, ax = plt.subplots(figsize=(5, 5))
ax.scatter(f0, y_hat_mean, alpha=0.5, color="#1a7a9c", s=20)
lims = [min(f0.min(), y_hat_mean.min()), max(f0.max(), y_hat_mean.max())]
ax.plot(lims, lims, color="firebrick", linewidth=1.5)
ax.set_xlabel("True f(u, X)")
ax.set_ylabel("Posterior mean prediction")
ax.set_title("In-sample fit")
plt.tight_layout()
plt.show()
```

```{python}
rmse = np.sqrt(np.mean((y_hat_mean - f0)**2))
print(f"RMSE: {rmse:.4f}")
```

::::

## Posterior GP Curves

A distinctive feature of tsBART is that we can draw posterior GP curves: for a fixed covariate vector $x^*$, we evaluate the model over a fine grid of $u$ values to visualise the smooth function $g(u \mid x^*)$.

This works by constructing a test set in which $X_{\text{test}}$ has $x^*$ repeated for every grid point and `leaf_basis` sweeps over the grid basis `omega_grid_scaled`. Because every grid point shares the same $X$, they all fall into the same leaf in each tree — so the prediction traces the GP assigned to that partition cell.

::::{.panel-tabset group="language"}

## R

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Posterior GP curves at three randomly selected observations. The dotted line is the true f(u, x*), the dark line is the posterior mean, and the shaded band is the 90% credible interval."
plot_ids <- sample(n, 3)

par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
for (ix in plot_ids) {
  xx <- X[ix, , drop = FALSE]

  # Repeat x* for every grid point; vary only the leaf basis over the u grid
  X_grid <- matrix(
    rep(xx, nrow(omega_grid_scaled)),
    nrow = nrow(omega_grid_scaled),
    byrow = TRUE
  )
  pred_grid <- predict(
    bart_model,
    X = X_grid,
    leaf_basis = omega_grid_scaled,
    terms = "y_hat"
  )
  # pred_grid is (n_grid x num_mcmc)

  pred_mean <- rowMeans(pred_grid)
  pred_lo <- apply(pred_grid, 1, quantile, 0.05)
  pred_hi <- apply(pred_grid, 1, quantile, 0.95)

  f_true_curve <- apply(
    cbind(u_grid, matrix(rep(xx, length(u_grid)), ncol = p, byrow = TRUE)),
    1,
    f_true
  )

  plot(
    u_grid,
    pred_mean,
    type = "l",
    lwd = 2,
    ylim = range(c(pred_lo, pred_hi, f_true_curve)),
    xlab = "u",
    ylab = "f(u, x*)",
    main = paste0("Obs. ", ix)
  )
  polygon(
    c(u_grid, rev(u_grid)),
    c(pred_hi, rev(pred_lo)),
    col = "#1a7a9c20",
    border = NA
  )
  lines(u_grid, pred_mean, lwd = 2, col = "#1a7a9c")
  lines(u_grid, f_true_curve, col = "firebrick", lwd = 1.5, lty = 2)
  points(u[ix], y[ix], pch = 19, col = "black")
}
par(mfrow = c(1, 1))
```

## Python

```{python}
#| fig-cap: "Posterior GP curves at three randomly selected observations. The dotted line is the true f(u, x*), the dark line is the posterior mean, and the shaded band is the 90% credible interval."
rng2 = np.random.default_rng(42)
plot_ids = rng2.choice(n, 3, replace=False)

fig, axes = plt.subplots(1, 3, figsize=(10, 4))
for ax, ix in zip(axes, plot_ids):
    xx = X[[ix], :]  # shape (1, p)

    # Repeat x* for every grid point; vary only the leaf basis over the u grid
    X_grid = np.repeat(xx, len(u_grid), axis=0)  # (n_grid, p)
    pred_grid = bart_model.predict(
        X=X_grid, leaf_basis=omega_grid_scaled, terms="y_hat"
    )  # (n_grid, num_mcmc)

    pred_mean = pred_grid.mean(axis=1)
    pred_lo = np.quantile(pred_grid, 0.05, axis=1)
    pred_hi = np.quantile(pred_grid, 0.95, axis=1)

    f_true_curve = np.array([f_true(ug, xx[0]) for ug in u_grid])

    ax.fill_between(u_grid, pred_lo, pred_hi, alpha=0.15, color="#1a7a9c")
    ax.plot(u_grid, pred_mean, color="#1a7a9c", linewidth=2, label="Posterior mean")
    ax.plot(
        u_grid,
        f_true_curve,
        color="firebrick",
        linewidth=1.5,
        linestyle="--",
        label="True f",
    )
    ax.scatter([u[ix]], [y[ix]], color="black", zorder=5, s=30)
    ax.set_xlabel("u")
    ax.set_ylabel("f(u, x*)")
    ax.set_title(f"Obs. {ix}")

axes[0].legend(fontsize=8)
plt.tight_layout()
plt.show()
```

::::

The model recovers smooth curves in $u$ that adapt to the local covariate context defined by $x^*$.

# References
