---
title: "Examining Individual Trees in a Fitted Ensemble"
bibliography: vignettes.bib
execute:
  freeze: auto  # re-render only when source changes
---

```{r}
#| include: false
reticulate::use_python(
  Sys.getenv(
    "RETICULATE_PYTHON",
    unset = file.path(here::here(), ".venv", "bin", "python")
  ),
  required = TRUE
)
```

While out of sample evaluation and MCMC diagnostics on parametric BART components
(i.e. $\sigma^2$, the global error variance) are helpful, it's important to be able
to inspect the trees in a BART / BCF model. This vignette walks through some of the
features `stochtree` provides to query and understand the forests and trees in a model.

# Setup

Load necessary packages

::::{.panel-tabset group="language"}

## R

```{r}
library(stochtree)
```

## Python

```{python}
import numpy as np
import matplotlib.pyplot as plt
from stochtree import BARTModel
```

::::

Set a seed for reproducibility

::::{.panel-tabset group="language"}

## R

```{r}
random_seed = 1234
set.seed(random_seed)
```

## Python

```{python}
random_seed = 1234
rng = np.random.default_rng(random_seed)
```

::::

# Data Generation

Generate sample data where feature 10 is the only "important" feature

::::{.panel-tabset group="language"}

## R

```{r}
n <- 500
p_x <- 10
X <- matrix(runif(n*p_x), ncol = p_x)
f_XW <- (
    ((0 <= X[,10]) & (0.25 > X[,10])) * (-7.5) +
    ((0.25 <= X[,10]) & (0.5 > X[,10])) * (-2.5) +
    ((0.5 <= X[,10]) & (0.75 > X[,10])) * (2.5) +
    ((0.75 <= X[,10]) & (1 > X[,10])) * (7.5)
)
noise_sd <- 1
y <- f_XW + rnorm(n, 0, 1)*noise_sd
```

## Python

```{python}
n = 500
p_x = 10
X = rng.uniform(size=(n, p_x))
# Feature 10 (R) = feature index 9 (Python, 0-indexed)
f_XW = (
    ((X[:, 9] >= 0)    & (X[:, 9] < 0.25)) * (-7.5) +
    ((X[:, 9] >= 0.25) & (X[:, 9] < 0.5))  * (-2.5) +
    ((X[:, 9] >= 0.5)  & (X[:, 9] < 0.75)) * (2.5)  +
    ((X[:, 9] >= 0.75) & (X[:, 9] < 1.0))  * (7.5)
)
noise_sd = 1.0
y = f_XW + rng.standard_normal(n) * noise_sd
```

::::

Split into train and test sets

::::{.panel-tabset group="language"}

## R

```{r}
test_set_pct <- 0.2
n_test <- round(test_set_pct*n)
n_train <- n - n_test
test_inds <- sort(sample(1:n, n_test, replace = FALSE))
train_inds <- (1:n)[!((1:n) %in% test_inds)]
X_test <- as.data.frame(X[test_inds,])
X_train <- as.data.frame(X[train_inds,])
y_test <- y[test_inds]
y_train <- y[train_inds]
```

## Python

```{python}
n_test = round(0.2 * n)
test_inds = rng.choice(n, n_test, replace=False)
train_inds = np.setdiff1d(np.arange(n), test_inds)
X_test = X[test_inds]
X_train = X[train_inds]
y_test = y[test_inds]
y_train = y[train_inds]
```

::::

# Model Sampling

Sample a BART model with 10 GFR and 100 MCMC iterations

::::{.panel-tabset group="language"}

## R

```{r}
num_gfr <- 10
num_burnin <- 0
num_mcmc <- 100
general_params <- list(keep_gfr = T)
bart_model <- stochtree::bart(
    X_train = X_train, y_train = y_train, X_test = X_test,
    num_gfr = num_gfr, num_burnin = num_burnin, num_mcmc = num_mcmc,
    general_params = general_params
)
```

## Python

```{python}
num_gfr = 10
num_burnin = 0
num_mcmc = 100
bart_model = BARTModel()
bart_model.sample(
    X_train=X_train, y_train=y_train, X_test=X_test,
    num_gfr=num_gfr, num_burnin=num_burnin, num_mcmc=num_mcmc,
    general_params={"num_threads": 1, "keep_gfr": True},
)
```

::::

# Model Inspection

Assess the global error variance traceplot and test set prediction quality

::::{.panel-tabset group="language"}

## R

```{r}
sigma2_samples <- extractParameter(bart_model, "sigma2_global")
plot(sigma2_samples, ylab="sigma^2")
abline(h=noise_sd^2,col="red",lty=2,lwd=2.5)
y_hat_test <- predict(bart_model, X=X_test, type="mean", terms="y_hat")
plot(y_hat_test, y_test, pch=16, cex=0.75, xlab = "pred", ylab = "actual")
abline(0,1,col="red",lty=2,lwd=2.5)
```

## Python

```{python}
fig, (ax1, ax2) = plt.subplots(1, 2, figsize=(12, 4))

sigma2_samples = bart_model.extract_parameter("sigma2_global")
ax1.plot(sigma2_samples)
ax1.axhline(noise_sd**2, color="red", linestyle="dashed", linewidth=2)
ax1.set_ylabel(r"$\sigma^2$")

y_hat_test = bart_model.predict(X=X_test, terms="y_hat", type="mean")
ax2.scatter(y_hat_test, y_test, s=15, alpha=0.6)
lo = min(y_hat_test.min(), y_test.min())
hi = max(y_hat_test.max(), y_test.max())
ax2.plot([lo, hi], [lo, hi], color="red", linestyle="dashed", linewidth=2)
ax2.set_xlabel("pred")
ax2.set_ylabel("actual")

plt.tight_layout()
plt.show()
```

::::

## Variable Split Counts

The `get_forest_split_counts` method of a BART model's internal forest objects allows us to compute the number of times each variable was used in a split rule across all trees in a given forest. 

Below we query this vector for the final GFR sample (1-indexed as 10 in R, 0-indexed as 9 in Python), where the second argument is the dimensionality of the covariates.

::::{.panel-tabset group="language"}

## R

```{r}
bart_model$mean_forests$get_forest_split_counts(10, p_x)
```

## Python

```{python}
bart_model.forest_container_mean.get_forest_split_counts(9, p_x)
```

::::

We can also compute split counts for each feature aggregated over all forests

::::{.panel-tabset group="language"}

## R

```{r}
bart_model$mean_forests$get_aggregate_split_counts(p_x)
```

## Python

```{python}
bart_model.forest_container_mean.get_overall_split_counts(p_x)
```

::::

The split counts appear relatively uniform across features, so let's dig deeper and
look at individual trees.

The `get_granular_split_counts` method returns a 3-dimensional array of shape `(num_forests, num_trees, num_features)`, where each entry represents the number of times a feature was used in a split for a specific tree in a specific forest. 

That is, we can count the number of times feature $k$ was split on in tree $j$ of forest $i$ by looking at the `(i,j,k)` entry of this array.

Below we compute the split count for all features in the first tree of the last GFR sample in our model (noting again the use of 1-indexing in R and 0-indexing in Python).

::::{.panel-tabset group="language"}

## R

```{r}
splits = bart_model$mean_forests$get_granular_split_counts(p_x)
splits[10,1,]
```

## Python

```{python}
splits = bart_model.forest_container_mean.get_granular_split_counts(p_x)
splits[9, 0, :]
```

::::

This tree has a single split on the only "important" feature (10). Now, let's look at
the second tree.

::::{.panel-tabset group="language"}

## R

```{r}
splits[10,2,]
```

## Python

```{python}
splits[9, 1, :]
```

::::

And the 20th and 30th trees

::::{.panel-tabset group="language"}

## R

```{r}
splits[10,20,]
```

## Python

```{python}
splits[9, 19, :]
```

::::

::::{.panel-tabset group="language"}

## R

```{r}
splits[10,30,]
```

## Python

```{python}
splits[9, 29, :]
```

::::

We see that "later" trees are splitting on other features, but we also note that these
trees are fitting an outcome that is already residualized by many "relevant splits"
made by trees 1 and 2.

## Tree Structure

Now, let's inspect the first tree for the last GFR sample in more depth, following
[this scikit-learn vignette](https://scikit-learn.org/stable/auto_examples/tree/plot_unveil_tree_structure.html).

::::{.panel-tabset group="language"}

## R

```{r}
forest_num <- 9
tree_num <- 0
nodes <- sort(bart_model$mean_forests$nodes(forest_num, tree_num))
for (nid in nodes) {
    if (bart_model$mean_forests$is_leaf_node(forest_num, tree_num, nid)) {
        node_depth <- bart_model$mean_forests$node_depth(forest_num, tree_num, nid)
        space_text <- rep("\t", node_depth)
        leaf_values <- bart_model$mean_forests$node_leaf_values(forest_num, tree_num, nid)
        cat(space_text, "node=", nid, " is a leaf node with value=",
            format(leaf_values, digits = 3), "\n", sep = "")
    } else {
        node_depth <- bart_model$mean_forests$node_depth(forest_num, tree_num, nid)
        space_text <- rep("\t", node_depth)
        left <- bart_model$mean_forests$left_child_node(forest_num, tree_num, nid)
        feature <- bart_model$mean_forests$node_split_index(forest_num, tree_num, nid)
        threshold <- bart_model$mean_forests$node_split_threshold(forest_num, tree_num, nid)
        right <- bart_model$mean_forests$right_child_node(forest_num, tree_num, nid)
        cat(space_text, "node=", nid, " is a split node, which tells us to go to node ",
            left, " if X[:, ", feature, "] <= ", format(threshold, digits = 3),
            " else to node ", right, "\n", sep = "")
    }
}
```

## Python

```{python}
forest_num = 9
tree_num = 0
fc = bart_model.forest_container_mean
nodes = np.sort(fc.nodes(forest_num, tree_num))
for nid in nodes:
    depth = fc.node_depth(forest_num, tree_num, nid)
    indent = "\t" * depth
    if fc.is_leaf_node(forest_num, tree_num, nid):
        value = np.round(fc.node_leaf_values(forest_num, tree_num, nid), 3)
        print(f"{indent}node={nid} is a leaf node with value={value}")
    else:

        left = fc.left_child_node(forest_num, tree_num, nid)
        feature = fc.node_split_index(forest_num, tree_num, nid)
        threshold = round(fc.node_split_threshold(forest_num, tree_num, nid), 3)
        right = fc.right_child_node(forest_num, tree_num, nid)
        print(f"{indent}node={nid} is a split node, which tells us to go to node "
              f"{left} if X[:, {feature}] <= {threshold} else to node {right}")
```

::::
