---
title: "Tree Inspection"
---

Examining the structure and parameters of individual trees within a fitted
`stochtree` ensemble — useful for interpretability and debugging.

::: {.callout-note}
This vignette is under construction. Content will be ported from:

- **Python**: `stochtree_repo/demo/notebooks/tree_inspection.ipynb`
- **R**: pkgdown article `TreeInspection`
:::

## Setup

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
library(stochtree)
```

## Python

```{python}
#| eval: false
import stochtree
import numpy as np
```

::::

## Fitting a Model

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Fit a model to inspect
```

## Python

```{python}
#| eval: false
# Fit a model to inspect
```

::::

## Accessing the Forest Container

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Extract forest container from fitted model
```

## Python

```{python}
#| eval: false
# Extract forest container from fitted model
```

::::

## Inspecting Tree Structure

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Navigate individual trees: split variables, thresholds, leaf values
```

## Python

```{python}
#| eval: false
# Navigate individual trees: split variables, thresholds, leaf values
```

::::

## Summarizing Split Frequencies

::::{.panel-tabset group="language"}

## R

```{r}
#| eval: false
# Variable importance via split counts
```

## Python

```{python}
#| eval: false
# Variable importance via split counts
```

::::
